---
title: "Analysis-Resynthesis Workflow: FormantPath + KlattGrid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis-Resynthesis Workflow: FormantPath + KlattGrid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates the complete **analysis-resynthesis workflow** in pladdrr, combining:

1. **FormantPath** - Robust formant extraction with multiple ceiling candidates
2. **KlattGrid** - Parametric speech synthesis from formant specifications

This workflow enables powerful speech research applications:

- **Voice morphing** - Transform speaker characteristics
- **Perceptual experiments** - Manipulate acoustic parameters systematically
- **Quality assessment** - Verify formant tracking accuracy via resynthesis
- **Speech restoration** - Clean and regenerate degraded audio

```{r setup, message=FALSE}
library(pladdrr)
library(ggplot2)
library(dplyr)
```

## Why Analysis-Resynthesis?

### The Perceptual Loop

The analysis-resynthesis cycle closes the loop between acoustic analysis and perceptual reality:

```
Real Speech → FormantPath Analysis → Formant Parameters
      ↑                                        ↓
Perceptual Validation ← KlattGrid Synthesis ← Modified Parameters
```

**Benefits**:

1. **Validation** - If resynthesis sounds similar, analysis is accurate
2. **Manipulation** - Systematically modify acoustic parameters
3. **Understanding** - Bridge between measurements and perception
4. **Control** - Create impossible stimuli (e.g., male formants + female pitch)

### Applications

- **Phonetics research**: Vowel space mapping, coarticulation studies
- **Speech technology**: Voice conversion, synthesis evaluation
- **Clinical**: Voice disorder assessment and therapy
- **Education**: Teaching formant-perception relationships

## Workflow 1: Basic Analysis → Synthesis

### Step 1: Load and Analyze Audio

```{r load-analyze}
# Load audio file
sound_orig <- Sound(system.file("extdata", "test.wav", package = "pladdrr"))

cat("Original audio:\n")
cat("  Duration:", sound_orig$get_duration(), "s\n")
cat("  Samples:", sound_orig$get_number_of_samples(), "\n")

# Extract formants with FormantPath (robust multi-ceiling)
fp <- sound_orig$to_formant_path(
  time_step = 0.005,
  max_num_formants = 5,
  formant_ceiling = 5500,
  num_steps_up_down = 2L  # Test 5 different ceilings
)

cat("\nFormantPath analysis:\n")
cat("  Candidates tested:", fp$get_number_of_candidates(), "\n")
cat("  Ceiling range:", 
    round(min(fp$get_all_ceiling_frequencies())), "-",
    round(max(fp$get_all_ceiling_frequencies())), "Hz\n")

# Extract optimal formant track
formant <- fp$extract_formant()
cat("  Formant frames:", formant$get_number_of_frames(), "\n")
```

### Step 2: Extract Mean Formant Values

```{r extract-formants}
# Convert to data frame for easy manipulation
df <- as.data.frame(formant)

# Calculate mean F1, F2, F3 across all frames
f1_mean <- mean(df[df$formant == 1, "frequency"], na.rm = TRUE)
f2_mean <- mean(df[df$formant == 2, "frequency"], na.rm = TRUE)
f3_mean <- mean(df[df$formant == 3, "frequency"], na.rm = TRUE)

cat("Mean formant values:\n")
cat(sprintf("  F1: %.0f Hz\n", f1_mean))
cat(sprintf("  F2: %.0f Hz\n", f2_mean))
cat(sprintf("  F3: %.0f Hz\n", f3_mean))

# Also extract pitch for realistic resynthesis
pitch <- sound_orig$to_pitch()
f0_mean <- mean(as.data.frame(pitch)$frequency, na.rm = TRUE)

cat(sprintf("  F0 (pitch): %.0f Hz\n", f0_mean))
```

### Step 3: Resynthesize with KlattGrid

```{r resynthesize}
# Create KlattGrid with extracted parameters
kg <- KlattGrid_createFromVowel(
  duration = min(sound_orig$get_duration(), 1.0),  # Limit for demo
  f0start = f0_mean,
  f1 = f1_mean, b1 = 80,    # Use typical bandwidths
  f2 = f2_mean, b2 = 120,
  f3 = f3_mean, b3 = 150
)

# Synthesize audio
sound_resynth <- kg$to_sound()

cat("Resynthesized audio:\n")
cat("  Duration:", sound_resynth$get_duration(), "s\n")
cat("  Samples:", sound_resynth$get_number_of_samples(), "\n")
```

### Step 4: Compare Original vs Resynthesized

```{r compare-spectrograms, fig.width=8, fig.height=6}
# Create spectrograms for comparison
spec_orig <- sound_orig$to_spectrogram(
  window_length = 0.005,
  maximum_frequency = 5000
)

spec_resynth <- sound_resynth$to_spectrogram(
  window_length = 0.005,
  maximum_frequency = 5000
)

# Extract formants for overlay
formant_orig <- sound_orig$to_formant_burg(formant_ceiling = 5500)
formant_resynth <- sound_resynth$to_formant_burg(formant_ceiling = 5500)

df_orig <- as.data.frame(formant_orig) %>% 
  filter(formant %in% 1:3) %>%
  mutate(source = "Original")

df_resynth <- as.data.frame(formant_resynth) %>% 
  filter(formant %in% 1:3) %>%
  mutate(source = "Resynthesized")

# Combine for plotting
df_combined <- bind_rows(df_orig, df_resynth)

# Plot formant tracks comparison
ggplot(df_combined, aes(time, frequency, color = factor(formant))) +
  geom_line(alpha = 0.7, linewidth = 0.8) +
  facet_wrap(~ source, ncol = 1) +
  scale_color_manual(
    values = c("1" = "#E41A1C", "2" = "#377EB8", "3" = "#4DAF4A"),
    labels = c("F1", "F2", "F3")
  ) +
  labs(
    title = "Original vs Resynthesized Formant Tracks",
    subtitle = "FormantPath analysis → KlattGrid synthesis",
    x = "Time (s)",
    y = "Frequency (Hz)",
    color = "Formant"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

## Workflow 2: Vowel Round-Trip Validation

### Create Synthetic Vowel with Known Parameters

```{r synthetic-vowel}
# Define target /a/ vowel parameters
target_f1 <- 730
target_f2 <- 1090
target_f3 <- 2440

cat("Target vowel: /a/\n")
cat(sprintf("  F1: %d Hz\n", target_f1))
cat(sprintf("  F2: %d Hz\n", target_f2))
cat(sprintf("  F3: %d Hz\n", target_f3))

# Synthesize
kg_vowel <- KlattGrid_createFromVowel(
  duration = 0.5,
  f0start = 120,
  f1 = target_f1, b1 = 80,
  f2 = target_f2, b2 = 120,
  f3 = target_f3, b3 = 150
)

sound_vowel <- kg_vowel$to_sound()
```

### Analyze Synthetic Vowel

```{r analyze-synthetic}
# Analyze with FormantPath
fp_vowel <- sound_vowel$to_formant_path(
  time_step = 0.005,
  formant_ceiling = 5500,
  num_steps_up_down = 2L
)

formant_vowel <- fp_vowel$extract_formant()

# Extract values at midpoint
mid_time <- formant_vowel$get_duration() / 2

f1_extracted <- formant_vowel$get_value_at_time(1, mid_time, "HERTZ")
f2_extracted <- formant_vowel$get_value_at_time(2, mid_time, "HERTZ")
f3_extracted <- formant_vowel$get_value_at_time(3, mid_time, "HERTZ")

cat("\nExtracted formants (at midpoint):\n")
cat(sprintf("  F1: %.0f Hz\n", f1_extracted))
cat(sprintf("  F2: %.0f Hz\n", f2_extracted))
cat(sprintf("  F3: %.0f Hz\n", f3_extracted))

# Calculate errors
f1_error <- abs(f1_extracted - target_f1) / target_f1 * 100
f2_error <- abs(f2_extracted - target_f2) / target_f2 * 100
f3_error <- abs(f3_extracted - target_f3) / target_f3 * 100

cat("\nExtraction accuracy:\n")
cat(sprintf("  F1 error: %.1f%%\n", f1_error))
cat(sprintf("  F2 error: %.1f%%\n", f2_error))
cat(sprintf("  F3 error: %.1f%%\n", f3_error))
```

### Visualize Round-Trip

```{r round-trip-viz, fig.width=8, fig.height=4}
# Create comparison data frame
df_comparison <- data.frame(
  Formant = rep(c("F1", "F2", "F3"), 2),
  Frequency = c(target_f1, target_f2, target_f3,
                f1_extracted, f2_extracted, f3_extracted),
  Type = rep(c("Target", "Extracted"), each = 3)
)

ggplot(df_comparison, aes(Formant, Frequency, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.0f Hz", Frequency)), 
            position = position_dodge(width = 0.8),
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("Target" = "#377EB8", "Extracted" = "#E41A1C")) +
  labs(
    title = "Round-Trip Validation: Target vs Extracted Formants",
    subtitle = "Synthesis → Analysis cycle for /a/ vowel",
    y = "Frequency (Hz)"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

## Workflow 3: Vowel Space Mapping

### Synthesize Vowel Triangle

```{r vowel-triangle-synth}
# Define cardinal vowels
vowels <- list(
  i = list(label = "/i/ (beet)", f1 = 280, f2 = 2250, f3 = 2890),
  a = list(label = "/a/ (father)", f1 = 730, f2 = 1090, f3 = 2440),
  u = list(label = "/u/ (boot)", f1 = 310, f2 = 870, f3 = 2250)
)

# Synthesize each vowel
vowel_sounds <- list()
vowel_results <- list()

for (v_name in names(vowels)) {
  v <- vowels[[v_name]]
  
  # Synthesize
  kg <- KlattGrid_createFromVowel(
    duration = 0.3,
    f0start = 150,
    f1 = v$f1, b1 = 60,
    f2 = v$f2, b2 = 100,
    f3 = v$f3, b3 = 140
  )
  
  vowel_sounds[[v_name]] <- kg$to_sound()
  
  cat(sprintf("Synthesized %s: F1=%d, F2=%d, F3=%d Hz\n",
              v$label, v$f1, v$f2, v$f3))
}
```

### Analyze Vowel Triangle

```{r vowel-triangle-analyze}
# Analyze each synthetic vowel
for (v_name in names(vowels)) {
  sound <- vowel_sounds[[v_name]]
  
  # Robust formant tracking
  fp <- sound$to_formant_path(num_steps_up_down = 1L)
  formant <- fp$extract_formant()
  
  # Get mean formants
  df <- as.data.frame(formant)
  f1_mean <- mean(df[df$formant == 1, "frequency"], na.rm = TRUE)
  f2_mean <- mean(df[df$formant == 2, "frequency"], na.rm = TRUE)
  
  vowel_results[[v_name]] <- data.frame(
    vowel = v_name,
    label = vowels[[v_name]]$label,
    target_f1 = vowels[[v_name]]$f1,
    target_f2 = vowels[[v_name]]$f2,
    extracted_f1 = f1_mean,
    extracted_f2 = f2_mean
  )
  
  cat(sprintf("Analyzed %s: F1=%.0f, F2=%.0f Hz\n",
              vowels[[v_name]]$label, f1_mean, f2_mean))
}

# Combine results
vowel_space <- do.call(rbind, vowel_results)
```

### Plot Vowel Space

```{r vowel-space-plot, fig.width=8, fig.height=6}
# Create vowel space plot (F1 vs F2)
vowel_space_long <- rbind(
  data.frame(
    vowel = vowel_space$vowel,
    label = vowel_space$label,
    F1 = vowel_space$target_f1,
    F2 = vowel_space$target_f2,
    type = "Target"
  ),
  data.frame(
    vowel = vowel_space$vowel,
    label = vowel_space$label,
    F1 = vowel_space$extracted_f1,
    F2 = vowel_space$extracted_f2,
    type = "Extracted"
  )
)

ggplot(vowel_space_long, aes(F2, F1, color = type, shape = vowel)) +
  geom_point(size = 5, alpha = 0.8) +
  geom_line(aes(group = vowel), color = "gray50", linetype = "dashed") +
  geom_text(aes(label = vowel), vjust = -1.5, size = 5, fontface = "bold") +
  scale_x_reverse() +  # Traditional vowel space orientation
  scale_y_reverse() +
  scale_color_manual(values = c("Target" = "#377EB8", "Extracted" = "#E41A1C")) +
  scale_shape_manual(values = c("i" = 16, "a" = 17, "u" = 15)) +
  labs(
    title = "Vowel Space: Synthesis-Analysis Round Trip",
    subtitle = "Target (blue) vs Extracted (red) formant values",
    x = "F2 (Hz)",
    y = "F1 (Hz)",
    color = "Type",
    shape = "Vowel"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major = element_line(color = "gray90")
  )
```

### Verify Vowel Space Relationships

```{r vowel-relationships}
# Check phonetic relationships are preserved
cat("Vowel space relationships:\n\n")

# F1: /a/ should be highest (lowest vowel)
cat("F1 ordering (high to low vowel height):\n")
cat(sprintf("  /a/: %.0f Hz (low vowel, high F1)\n", 
            vowel_space[vowel_space$vowel == "a", "extracted_f1"]))
cat(sprintf("  /i/: %.0f Hz (high vowel, low F1)\n", 
            vowel_space[vowel_space$vowel == "i", "extracted_f1"]))
cat(sprintf("  /u/: %.0f Hz (high vowel, low F1)\n", 
            vowel_space[vowel_space$vowel == "u", "extracted_f1"]))

f1_a <- vowel_space[vowel_space$vowel == "a", "extracted_f1"]
f1_i <- vowel_space[vowel_space$vowel == "i", "extracted_f1"]
f1_u <- vowel_space[vowel_space$vowel == "u", "extracted_f1"]

cat(sprintf("  ✓ /a/ F1 > /i/ F1: %s\n", f1_a > f1_i))
cat(sprintf("  ✓ /a/ F1 > /u/ F1: %s\n", f1_a > f1_u))

# F2: /i/ (front) > /a/ (central) > /u/ (back)
cat("\nF2 ordering (front to back):\n")
cat(sprintf("  /i/: %.0f Hz (front, high F2)\n", 
            vowel_space[vowel_space$vowel == "i", "extracted_f2"]))
cat(sprintf("  /a/: %.0f Hz (central, mid F2)\n", 
            vowel_space[vowel_space$vowel == "a", "extracted_f2"]))
cat(sprintf("  /u/: %.0f Hz (back, low F2)\n", 
            vowel_space[vowel_space$vowel == "u", "extracted_f2"]))

f2_i <- vowel_space[vowel_space$vowel == "i", "extracted_f2"]
f2_a <- vowel_space[vowel_space$vowel == "a", "extracted_f2"]
f2_u <- vowel_space[vowel_space$vowel == "u", "extracted_f2"]

cat(sprintf("  ✓ /i/ F2 > /a/ F2: %s\n", f2_i > f2_a))
cat(sprintf("  ✓ /a/ F2 > /u/ F2: %s\n", f2_a > f2_u))
```

## Advanced: Voice Morphing

### Morph Between Two Speakers

```{r voice-morph}
# Define two speaker profiles
speaker_male <- list(
  f1 = 730, f2 = 1090, f3 = 2440,
  f0 = 100, label = "Male"
)

speaker_female <- list(
  f1 = 850, f2 = 1300, f3 = 2800,
  f0 = 220, label = "Female"
)

# Create morph continuum (5 steps)
morph_alphas <- seq(0, 1, length.out = 5)

morph_results <- data.frame()

for (i in seq_along(morph_alphas)) {
  alpha <- morph_alphas[i]
  
  # Linear interpolation between speakers
  f1 <- speaker_male$f1 + alpha * (speaker_female$f1 - speaker_male$f1)
  f2 <- speaker_male$f2 + alpha * (speaker_female$f2 - speaker_male$f2)
  f3 <- speaker_male$f3 + alpha * (speaker_female$f3 - speaker_male$f3)
  f0 <- speaker_male$f0 + alpha * (speaker_female$f0 - speaker_male$f0)
  
  # Synthesize
  kg <- KlattGrid_createFromVowel(
    duration = 0.4,
    f0start = f0,
    f1 = f1, b1 = 80,
    f2 = f2, b2 = 120,
    f3 = f3, b3 = 150
  )
  
  sound <- kg$to_sound()
  
  morph_results <- rbind(morph_results, data.frame(
    step = i,
    alpha = alpha,
    f0 = f0,
    f1 = f1,
    f2 = f2,
    f3 = f3,
    label = sprintf("%.0f%% Female", alpha * 100)
  ))
  
  cat(sprintf("Step %d (%.0f%% Female): F0=%.0f, F1=%.0f, F2=%.0f Hz\n",
              i, alpha * 100, f0, f1, f2))
}
```

### Visualize Morph Continuum

```{r morph-viz, fig.width=8, fig.height=5}
# Plot formant trajectories across morph
morph_long <- tidyr::pivot_longer(
  morph_results,
  cols = c(f0, f1, f2, f3),
  names_to = "parameter",
  values_to = "frequency"
)

ggplot(morph_long, aes(alpha, frequency, color = parameter)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("f0" = "#984EA3", "f1" = "#E41A1C", 
               "f2" = "#377EB8", "f3" = "#4DAF4A"),
    labels = c("F0 (pitch)", "F1", "F2", "F3")
  ) +
  labs(
    title = "Voice Morphing: Male → Female Continuum",
    subtitle = "Linear interpolation of formant and pitch parameters",
    x = "Morph Parameter (0 = Male, 1 = Female)",
    y = "Frequency (Hz)",
    color = "Parameter"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

## Use Cases

### 1. Formant Manipulation Experiments

Systematically vary one formant while holding others constant:

```{r formant-manipulation, eval=FALSE}
# Create F1 continuum (F2/F3 constant)
f1_values <- seq(400, 800, by = 50)  # Low to high

stimuli <- lapply(f1_values, function(f1) {
  kg <- KlattGrid_createFromVowel(
    duration = 0.5, f0start = 120,
    f1 = f1, b1 = 80,        # VARIED
    f2 = 1500, b2 = 120,      # CONSTANT
    f3 = 2500, b3 = 150       # CONSTANT
  )
  
  sound <- kg$to_sound()
  sound$save(sprintf("f1_%04d.wav", f1), "WAV")
  
  sound
})
```

### 2. Pitch-Formant Decoupling

Test perception with impossible voice combinations:

```{r pitch-formant-decouple, eval=FALSE}
# Male formants + Female pitch (impossible in nature)
kg_impossible <- KlattGrid_createFromVowel(
  duration = 0.5,
  f0start = 220,              # Female pitch
  f1 = 730, b1 = 80,          # Male formants
  f2 = 1090, b2 = 120,
  f3 = 2440, b3 = 150
)

sound_impossible <- kg_impossible$to_sound()
```

### 3. Quality Assessment

Verify formant tracking accuracy on known stimuli:

```{r quality-assessment, eval=FALSE}
# Generate test vowels with known formants
test_vowels <- expand.grid(
  f1 = seq(300, 800, by = 100),
  f2 = seq(900, 2400, by = 300)
)

accuracy <- apply(test_vowels, 1, function(params) {
  # Synthesize
  kg <- KlattGrid_createFromVowel(
    duration = 0.5, f0start = 120,
    f1 = params[1], b1 = 80,
    f2 = params[2], b2 = 120,
    f3 = 2500, b3 = 150
  )
  sound <- kg$to_sound()
  
  # Analyze
  fp <- sound$to_formant_path(num_steps_up_down = 2L)
  formant <- fp$extract_formant()
  
  # Compare
  df <- as.data.frame(formant)
  f1_extracted <- mean(df[df$formant == 1, "frequency"], na.rm = TRUE)
  f2_extracted <- mean(df[df$formant == 2, "frequency"], na.rm = TRUE)
  
  data.frame(
    target_f1 = params[1],
    target_f2 = params[2],
    extracted_f1 = f1_extracted,
    extracted_f2 = f2_extracted,
    error_f1 = abs(f1_extracted - params[1]),
    error_f2 = abs(f2_extracted - params[2])
  )
})

# Analyze accuracy across vowel space
accuracy_df <- do.call(rbind, accuracy)
mean(accuracy_df$error_f1, na.rm = TRUE)  # Mean F1 error in Hz
```

## Best Practices

### ✓ DO

1. **Use FormantPath for analysis** - More robust than single-ceiling methods
2. **Extract pitch from original** - Use `to_pitch()` for realistic F0
3. **Match bandwidths to formant frequency** - Higher formants = wider bandwidths
4. **Validate on synthetic vowels first** - Test round-trip accuracy
5. **Save intermediate results** - Keep formant data and KlattGrid parameters
6. **Compare spectrograms** - Visual inspection reveals synthesis quality
7. **Document parameters** - Record all synthesis settings for reproducibility

### ✗ DON'T

1. **Don't use single ceiling for diverse speakers** - Use FormantPath instead
2. **Don't ignore bandwidth** - Critical for natural synthesis
3. **Don't expect perfect reconstruction** - Some loss is inherent
4. **Don't synthesize consonants** - KlattGrid is vowel-focused
5. **Don't skip validation** - Always check round-trip accuracy
6. **Don't use extreme formant values** - Stay within natural ranges

## Troubleshooting

### Poor Resynthesis Quality

**Symptoms**: Synthetic speech sounds unnatural, buzzy, or robotic

**Solutions**:
- Verify F1 < F2 < F3 relationship
- Check bandwidth values (B1 ≈ 80, B2 ≈ 120, B3 ≈ 150 Hz)
- Ensure pitch is realistic (80-300 Hz)
- Use mean formants rather than single time point

### High Round-Trip Error

**Symptoms**: Extracted formants differ greatly from target (>20% error)

**Solutions**:
- Increase FormantPath candidates (`num_steps_up_down = 3L`)
- Adjust formant ceiling based on speaker
- Check for formant tracking errors (visual inspection)
- Use longer vowel duration (>300 ms)

### Segfault During Synthesis

**Symptoms**: R crashes when calling `kg$to_sound()`

**Solutions**:
- Always use `KlattGrid_createFromVowel()` (not empty constructor)
- Verify all formant values are positive and finite
- Check F1 < F2 < F3 constraint

## Performance Notes

### Memory Usage

Analysis-resynthesis stores multiple objects:

- Original Sound: ~1 MB per minute
- FormantPath: ~5× Formant size (5 candidates)
- KlattGrid: Small (<1 KB)
- Synthetic Sound: ~1 MB per minute

**Tip**: Extract formant values and discard FormantPath object to save memory.

### Computation Time

Approximate times for 1-minute audio:

- FormantPath analysis: 5-10 seconds
- KlattGrid synthesis: 1-2 seconds
- Total round-trip: 6-12 seconds

FormantPath dominates computation (proportional to number of candidates).

## Summary

**The analysis-resynthesis workflow enables**:
- Validation of formant tracking accuracy
- Systematic acoustic parameter manipulation
- Voice morphing and transformation
- Perceptual experiment stimulus generation

**Key functions**:
- `sound$to_formant_path()` - Robust analysis
- `fp$extract_formant()` - Get optimal track
- `KlattGrid_createFromVowel()` - Safe synthesis
- `as.data.frame()` - Extract formant values

**Typical pipeline**:
1. Load audio: `Sound("file.wav")`
2. Analyze: `fp <- sound$to_formant_path(num_steps_up_down=2L)`
3. Extract: `formant <- fp$extract_formant()`
4. Get values: `df <- as.data.frame(formant)`
5. Synthesize: `kg <- KlattGrid_createFromVowel(...)`
6. Compare: Visual/perceptual validation

## Further Reading

- `vignette("formantpath-robust-tracking")` - FormantPath details
- `vignette("speech-synthesis-klattgrid")` - KlattGrid synthesis
- `vignette("formant-analysis")` - Standard formant extraction
- Praat manual: [Sound synthesis](http://www.fon.hum.uva.nl/praat/manual/Sound__Synthesis___.html)

## Session Info

```{r session-info}
sessionInfo()
```
