---
title: "Auditory Modeling with Cochleagram and Excitation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditory Modeling with Cochleagram and Excitation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

pladdrr provides auditory modeling through the **Cochleagram** and **Excitation** objects. These model how the human auditory system processes acoustic information, as opposed to a purely physical (spectral) analysis.

### Key Concepts

- **Cochleagram**: Models the basilar membrane response in the cochlea using the **Bark scale** (0-25.6 Bark ≈ 0-20,000 Hz)
- **Excitation**: Represents auditory nerve firing patterns on the **ERB (Equivalent Rectangular Bandwidth) scale**
- **Loudness**: Perceptual loudness measured in sones (not physical intensity)
- **Perceptual Distance**: Quantifies how different two sounds "sound" to human listeners

```{r setup}
library(pladdrr)
library(ggplot2)
```

## Cochleagram Analysis

### Creating a Cochleagram

The cochleagram represents sound on a frequency scale that matches the human cochlea's tonotopic organization.

```{r cochleagram-basic}
# Load or create a sound
sound <- Sound$create_tone(frequency = 440, duration = 0.5, sampling_rate = 44100)

# Create cochleagram
cochlea <- sound$to_cochleagram(
  dt = 0.01,                    # Time step (10 ms)
  df = 0.1,                     # Frequency step in Bark
  window_length = 0.025,        # Analysis window (25 ms)
  forward_masking_time = 0.03   # Temporal masking (30 ms)
)
```

### Understanding the Bark Scale

The Bark scale is a perceptual frequency scale where:
- 1 Bark ≈ 1 critical band (frequency resolution of human hearing)
- 0 Bark = 0 Hz
- 13 Bark ≈ 1500 Hz (center of speech range)
- 25.6 Bark ≈ 20,000 Hz

```{r bark-conversion, eval=FALSE}
# Common frequency conversions
# F1 (250 Hz) ≈ 2.5 Bark
# F2 (2000 Hz) ≈ 15 Bark
# F3 (3000 Hz) ≈ 19 Bark
```

### Querying Cochleagram Values

```{r cochleagram-query}
# Get excitation at specific time and frequency
# 440 Hz ≈ 4.2 Bark
excitation_440hz <- cochlea$get_value_at_time_and_frequency(
  time = 0.25,      # At 250 ms
  freq_bark = 4.2   # ~440 Hz
)

print(paste("Excitation at 440 Hz:", round(excitation_440hz, 4)))
```

### Visualizing Cochleagrams

```{r cochleagram-plot}
# Export to matrix for plotting
cochlea_matrix <- as.matrix(cochlea$as_matrix())

# Convert to long format for ggplot2 (rows = frequency bands, columns = time frames)
times <- sapply(seq_len(cochlea$get_number_of_frames()), cochlea$get_time_from_column)
barks <- sapply(seq_len(cochlea$get_number_of_frequency_bands()), cochlea$get_frequency_from_row)
df_long <- expand.grid(Bark = barks, Time = times)
df_long$Excitation <- as.vector(cochlea_matrix)

# Plot
ggplot(df_long, aes(x = Time, y = Bark, fill = Excitation)) +
  geom_raster() +
  scale_fill_viridis_c(option = "inferno") +
  labs(
    title = "Cochleagram of 440 Hz Tone",
    x = "Time (s)",
    y = "Frequency (Bark)",
    fill = "Excitation"
  ) +
  theme_minimal()
```

### Advanced: Ear-Drum-Brain Model

For more realistic auditory modeling, use the EDB (Ear-Drum-Brain) method which includes synaptic processing:

```{r cochleagram-edb}
cochlea_edb <- sound$to_cochleagram_edb(
  dtime = 0.01,
  dfreq = 0.1,
  has_synapse = TRUE,         # Include synaptic adaptation
  replenishment_rate = 0.01,  # Neurotransmitter replenishment
  loss_rate = 0.1,            # Synaptic depletion
  return_rate = 0.05,         # Recovery rate
  reprocessing_rate = 0.01    # Reprocessing rate
)
```

The EDB model is more computationally intensive but provides:
- Synaptic adaptation (important for continuous sounds)
- More realistic temporal response
- Better modeling of masking effects

## Excitation Patterns

### Creating Excitation from Spectrum

Excitation patterns represent the instantaneous firing rate of auditory nerve fibers:

```{r excitation-spectrum}
# Create excitation from spectrum
spectrum <- sound$to_spectrum()
excitation <- spectrum$to_excitation(erb_density = 0.1)

# Get perceptual loudness
loudness <- excitation$get_loudness()
print(paste("Loudness:", round(loudness, 2), "sones"))
```

### Creating Excitation from Cochleagram

You can also extract excitation patterns at specific times from a cochleagram:

```{r excitation-cochlea}
# Extract excitation pattern at t = 0.25s
excitation_t <- cochlea$to_excitation(0.25)

# Query excitation at specific frequency
exc_value <- excitation_t$get_value_at_frequency(4.2)  # ~440 Hz
print(paste("Excitation at 440 Hz:", round(exc_value, 4)))
```

### Perceptual Distance

Quantify how different two sounds are perceptually:

```{r excitation-distance}
# Create two different sounds
sound1 <- Sound$create_tone(frequency = 440, duration = 0.2, sampling_rate = 22050)
sound2 <- Sound$create_tone(frequency = 550, duration = 0.2, sampling_rate = 22050)

# Get excitation patterns
exc1 <- sound1$to_spectrum()$to_excitation()
exc2 <- sound2$to_spectrum()$to_excitation()

# Calculate perceptual distance
distance <- exc1$get_distance(exc2)
print(paste("Perceptual distance:", round(distance, 4)))
```

Higher distance values indicate sounds that are more perceptually different.

## Clinical Applications

### Hearing Loss Simulation

Compare cochleagrams of normal and impaired hearing:

```{r hearing-loss}
# Original speech sound
sound_original <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))

# In practice, apply a low-pass filter to sound_filtered here to simulate
# high-frequency hearing loss (Sound objects have no clone() method, so
# start from a fresh load rather than copying sound_original)
sound_filtered <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))

# Create cochleagrams
cochlea_normal <- sound_original$to_cochleagram()
cochlea_impaired <- sound_filtered$to_cochleagram()

# Calculate perceptual difference
difference <- cochlea_normal$get_difference(
  cochlea_impaired,
  tmin = 0,
  tmax = 0  # Full duration
)

print(paste("Hearing loss impact (distance):", round(difference, 4)))
```

Without an actual filtering step, `sound_original` and `sound_filtered` are
identical here, so `difference` is 0; apply a real low-pass filter to
`sound_filtered` to get a non-zero result.

### Speech Intelligibility Prediction

Excitation patterns can predict speech intelligibility. This requires two
recordings of the same material (clean and noisy); the pattern below uses
the bundled test file for both, so `distance` will be 0.

```{r intelligibility}
# Clean speech
speech_clean <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))

# Noisy speech (in practice, a separate recording made in noise)
speech_noisy <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))

# Get excitation patterns
exc_clean <- speech_clean$to_spectrum()$to_excitation()
exc_noisy <- speech_noisy$to_spectrum()$to_excitation()

# Perceptual distance correlates with intelligibility loss
distance <- exc_clean$get_distance(exc_noisy)

# Lower distance = better preserved intelligibility
distance
```

### Loudness Recruitment Assessment

Track loudness perception across intensity levels:

```{r loudness-recruitment}
# Create sounds at different relative levels by scaling tone amplitude
intensities <- seq(40, 80, by = 10)  # relative dB SPL
loudnesses <- numeric(length(intensities))

for (i in seq_along(intensities)) {
  # Approximate a level in dB by scaling amplitude (20*log10(amplitude))
  amplitude <- min(0.99, 10 ^ ((intensities[i] - 80) / 20))
  sound <- Sound$create_tone(frequency = 1000, duration = 0.3,
                             amplitude = amplitude, sampling_rate = 44100)

  # Measure perceptual loudness
  excitation <- sound$to_spectrum()$to_excitation()
  loudnesses[i] <- excitation$get_loudness()
}

# Plot loudness growth function
plot(intensities, loudnesses,
     xlab = "Intensity (dB SPL)",
     ylab = "Loudness (sones)",
     main = "Loudness Growth Function",
     type = "b")
```

## Formant Extraction from Excitation

Excitation patterns provide an alternative method for formant extraction:

```{r formant-from-excitation}
# Create vowel sound
sound_vowel <- Sound$create_tone(duration = 0.2, sampling_rate = 22050)

# Get excitation pattern
excitation <- sound_vowel$to_spectrum()$to_excitation()

# Extract formants from excitation pattern
formant <- excitation$to_formant(max_formants = 5)

# Query formants
f1 <- formant$get_value_at_time(1, 0.1, unit = "hertz")
f2 <- formant$get_value_at_time(2, 0.1, unit = "hertz")

if (!is.na(f1) && !is.na(f2)) {
  print(paste("F1:", round(f1), "Hz"))
  print(paste("F2:", round(f2), "Hz"))
}
```

This method can be more robust than LPC for:
- Noisy speech
- Breathy voice quality
- Non-modal phonation

## Best Practices

### Choosing Parameters

**Cochleagram time step (dt)**:
- Speech analysis: 0.005-0.010 s (5-10 ms)
- Music: 0.010-0.020 s
- Sustained sounds: 0.020-0.050 s

**Frequency step (df)**:
- Detailed analysis: 0.05-0.1 Bark
- Standard analysis: 0.1-0.2 Bark
- Coarse analysis: 0.2-0.5 Bark

**Window length**:
- Shorter (0.005-0.010 s): Better time resolution
- Medium (0.020-0.030 s): Balanced
- Longer (0.050-0.100 s): Better frequency resolution

**Forward masking**:
- Standard: 0.03 s
- Strong masking: 0.05-0.1 s
- Minimal masking: 0.01-0.02 s

### Performance Considerations

```{r performance}
# For batch processing, reuse objects when possible
sounds <- list(sound1, sound2, sound_vowel)

# Vectorized approach: build the cochleagrams once, then query all of them
cochleagrams <- lapply(sounds, function(s) {
  s$to_cochleagram(dt = 0.01, df = 0.1)
})

# Extract loudness from all
loudnesses <- sapply(cochleagrams, function(c) {
  c$get_loudness_at_time(0.1)
})
```

### Memory Management

Cochleagrams and excitation patterns can be large. Clean up when done:

```{r memory}
# Process and extract only what you need
cochlea <- sound$to_cochleagram()
loudness_time_series <- sapply(seq(0, 0.5, by = 0.01), cochlea$get_loudness_at_time)

# Remove large object if no longer needed
rm(cochlea)
gc()  # Force garbage collection
```

## Comparison with Traditional Analysis

| Traditional | Auditory Modeling |
|------------|-------------------|
| Spectrogram (linear/log Hz) | Cochleagram (Bark scale) |
| Spectrum power (dB) | Excitation (nerve firing rate) |
| Sound level (dB SPL) | Loudness (sones) |
| Spectral distance (Euclidean) | Perceptual distance |

Auditory modeling is preferred when:
- Modeling human perception
- Clinical audiology applications
- Speech intelligibility prediction
- Psychoacoustic research
- Hearing aid algorithm development

## Further Reading

- **Bark scale**: Zwicker & Terhardt (1980)
- **ERB scale**: Glasberg & Moore (1990)
- **Cochleagram**: Patterson et al. (1992)
- **Excitation patterns**: Moore (2012) - *An Introduction to the Psychology of Hearing*

## Session Info

```{r session}
sessionInfo()
```
