---
title: "Composable Plotting with autoplot and autolayer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composable Plotting with autoplot and autolayer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-packages, message=FALSE}
library(pladdrr)
library(ggplot2)
```

## Overview

pladdrr provides `autoplot(
  )` and `autolayer()` methods following the ggplot2 convention for visualizing acoustic data.

- **`autoplot(object)`** - Creates a complete ggplot for a single Praat object
- **`autolayer(object)`** - Returns layers to add to an existing plot

This pattern allows combining multiple objects in one plot:

```r
autoplot(spec) +
  autolayer(pitch) +
  autolayer(formant)
```

## Basic Usage: Single Object Plots

Each Praat object type has an `autoplot(
  )` method that creates an appropriate visualization.

```{r load-sound}
# Load example sound
sound <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))
```
### Sound Waveform

```{r autoplot-sound}
autoplot(sound)
```

### Pitch Contour

```{r autoplot-pitch}
pitch <- sound$to_pitch()
autoplot(pitch)
```

### Spectrogram

```{r autoplot-spectrogram}
spec <- sound$to_spectrogram()
autoplot(spec, to_freq = 5000)
```

### Spectrum

```{r autoplot-spectrum}
spectrum <- sound$to_spectrum()
autoplot(spectrum, to_freq = 5000)
```

### Intensity

```{r autoplot-intensity}
intensity <- sound$to_intensity()
autoplot(intensity)
```

### Formants

```{r autoplot-formant}
# Formant extraction may fail on some sounds (e.g., if too short or no voiced
#  content)
formant <- tryCatch(
  sound$to_formant_burg(),
  error = function(e) {
    message("Formant extraction failed, using synthesized vowel")
    kg <- klattgrid_create_from_vowel(duration = 0.3, f0start = 120)
    kg$to_sound()$to_formant_burg()
  }
)
autoplot(formant, max_formant = 4)
```

### Pitch Tier

A `PitchTier` stores sparse, hand-editable F0 points (as opposed to `Pitch`'s
dense per-frame contour) — it's what you get from `pitch$down_to_pitch_tier()`,
and it's the object type Praat scripts manipulate when resynthesizing prosody.

```{r autoplot-pitchtier, eval = requireNamespace("ggplot2", quietly = TRUE)}
pitch_tier <- pitch$down_to_pitch_tier()
autoplot(pitch_tier)
```

Overlay the sparse tier points on the dense pitch contour with `autolayer()`:

```{r pitch-pitchtier-overlay, eval = requireNamespace("ggplot2", quietly = TRUE)}
autoplot(pitch) +
  autolayer(pitch_tier, color = "red")
```

### Formant Grid and Formant Tier Styles

A `FormantGrid` holds formant frequency points directly (the editable
structure behind Klatt synthesis), while a `FormantTier` is a downsampled
view of a `Formant` object. `FormantTier`'s `style` argument switches between
`"speckle"` (default, one point per Praat-selected candidate) and `"line"`
(one interpolated point per fixed time step).

```{r autoplot-formantgrid, eval = requireNamespace("ggplot2", quietly = TRUE)}
grid <- FormantGrid(tmin = 0, tmax = 1, number_of_formants = 3)
grid$add_formant_point(1, 0.5, 500)
grid$add_formant_point(2, 0.5, 1500)
grid$add_formant_point(3, 0.5, 2500)
autoplot(grid)
```

```{r autoplot-formanttier-styles, eval = requireNamespace("ggplot2", quietly = TRUE)}
formant_tier <- FormantTier$from_formant(formant)
autoplot(formant_tier)                  # style = "speckle" (default)
autoplot(formant_tier, style = "line")  # interpolated at a fixed time step
```

### Cepstral Analysis

`Cepstrum` supports two views of the same data via its `power` argument: the
default raw (signed) cepstrum, and a `power = TRUE` dB view used for peak
picking. `PowerCepstrogram` extends this over time and underlies CPPS-based
voice-quality measures.

```{r autoplot-cepstrum-power, eval = requireNamespace("ggplot2", quietly = TRUE)}
cepstrum <- sound$to_cepstrum()
autoplot(cepstrum)                # raw view
autoplot(cepstrum, power = TRUE)  # power (dB) view
```

```{r autoplot-powercepstrogram, eval = requireNamespace("ggplot2", quietly = TRUE)}
power_cepstrogram <- sound$to_powercepstrogram()
autoplot(power_cepstrogram)
```

## Composing Multi-Layer Plots

Combine objects using `autolayer()`.

### Spectrogram with Pitch Overlay

```{r spectrogram-pitch}
autoplot(spec, to_freq = 5000) +
  autolayer(pitch, color = "cyan", geom = "point")
```

### Spectrogram with Formant Tracks

```{r spectrogram-formant}
autoplot(spec, to_freq = 5000) +
  autolayer(formant, max_formant = 3)
```

### Spectrogram with Pitch and Formants

```{r spectrogram-pitch-formant}
autoplot(spec, to_freq = 5000) +
  autolayer(formant, max_formant = 3) +
  autolayer(pitch, color = "white", geom = "point")
```

### Sound with Pitch Overlay

Overlay pitch points on waveform (useful for seeing voicing):

```{r sound-pitch}
# Scale pitch to amplitude range for overlay
pitch_df <- pitch$as_data_frame()
pitch_df <- pitch_df[!is.na(pitch_df$frequency) & pitch_df$frequency > 0, ]

autoplot(sound) +
  geom_point(
    data = pitch_df,
    aes(x = time, y = 0),
    color = "red", size = 0.5, alpha = 0.5
  )
```

## Customization with ggplot2

All autoplot/autolayer methods return ggplot2 objects, so you can customize freely.

### Custom Themes

```{r custom-theme}
autoplot(pitch) +
  theme_classic() +
  labs(
    title = "Fundamental Frequency",
    subtitle = "Extracted using autocorrelation method"
  )
```

### Custom Colors

```{r custom-colors}
autoplot(spec, to_freq = 5000) +
  scale_fill_viridis_c(option = "magma") +
  theme_dark()
```

### Faceting by Time Windows

```{r facet-example}
# Create intensity plot with reference lines
autoplot(intensity) +
  geom_hline(yintercept = c(50, 60, 70), linetype = "dashed", alpha = 0.3) +
  annotate("text", x = 0, y = 70, label = "Loud", hjust = 0, size = 3) +
  annotate("text", x = 0, y = 50, label = "Quiet", hjust = 0, size = 3)
```

## Statistical and Alignment Tools

pladdrr also wraps a few non-acoustic-domain Praat objects used in analysis
pipelines: PCA for dimensionality reduction, and DTW for aligning two sounds.

### PCA: Scree, Scores, and Combined Plots

`autoplot.PCA()`'s `type` argument selects a scree plot (variance explained
per component), a scores plot (observations projected onto the first two
components), or `"both"` combined side-by-side via the optional patchwork
package.

```{r autoplot-pca, eval = requireNamespace("ggplot2", quietly = TRUE)}
set.seed(1)
formant_matrix <- matrix(rnorm(200), nrow = 20)
pca <- pca_from_matrix(formant_matrix)
autoplot(pca, type = "scree")
autoplot(pca, type = "scores")
```

```{r autoplot-pca-both, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE)}
autoplot(pca, type = "both")
```

### DTW Warping Path

Dynamic Time Warping aligns two sounds (or feature sequences) that differ in
timing — `autoplot.DTW()` plots the resulting warping path, and
`autolayer.DTW()` overlays it on an existing plot.

```{r autoplot-dtw, eval = requireNamespace("ggplot2", quietly = TRUE)}
dtw_sound <- generate_sine_wave(440, 0.1, sampling_rate = 16000)
dtw <- sounds_to_dtw(dtw_sound, dtw_sound)
autoplot(dtw)
```

```{r autolayer-dtw, eval = requireNamespace("ggplot2", quietly = TRUE)}
autoplot(dtw) +
  autolayer(dtw, color = "orange")
```

## Parameter Reference

### Common Parameters

| Parameter | Description | Default |
|-----------|-------------|---------|
| `from_time` | Start time (seconds) | NULL (start) |
| `to_time` | End time (seconds) | NULL (end) |
| `color` | Line/point color | varies |

### Spectrogram Parameters

| Parameter | Description | Default |
|-----------|-------------|---------|
| `from_freq` | Min frequency (Hz) | NULL (0) |
| `to_freq` | Max frequency (Hz) | NULL (Nyquist) |
| `dynamic_range` | Dynamic range (dB) | 70 |

### Formant Parameters

| Parameter | Description | Default |
|-----------|-------------|---------|
| `max_formant` | Max formant number | 3 |
| `colors` | Color vector for formants | auto |

### Pitch autolayer Parameters

| Parameter | Description | Default |
|-----------|-------------|---------|
| `geom` | "line" or "point" | "line" |

## Comparison with plot() Methods

pladdrr also provides traditional `plot()` S3 methods. Here's when to use each:

| Use Case | Recommended |
|----------|-------------|
| Quick single-object plot | `plot(object)` or `autoplot(object)` |
| Multi-object composition | `autoplot() + autolayer()` |
| Maximum customization | `autoplot()` (returns ggplot2) |
| Base R graphics pipeline | `plot()` |

## Saving Plots

```{r save-example, eval=FALSE}
# Create composed plot
p <- autoplot(spec, to_freq = 5000) +
  autolayer(formant, max_formant = 3) +
  autolayer(pitch, color = "cyan") +
  theme_minimal() +
  labs(title = "Acoustic Analysis")

# Save as PDF (vector)
ggsave("analysis.pdf", p, width = 10, height = 6)

# Save as PNG (high-res)
ggsave("analysis.png", p, width = 10, height = 6, dpi = 300)
```

## Summary

The autoplot/autolayer pattern provides:

- **Composability** - Layer multiple acoustic objects
- **Consistency** - Same interface for all object types
- **Customization** - Full ggplot2 compatibility
- **Publication quality** - Vector graphics output

For more visualization options, see `vignette("visualization")`.
