---
title: "Advanced Formant Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Formant Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

pladdrr provides multiple methods for formant extraction. This vignette covers when to use each method, how to troubleshoot formant detection issues, and parameter tuning for formant analysis.

```{r setup}
library(pladdrr)
library(ggplot2)
```

## Formant Extraction Methods

### Available Methods

1. **Burg** (`to_formant_burg`) - LPC-based, recommended for general use
2. **Keep All** (`to_formant_keepall`) - Retains all LPC formants without filtering
3. **Robust** (`to_formant_robust`) - Outlier-resistant iterative refit of Burg
4. **Willems** (`to_formant_willems`) - Alternative LPC formant tracker
5. **Split-Levinson** (`to_formant_sl`) - LPC via split-Levinson recursion
6. **From Excitation** (`excitation$to_formant`) - Perceptual-based extraction, via `Spectrum$to_excitation()` or `Cochleagram$to_excitation()`

### Method Comparison

| Method | Speed | Accuracy | Best For |
|--------|-------|----------|----------|
| **Burg** | Fast | Good | General speech analysis (recommended) |
| **Keep All** | Fast | Variable | Research, exploration, comparison |
| **Robust** | Slower | Good | Recordings with occasional outlier frames |
| **Excitation** | Slow | Good | Noisy/breathy speech |

## 1. Burg's Method (Standard)

The most commonly used method, based on the Burg algorithm for LPC analysis:

```{r burg-basic, error=TRUE}
# Create a synthesized vowel sound (pure tones don't have formants)
# KlattGrid synthesizes speech with formant structure
# Use a longer duration (1s) for stable analysis
kg <- tryCatch({
  KlattGrid_createFromVowel(
    duration = 1.0,          # Longer duration for stable analysis
    f0start = 120,           # Fundamental frequency
    f1 = 730, b1 = 90,       # F1 for /a/ vowel
    f2 = 1090, b2 = 110,     # F2
    f3 = 2440, b3 = 170      # F3
  )
}, error = function(e) {
  message("KlattGrid synthesis failed: ", e$message)
  NULL
})

if (!is.null(kg)) {
  sound <- kg$to_sound()
  
  # Standard Burg formant extraction with error handling
  formant_burg <- tryCatch({
    sound$to_formant_burg(
      time_step = 0.005,                   # 5 ms steps
      max_formants = 5,                    # Track up to F5
      max_frequency = 5500,                # Ceiling (Hz)
      window_length = 0.025,               # 25 ms analysis window
      pre_emphasis_from = 50               # Pre-emphasis from 50 Hz
    )
  }, error = function(e) {
    message("Formant extraction failed: ", e$message)
    message("Note: This requires completed polynomial root finding implementation")
    NULL
  })
  
  # Query formants at midpoint if extraction succeeded
  if (!is.null(formant_burg)) {
    f1 <- formant_burg$get_value_at_time(1, 0.5, unit = "hertz")
    f2 <- formant_burg$get_value_at_time(2, 0.5, unit = "hertz")
    
    if (!is.na(f1) && !is.na(f2)) {
      print(paste("F1:", round(f1), "Hz (expected ~730)"))
      print(paste("F2:", round(f2), "Hz (expected ~1090)"))
    }
  } else {
    message("Skipping formant queries - extraction not available")
  }
} else {
  message("Skipping formant analysis - synthesis not available")
}
```

### When to Use Burg

Use Burg for:
- Standard speech analysis (recommended default)
- Fast processing
- General vowel formant tracking
- Batch processing of many files

Consider alternatives for:
- Very noisy recordings (try Excitation-based method)
- Exploratory research (consider Keep All to see all detected formants)

## 2. Keep All Method

Retains all detected LPC formants without filtering:

```{r keepall}
formant_keepall <- sound$to_formant_keepall(
  time_step = 0.005,
  max_formants = 5,                    # Maximum formants to track
  max_frequency = 5500,
  window_length = 0.025,
  pre_emphasis_from = 50
)
formant_keepall
```

### When to Use Keep All

Use Keep All for:
- Exploratory analysis
- Research applications
- Comparison with the Burg method

Avoid Keep All for:
- Cases where you need only specific formants filtered out

### Keep All vs. Burg Example

```{r compare-keepall-burg}
# Same sound, both methods
formant_b <- sound$to_formant_burg(max_formants = 4)
formant_k <- sound$to_formant_keepall(max_formants = 4)

# Extract F1 and F2
time <- 0.25
f1_burg <- formant_b$get_value_at_time(1, time, unit = "hertz")
f1_keep <- formant_k$get_value_at_time(1, time, unit = "hertz")

f2_burg <- formant_b$get_value_at_time(2, time, unit = "hertz")
f2_keep <- formant_k$get_value_at_time(2, time, unit = "hertz")

# Comparison
if (!any(is.na(c(f1_burg, f1_keep, f2_burg, f2_keep)))) {
  cat("Method Comparison:\n")
  cat(sprintf("  Burg:     F1 = %d Hz, F2 = %d Hz\n", round(f1_burg), round(f2_burg)))
  cat(sprintf("  Keep All: F1 = %d Hz, F2 = %d Hz\n", round(f1_keep), round(f2_keep)))
  cat(sprintf("  Diff:     F1 = %d Hz, F2 = %d Hz\n",
              round(abs(f1_burg - f1_keep)),
              round(abs(f2_burg - f2_keep))))
}
```

## 3. Alternative Methods (Advanced)

`to_formant_willems()` and `to_formant_sl()` (split-Levinson) provide alternative
LPC-based extraction. `to_formant_robust()` iteratively refits Burg while down-weighting
outlier frames (`num_std_dev`, `max_iterations`). Burg (recommended) or Keep All cover
most use cases; these are for comparison or specific research needs.

```{r robust-basic}
formant_robust <- sound$to_formant_robust(
  time_step = 0.005,
  max_formants = 5,
  max_frequency = 5500,
  window_length = 0.025,
  pre_emphasis_from = 50,
  num_std_dev = 1.5,       # outlier threshold in standard deviations
  max_iterations = 5
)
formant_robust
```

## Choosing Formant Ceiling

The maximum formant frequency affects which formants are tracked:

### Speaker-Specific Ceilings

```{r formant-ceiling, error=TRUE}
# Speaker-specific formant ceilings demonstration
# Only run if sound was created successfully
if (exists("sound") && !is.null(sound)) {
  # Adult male
  formant_male <- tryCatch(
    sound$to_formant_burg(max_frequency = 5000),
    error = function(e) { message("Male ceiling failed: ", e$message); NULL }
  )

  # Adult female
  formant_female <- tryCatch(
    sound$to_formant_burg(max_frequency = 5500),
    error = function(e) { message("Female ceiling failed: ", e$message); NULL }
  )

  # Child
  formant_child <- tryCatch(
    sound$to_formant_burg(max_frequency = 8000),
    error = function(e) { message("Child ceiling failed: ", e$message); NULL }
  )

  # Report results
  cat("Formant extraction results:\n")
  cat("  Male (5000 Hz):", if(!is.null(formant_male)) "OK" else "failed", "\n")
  cat("  Female (5500 Hz):", if(!is.null(formant_female)) "OK" else "failed", "\n")
  cat("  Child (8000 Hz):", if(!is.null(formant_child)) "OK" else "failed", "\n")
} else {
  message("Skipping ceiling demo - sound not available")
}
```

### Guidelines

| Speaker | F5 Ceiling | Reasoning |
|---------|-----------|-----------|
| Adult male | 5000 Hz | Lower formants, longer vocal tract |
| Adult female | 5500 Hz | Higher formants, shorter vocal tract |
| Child | 7000-8000 Hz | Much higher formants |
| Infant | 9000-10000 Hz | Very high formants |

**Rule of thumb**: Ceiling ≈ (1000 × (number of formants + 1)) Hz

## Parameter Tuning

### Window Length

```{r window-length, error=TRUE}
if (exists("sound") && !is.null(sound)) {
  # Short window (better time resolution)
  formant_short <- tryCatch(
    sound$to_formant_burg(window_length = 0.010),
    error = function(e) { message("10ms window failed"); NULL }
  )

  # Standard window (balanced)
  formant_std <- tryCatch(
    sound$to_formant_burg(window_length = 0.025),
    error = function(e) { message("25ms window failed"); NULL }
  )

  # Long window (better frequency resolution)
  formant_long <- tryCatch(
    sound$to_formant_burg(window_length = 0.050),
    error = function(e) { message("50ms window failed"); NULL }
  )
} else {
  message("Skipping window-length demo - sound not available")
}
```

**Recommendations**:
- **Vowels** (sustained): 0.025-0.050 s
- **Consonants** (rapid): 0.010-0.020 s
- **General speech**: 0.025 s (default)
- **Singing**: 0.050-0.100 s

### Time Step

```{r time-step, error=TRUE}
if (exists("sound") && !is.null(sound)) {
  # Coarse (fast, less detail)
  formant_coarse <- tryCatch(
    sound$to_formant_burg(time_step = 0.010),
    error = function(e) { message("10ms step failed"); NULL }
  )

  # Standard (good balance)
  formant_std <- tryCatch(
    sound$to_formant_burg(time_step = 0.005),
    error = function(e) { message("5ms step failed"); NULL }
  )

  # Fine (slow, more detail)
  formant_fine <- tryCatch(
    sound$to_formant_burg(time_step = 0.002),
    error = function(e) { message("2ms step failed"); NULL }
  )
} else {
  message("Skipping time-step demo - sound not available")
}
```

**Recommendations**:
- **Static vowels**: 0.010 s
- **Conversational speech**: 0.005 s
- **Rapid transitions**: 0.002 s

### Number of Formants

```{r num-formants, error=TRUE}
if (exists("sound") && !is.null(sound)) {
  # Track only F1-F2 (faster)
  # Note: Low max_formants can fail on some sounds due to polynomial constraints
  formant_2 <- tryCatch(
    sound$to_formant_burg(max_formants = 2),
    error = function(e) tryCatch(
      sound$to_formant_burg(max_formants = 4),
      error = function(e2) { message("F2 tracking failed"); NULL }
    )
  )

  # Standard (F1-F5) - most reliable
  formant_5 <- tryCatch(
    sound$to_formant_burg(max_formants = 5),
    error = function(e) { message("F5 tracking failed"); NULL }
  )

  # Extended (up to F7, slower)
  formant_7 <- tryCatch(
    sound$to_formant_burg(max_formants = 7, max_frequency = 7000),
    error = function(e) tryCatch(
      sound$to_formant_burg(max_formants = 5, max_frequency = 5500),
      error = function(e2) { message("F7 tracking failed"); NULL }
    )
  )
} else {
  message("Skipping num-formants demo - sound not available")
}
```

## Troubleshooting Formant Detection

### Problem: No Formants Detected

```{r no-formants, eval=FALSE}
formant <- sound$to_formant_burg()
f1 <- formant$get_value_at_time(1, 0.1, unit = "hertz")
# Returns: NA
```

**Solutions**:

1. **Lower the ceiling**:
```{r lower-ceiling, eval=FALSE}
# Try lower maximum frequency
formant <- sound$to_formant_burg(max_frequency = 4500)
```

2. **Increase window length**:
```{r increase-window, eval=FALSE}
formant <- sound$to_formant_burg(window_length = 0.050)
```

3. **Check if sound is silent**:
```{r check-silent, eval=FALSE}
intensity <- sound$to_intensity()
mean_intensity <- intensity$get_mean(0, 0)  # in dB
# If < 30 dB, may be too quiet
```

### Problem: Formants Jump Erratically

```{r erratic, eval=FALSE}
# F1 jumps from 500 Hz to 2000 Hz between frames
```

**Solutions**:

1. **Use robust tracking**:
```{r robust, eval=FALSE}
formant <- sound$to_formant_robust(
  num_std_dev = 1.5,       # outlier threshold in standard deviations
  max_iterations = 5       # refit iterations
)
```

2. **Increase window length**:
```{r smooth-window, eval=FALSE}
formant <- sound$to_formant_burg(window_length = 0.040)
```

3. **Reduce time step**:
```{r reduce-step, eval=FALSE}
formant <- sound$to_formant_burg(time_step = 0.002)
```

### Problem: F3 Detected as F2

```{r wrong-formant, eval=FALSE}
# F2 shows values ~3000 Hz (likely F3)
```

**Solutions**:

1. **Adjust formant ceiling**:
```{r adjust-ceiling, eval=FALSE}
# Lower ceiling may force re-evaluation
formant <- sound$to_formant_burg(max_frequency = 4800)
```

2. **Try Keep All method for comparison**:
```{r try-keepall, eval=FALSE}
formant <- sound$to_formant_keepall(max_formants = 5)
```

3. **Check bandwidth**:
```{r check-bandwidth, eval=FALSE}
bandwidth <- formant$get_bandwidth_at_time(2, 0.1, unit = "hertz")
# Very wide bandwidth may indicate spurious formant
```

## Vowel Space Analysis

### Extracting F1-F2 for Vowel Plot

```{r vowel-space}
# Analyze multiple vowel tokens. In practice, pass paths to your own
# recordings of each vowel; the test file is reused here for illustration.
wav_path <- system.file("extdata", "test.wav", package = "pladdrr")
vowels <- rep(wav_path, 3)
vowel_labels <- c("token1", "token2", "token3")

vowel_data <- data.frame()

for (i in seq_along(vowels)) {
  sound <- Sound$new(vowels[i])
  formant <- sound$to_formant_burg()
  
  # Extract at vowel midpoint
  duration <- sound$get_total_duration()
  midpoint <- duration / 2
  
  f1 <- formant$get_value_at_time(1, midpoint, unit = "hertz")
  f2 <- formant$get_value_at_time(2, midpoint, unit = "hertz")
  
  vowel_data <- rbind(vowel_data, data.frame(
    vowel = vowel_labels[i],
    F1 = f1,
    F2 = f2
  ))
}

# Plot vowel space
ggplot(vowel_data, aes(x = F2, y = F1, label = vowel)) +
  geom_point(size = 3) +
  geom_text(vjust = -1) +
  scale_x_reverse() +  # F2 decreases left to right
  scale_y_reverse() +  # F1 decreases bottom to top
  labs(title = "Vowel Space",
       x = "F2 (Hz)", y = "F1 (Hz)") +
  theme_minimal()
```

### Formant Tracking Over Time

```{r formant-tracking}
sound <- Sound$new(wav_path)
formant <- sound$to_formant_burg(time_step = 0.005)

# Extract F1 and F2 across time
times <- seq(0, sound$get_total_duration(), by = 0.005)
f1_values <- sapply(times, function(t) {
  formant$get_value_at_time(1, t, unit = "hertz")
})
f2_values <- sapply(times, function(t) {
  formant$get_value_at_time(2, t, unit = "hertz")
})

# Plot formant tracks
df <- data.frame(
  Time = rep(times, 2),
  Frequency = c(f1_values, f2_values),
  Formant = rep(c("F1", "F2"), each = length(times))
)

ggplot(df, aes(x = Time, y = Frequency, color = Formant)) +
  geom_line(linewidth = 1) +
  labs(title = "Formant Tracks",
       x = "Time (s)", y = "Frequency (Hz)") +
  theme_minimal()
```

## Advanced: Formant from Excitation

For noisy or breathy speech, excitation-based formant extraction is an alternative:

```{r excitation-formant}
# For noisy recordings, use excitation-based formant extraction
sound <- Sound$new(wav_path)

# Method 1: Via Spectrum
spectrum <- sound$to_spectrum()
excitation <- spectrum$to_excitation()
formant_exc <- excitation$to_formant(max_formants = 5)

# Method 2: Via Cochleagram (more complex but robust)
cochlea <- sound$to_cochleagram()
excitation2 <- cochlea$to_excitation(0.15)  # At specific time
formant_exc2 <- excitation2$to_formant(max_formants = 5)
```

**When to use**:
- Breathy voice
- Noisy recordings
- Pathological voice
- Whispered speech

## Best Practices Summary

### General Recommendations

1. **Start with Burg** - Recommended default for most cases
2. **Compare with Keep All** - If results look suspicious
3. **Adjust ceiling** - Based on speaker demographics
4. **Use appropriate time step** - Balance speed vs. detail
5. **Check intensity** - Low intensity → unreliable formants
6. **Try Excitation method** - For very noisy recordings

### Quality Checks

```{r quality-checks, eval=FALSE}
# 1. Check if formants are in expected ranges
f1 <- formant$get_value_at_time(1, time, unit = "hertz")
f2 <- formant$get_value_at_time(2, time, unit = "hertz")

if (!is.na(f1) && !is.na(f2)) {
  # F1 should be < F2
  if (f1 >= f2) warning("F1 >= F2, likely error")
  
  # Formants should be in typical ranges
  if (f1 < 200 || f1 > 1200) warning("F1 out of typical range")
  if (f2 < 700 || f2 > 3500) warning("F2 out of typical range")
}

# 2. Check bandwidth (shouldn't be too large)
b1 <- formant$get_bandwidth_at_time(1, time, unit = "hertz")
if (!is.na(b1) && b1 > 400) warning("Large F1 bandwidth")

# 3. Check formant continuity
f1_prev <- formant$get_value_at_time(1, time - 0.010, unit = "hertz")
if (!is.na(f1) && !is.na(f1_prev)) {
  jump <- abs(f1 - f1_prev)
  if (jump > 300) warning("Large formant jump (>300 Hz)")
}
```

## Session Info

```{r session}
sessionInfo()
```
