---
title: "Advanced Formant Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Formant Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

pladdrr provides multiple methods for formant extraction, each with different strengths and optimal use cases. This vignette explains when to use each method, how to troubleshoot formant detection issues, and best practices for robust formant analysis.

```{r setup}
library(pladdrr)
library(ggplot2)
```

## Formant Extraction Methods

### Available Methods

1. **Burg** (`to_formant_burg`) - Most common, good general-purpose
2. **Willems** (`to_formant_willems`) - More accurate for specific number of formants
3. **Split-Levinson** (`to_formant_sl`) - Alternative algorithm for verification
4. **Keep All** (`to_formant_keepall`) - Retains all LPC formants
5. **From Excitation** (`excitation$to_formant`) - Perceptual-based extraction

### Method Comparison

| Method | Speed | Accuracy | Best For |
|--------|-------|----------|----------|
| **Burg** | Fast | Good | General speech analysis |
| **Willems** | Medium | Better | Synthesis, specific F count |
| **Split-Levinson** | Fast | Good | Verification, comparison |
| **Keep All** | Fast | Variable | Research, exploration |
| **Excitation** | Slow | Good | Noisy/breathy speech |

## 1. Burg's Method (Standard)

The most commonly used method, based on the Burg algorithm for LPC analysis:

```{r burg-basic}
# Create or load a sound
sound <- Sound$create_tone(duration = 0.5, sampling_rate = 22050)

# Standard Burg formant extraction
formant_burg <- sound$to_formant_burg(
  time_step = 0.005,                   # 5 ms steps
  max_formants = 5,                    # Track up to F5
  max_frequency = 5500,                # Ceiling (Hz)
  window_length = 0.025,               # 25 ms analysis window
  pre_emphasis_from = 50               # Pre-emphasis from 50 Hz
)

# Query formants
f1 <- formant_burg$get_value_at_time(1, 0.25, unit = "hertz")
f2 <- formant_burg$get_value_at_time(2, 0.25, unit = "hertz")

if (!is.na(f1) && !is.na(f2)) {
  print(paste("F1:", round(f1), "Hz"))
  print(paste("F2:", round(f2), "Hz"))
}
```

### When to Use Burg

✅ **Use Burg for**:
- Standard speech analysis
- When you need fast processing
- General vowel formant tracking
- Batch processing of many files

❌ **Avoid Burg for**:
- Exact formant count requirements (use Willems)
- Verification (use Split-Levinson as second opinion)

## 2. Willems' Method

Optimized for extracting a specific number of formants:

```{r willems}
formant_willems <- sound$to_formant_willems(
  time_step = 0.005,
  number_of_formants = 5,              # Exact number to extract
  max_frequency = 5500,
  window_length = 0.025,
  pre_emphasis_from = 50
)
```

### When to Use Willems

✅ **Use Willems for**:
- When you know exact formant count needed
- Formant synthesis applications
- Better bandwidth estimation
- More accurate formant tracking

❌ **Avoid Willems for**:
- Exploratory analysis (formant count unknown)
- When speed is critical

### Willems vs. Burg Example

```{r compare-willems-burg}
# Same sound, both methods
formant_b <- sound$to_formant_burg(max_formants = 4)
formant_w <- sound$to_formant_willems(number_of_formants = 4)

# Extract F1 and F2
time <- 0.25
f1_burg <- formant_b$get_value_at_time(1, time, unit = "hertz")
f1_will <- formant_w$get_value_at_time(1, time, unit = "hertz")

f2_burg <- formant_b$get_value_at_time(2, time, unit = "hertz")
f2_will <- formant_w$get_value_at_time(2, time, unit = "hertz")

# Comparison
if (!any(is.na(c(f1_burg, f1_will, f2_burg, f2_will)))) {
  cat("Method Comparison:\n")
  cat(sprintf("  Burg:    F1 = %d Hz, F2 = %d Hz\n", round(f1_burg), round(f2_burg)))
  cat(sprintf("  Willems: F1 = %d Hz, F2 = %d Hz\n", round(f1_will), round(f2_will)))
  cat(sprintf("  Diff:    F1 = %d Hz, F2 = %d Hz\n",
              round(abs(f1_burg - f1_will)),
              round(abs(f2_burg - f2_will))))
}
```

## 3. Split-Levinson Method

Alternative algorithm useful for verification:

```{r split-levinson}
formant_sl <- sound$to_formant_sl(
  time_step = 0.005,
  number_of_poles = 10,                # Number of LPC poles
  max_frequency = 5500,
  window_length = 0.025,
  pre_emphasis_from = 50
)
```

### When to Use Split-Levinson

✅ **Use Split-Levinson for**:
- Verifying results from other methods
- Comparison studies
- Research applications
- When Burg produces unexpected results

## Choosing Formant Ceiling

The maximum formant frequency affects which formants are tracked:

### Speaker-Specific Ceilings

```{r formant-ceiling}
# Adult male
formant_male <- sound$to_formant_burg(
  max_frequency = 5000    # Men: 5000 Hz
)

# Adult female
formant_female <- sound$to_formant_burg(
  max_frequency = 5500    # Women: 5500 Hz
)

# Child
formant_child <- sound$to_formant_burg(
  max_frequency = 8000    # Children: 7000-8000 Hz
)
```

### Guidelines

| Speaker | F5 Ceiling | Reasoning |
|---------|-----------|-----------|
| Adult male | 5000 Hz | Lower formants, longer vocal tract |
| Adult female | 5500 Hz | Higher formants, shorter vocal tract |
| Child | 7000-8000 Hz | Much higher formants |
| Infant | 9000-10000 Hz | Very high formants |

**Rule of thumb**: Ceiling ≈ (1000 × (number of formants + 1)) Hz

## Parameter Tuning

### Window Length

```{r window-length}
# Short window (better time resolution)
formant_short <- sound$to_formant_burg(window_length = 0.010)  # 10 ms

# Standard window (balanced)
formant_std <- sound$to_formant_burg(window_length = 0.025)    # 25 ms

# Long window (better frequency resolution)
formant_long <- sound$to_formant_burg(window_length = 0.050)   # 50 ms
```

**Recommendations**:
- **Vowels** (sustained): 0.025-0.050 s
- **Consonants** (rapid): 0.010-0.020 s
- **General speech**: 0.025 s (default)
- **Singing**: 0.050-0.100 s

### Time Step

```{r time-step}
# Coarse (fast, less detail)
formant_coarse <- sound$to_formant_burg(time_step = 0.010)  # 10 ms

# Standard (good balance)
formant_std <- sound$to_formant_burg(time_step = 0.005)     # 5 ms

# Fine (slow, more detail)
formant_fine <- sound$to_formant_burg(time_step = 0.002)    # 2 ms
```

**Recommendations**:
- **Static vowels**: 0.010 s
- **Conversational speech**: 0.005 s
- **Rapid transitions**: 0.002 s

### Number of Formants

```{r num-formants}
# Track only F1-F2 (faster)
formant_2 <- sound$to_formant_burg(max_formants = 2)

# Standard (F1-F5)
formant_5 <- sound$to_formant_burg(max_formants = 5)

# Extended (up to F7, slower)
formant_7 <- sound$to_formant_burg(
  max_formants = 7,
  max_frequency = 7000
)
```

## Troubleshooting Formant Detection

### Problem: No Formants Detected

```{r no-formants, eval=FALSE}
formant <- sound$to_formant_burg()
f1 <- formant$get_value_at_time(1, 0.1, unit = "hertz")
# Returns: NA
```

**Solutions**:

1. **Lower the ceiling**:
```{r lower-ceiling, eval=FALSE}
# Try lower maximum frequency
formant <- sound$to_formant_burg(max_frequency = 4500)
```

2. **Increase window length**:
```{r increase-window, eval=FALSE}
formant <- sound$to_formant_burg(window_length = 0.050)
```

3. **Check if sound is silent**:
```{r check-silent, eval=FALSE}
intensity <- sound$to_intensity()
mean_intensity <- intensity$get_mean(0, 0, unit = "dB")
# If < 30 dB, may be too quiet
```

### Problem: Formants Jump Erratically

```{r erratic, eval=FALSE}
# F1 jumps from 500 Hz to 2000 Hz between frames
```

**Solutions**:

1. **Use robust tracking** (if available):
```{r robust, eval=FALSE}
formant <- sound$to_formant_robust(
  smoothing_window = 0.05,      # Smooth over 50 ms
  max_formant_jump = 200        # Max 200 Hz jump
)
```

2. **Increase window length**:
```{r smooth-window, eval=FALSE}
formant <- sound$to_formant_burg(window_length = 0.040)
```

3. **Reduce time step**:
```{r reduce-step, eval=FALSE}
formant <- sound$to_formant_burg(time_step = 0.002)
```

### Problem: F3 Detected as F2

```{r wrong-formant, eval=FALSE}
# F2 shows values ~3000 Hz (likely F3)
```

**Solutions**:

1. **Adjust formant ceiling**:
```{r adjust-ceiling, eval=FALSE}
# Lower ceiling may force re-evaluation
formant <- sound$to_formant_burg(max_frequency = 4800)
```

2. **Try Willems method**:
```{r try-willems, eval=FALSE}
formant <- sound$to_formant_willems(number_of_formants = 5)
```

3. **Check bandwidth**:
```{r check-bandwidth, eval=FALSE}
bandwidth <- formant$get_bandwidth_at_time(2, 0.1, unit = "hertz")
# Very wide bandwidth may indicate spurious formant
```

## Vowel Space Analysis

### Extracting F1-F2 for Vowel Plot

```{r vowel-space, eval=FALSE}
# Analyze multiple vowel tokens
vowels <- c("vowel_a.wav", "vowel_e.wav", "vowel_i.wav",
            "vowel_o.wav", "vowel_u.wav")

vowel_data <- data.frame()

for (file in vowels) {
  sound <- Sound$new(file)
  formant <- sound$to_formant_burg()
  
  # Extract at vowel midpoint
  duration <- sound$get_total_duration()
  midpoint <- duration / 2
  
  f1 <- formant$get_value_at_time(1, midpoint, unit = "hertz")
  f2 <- formant$get_value_at_time(2, midpoint, unit = "hertz")
  
  vowel_data <- rbind(vowel_data, data.frame(
    vowel = tools::file_path_sans_ext(basename(file)),
    F1 = f1,
    F2 = f2
  ))
}

# Plot vowel space
ggplot(vowel_data, aes(x = F2, y = F1, label = vowel)) +
  geom_point(size = 3) +
  geom_text(vjust = -1) +
  scale_x_reverse() +  # F2 decreases left to right
  scale_y_reverse() +  # F1 decreases bottom to top
  labs(title = "Vowel Space",
       x = "F2 (Hz)", y = "F1 (Hz)") +
  theme_minimal()
```

### Formant Tracking Over Time

```{r formant-tracking, eval=FALSE}
sound <- Sound$new("diphthong.wav")
formant <- sound$to_formant_burg(time_step = 0.005)

# Extract F1 and F2 across time
times <- seq(0, sound$get_total_duration(), by = 0.005)
f1_values <- sapply(times, function(t) {
  formant$get_value_at_time(1, t, unit = "hertz")
})
f2_values <- sapply(times, function(t) {
  formant$get_value_at_time(2, t, unit = "hertz")
})

# Plot formant tracks
df <- data.frame(
  Time = rep(times, 2),
  Frequency = c(f1_values, f2_values),
  Formant = rep(c("F1", "F2"), each = length(times))
)

ggplot(df, aes(x = Time, y = Frequency, color = Formant)) +
  geom_line(size = 1) +
  labs(title = "Formant Tracks",
       x = "Time (s)", y = "Frequency (Hz)") +
  theme_minimal()
```

## Advanced: Formant from Excitation

For noisy or breathy speech, excitation-based formant extraction can be more robust:

```{r excitation-formant}
sound <- Sound$create_tone(duration = 0.3, sampling_rate = 22050)

# Method 1: Via Spectrum
spectrum <- sound$to_spectrum()
excitation <- spectrum$to_excitation()
formant_exc <- excitation$to_formant(max_formants = 5)

# Method 2: Via Cochleagram (more complex but robust)
cochlea <- sound$to_cochleagram()
excitation2 <- cochlea$to_excitation(0.15)  # At specific time
formant_exc2 <- excitation2$to_formant(max_formants = 5)
```

**When to use**:
- Breathy voice
- Noisy recordings
- Pathological voice
- Whispered speech

## Best Practices Summary

### General Recommendations

1. **Start with Burg** - Good default for most cases
2. **Verify with Willems or SL** - If results look suspicious
3. **Adjust ceiling** - Based on speaker demographics
4. **Use appropriate time step** - Balance speed vs. detail
5. **Check intensity** - Low intensity → unreliable formants

### Quality Checks

```{r quality-checks, eval=FALSE}
# 1. Check if formants are in expected ranges
f1 <- formant$get_value_at_time(1, time, unit = "hertz")
f2 <- formant$get_value_at_time(2, time, unit = "hertz")

if (!is.na(f1) && !is.na(f2)) {
  # F1 should be < F2
  if (f1 >= f2) warning("F1 >= F2, likely error")
  
  # Formants should be in typical ranges
  if (f1 < 200 || f1 > 1200) warning("F1 out of typical range")
  if (f2 < 700 || f2 > 3500) warning("F2 out of typical range")
}

# 2. Check bandwidth (shouldn't be too large)
b1 <- formant$get_bandwidth_at_time(1, time, unit = "hertz")
if (!is.na(b1) && b1 > 400) warning("Large F1 bandwidth")

# 3. Check formant continuity
f1_prev <- formant$get_value_at_time(1, time - 0.010, unit = "hertz")
if (!is.na(f1) && !is.na(f1_prev)) {
  jump <- abs(f1 - f1_prev)
  if (jump > 300) warning("Large formant jump (>300 Hz)")
}
```

## Session Info

```{r session}
sessionInfo()
```
