---
title: "Robust Formant Tracking with FormantPath"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust Formant Tracking with FormantPath}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

FormantPath is Praat's advanced formant tracking algorithm that tests multiple formant ceiling values and automatically selects the optimal tracking path. This approach is particularly valuable when:

- Speaker characteristics are unknown (male vs female vs child)
- Audio quality is poor or contains noise
- Standard formant tracking produces spurious values
- You need maximum robustness across diverse speakers

Unlike standard formant extraction (`to_formant_burg()`), FormantPath generates multiple candidate formant tracks with different ceiling frequencies, then uses statistical criteria to select the best path frame-by-frame.

```{r setup, message=FALSE}
library(pladdrr)
library(ggplot2)
```

## Why Use FormantPath?

### The Ceiling Frequency Problem

Formant tracking quality depends heavily on choosing the right **formant ceiling** - the maximum frequency for formant detection:

- **Too low** (e.g., 5000 Hz for female speaker): Misses F3, tracks harmonics as formants
- **Too high** (e.g., 6500 Hz for male speaker): Detects spurious formants, poor F1/F2 accuracy
- **Just right**: Accurate F1/F2/F3 tracking with minimal errors

**FormantPath solves this** by testing multiple ceilings automatically.

### Standard vs FormantPath Comparison

```{r standard-vs-path, fig.width=8, fig.height=4}
# Load example audio
sound <- Sound(system.file("extdata", "test.wav", package = "pladdrr"))

# Standard method (single ceiling)
formant_standard <- sound$to_formant_burg(
  time_step = 0.005,
  max_formants = 5,
  max_frequency = 5500
)

# FormantPath (multiple ceilings)
formant_path <- sound$to_formant_path(
  time_step = 0.005,
  max_num_formants = 5,
  formant_ceiling = 5500,
  num_steps_up_down = 2L  # Test 5 ceilings: ±10% around 5500 Hz
)

# Extract optimal formant from path
formant_optimal <- formant_path$extract_formant()

cat("Standard method: 1 ceiling (5500 Hz)\n")
cat("FormantPath: ", formant_path$get_number_of_candidates(), 
    " ceilings tested\n", sep = "")
cat("Ceiling range: ", 
    round(min(formant_path$get_all_ceiling_frequencies())), " - ",
    round(max(formant_path$get_all_ceiling_frequencies())), " Hz\n", sep = "")
```

## Basic Usage

### Creating a FormantPath

The key parameter is `num_steps_up_down` - how many ceiling values to test above and below the center:

```{r basic-creation}
# Test 3 ceilings (±5% around 5500 Hz)
fp <- sound$to_formant_path(
  time_step = 0.005,
  max_num_formants = 5,
  formant_ceiling = 5500,           # Center frequency
  ceiling_step_fraction = 0.05,     # ±5% steps
  num_steps_up_down = 1L            # 1 step up, 1 down = 3 total
)

cat("Created FormantPath with:\n")
cat("  Duration:", fp$get_duration(), "seconds\n")
cat("  Frames:", fp$get_nx(), "\n")
cat("  Candidates:", fp$get_number_of_candidates(), "\n")
cat("  Ceilings tested:", 
    paste(round(fp$get_all_ceiling_frequencies()), collapse = ", "), 
    "Hz\n")
```

### Interpreting num_steps_up_down

| `num_steps_up_down` | Ceilings Tested | Example (5500 Hz, 5% steps) |
|---------------------|-----------------|------------------------------|
| 0                   | 1               | 5500 Hz                      |
| 1                   | 3               | 5225, 5500, 5788 Hz         |
| 2                   | 5               | 4977, 5225, 5500, 5788, 6078 Hz |
| 3                   | 7               | 4737, 4977, ... 6378 Hz     |
| 4                   | 9               | 4507, ... 6696 Hz           |

More candidates = more robust but slower and more memory.

## Querying FormantPath

### Inspecting Candidates

```{r query-candidates}
fp <- sound$to_formant_path(num_steps_up_down = 2L)

# Get all tested ceilings
ceilings <- fp$get_all_ceiling_frequencies()
cat("Tested ceilings (Hz):\n")
print(round(ceilings))

# Check which candidate was chosen in middle frame
mid_frame <- fp$get_nx() %/% 2
chosen_candidate <- fp$get_candidate_in_frame(mid_frame)
chosen_ceiling <- fp$get_ceiling_frequency(chosen_candidate)

cat("\nAt frame", mid_frame, "(mid-point):\n")
cat("  Chosen candidate:", chosen_candidate, "\n")
cat("  Ceiling:", round(chosen_ceiling), "Hz\n")
```

### Stress Values (Quality Metric)

Each candidate has a "stress" value - lower is better:

```{r stress-values}
fp <- sound$to_formant_path(num_steps_up_down = 2L)

# Calculate stress for each candidate
n_candidates <- fp$get_number_of_candidates()
stresses <- sapply(1:n_candidates, function(i) {
  fp$get_stress_of_candidate(
    candidate = i,
    parameters = c(1, 1, 1, 1, 1),  # Equal weights
    powerf = 1.25
  )
})

# Find best candidate
best_idx <- which.min(stresses)
cat("Stress by candidate:\n")
print(data.frame(
  Candidate = 1:n_candidates,
  Ceiling = round(fp$get_all_ceiling_frequencies()),
  Stress = round(stresses, 4)
))
cat("\nLowest stress: Candidate", best_idx, 
    "(", round(fp$get_ceiling_frequency(best_idx)), "Hz )\n")
```

## Extracting the Optimal Formant

### Automatic Extraction

```{r extract-automatic}
fp <- sound$to_formant_path(num_steps_up_down = 2L)

# Extract the automatically selected optimal path
frm_result <- fp$extract_formant()

cat("Extracted formant:\n")
cat("  Duration:", sound$get_duration(), "seconds\n")
cat("  Frames:", frm_result$get_number_of_frames(), "\n")

# Query formant values
mid_time <- sound$get_duration() / 2
f1 <- frm_result$get_value_at_time(1, mid_time, "hertz")
f2 <- frm_result$get_value_at_time(2, mid_time, "hertz")
f3 <- frm_result$get_value_at_time(3, mid_time, "hertz")

cat("\nFormants at midpoint (", round(mid_time, 3), "s):\n", sep = "")
cat("  F1:", round(f1), "Hz\n")
cat("  F2:", round(f2), "Hz\n")
cat("  F3:", round(f3), "Hz\n")
```

### Manual Path Selection

You can manually override the automatic selection:

```{r manual-path, eval=FALSE}
# Set candidate 3 for time range 0.2-0.4s
fp$set_path(tmin = 0.2, tmax = 0.4, selected_candidate = 3)

# Or set optimal path for entire duration
fp$set_optimal_path(
  tmin = fp$get_xmin(),
  tmax = fp$get_xmax(),
  parameters = c(1, 1, 1, 1, 1),
  powerf = 1.25
)

# Then extract
frm_result <- fp$extract_formant()
```

## Visualizing FormantPath Results

### Comparing All Candidates

```{r plot-candidates, fig.width=8, fig.height=6}
fp <- sound$to_formant_path(num_steps_up_down = 2L)

# Export optimal formant track
df <- fp$as_data_frame()

# Plot F1 and F2
df_plot <- df[df$formant %in% 1:2, ]
df_plot$formant_num <- df_plot$formant

ggplot(df_plot, aes(time, frequency, color = factor(formant_num))) +
  geom_line(alpha = 0.7, linewidth = 1) +
  scale_color_manual(
    values = c("1" = "#E41A1C", "2" = "#377EB8"),
    labels = c("F1", "F2")
  ) +
  labs(
    title = "FormantPath: Optimal Formant Track",
    subtitle = paste("Selected from", fp$get_number_of_candidates(), "ceiling frequency candidates"),
    x = "Time (s)",
    y = "Frequency (Hz)",
    color = "Formant"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

### Optimal Path vs Standard

```{r plot-comparison, fig.width=8, fig.height=5}
# Get optimal formant from FormantPath
fp <- sound$to_formant_path(num_steps_up_down = 2L)
formant_optimal <- fp$extract_formant()
df_optimal <- as.data.frame(formant_optimal)
df_optimal$method <- "FormantPath (Optimal)"

# Standard single-ceiling with full parameters
formant_standard <- sound$to_formant_burg(
  time_step = 0.005,
  max_formants = 5,
  max_frequency = 5500,
  window_length = 0.025,
  pre_emphasis_from = 50
)
df_standard <- as.data.frame(formant_standard)
df_standard$method <- "Standard (5500 Hz)"

# Combine and plot
combined_df <- rbind(df_optimal, df_standard)
combined_df <- combined_df[combined_df$formant %in% 1:3, ]
ggplot(combined_df, aes(time, frequency, color = method)) +
  geom_line(alpha = 0.7, linewidth = 0.8) +
  facet_wrap(~ paste0("F", formant), ncol = 1, scales = "free_y") +
  scale_color_manual(values = c(
    "FormantPath (Optimal)" = "#E41A1C",
    "Standard (5500 Hz)" = "#377EB8"
  )) +
  labs(
    title = "FormantPath vs Standard Formant Tracking",
    x = "Time (s)",
    y = "Frequency (Hz)",
    color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

## Advanced Path Finding

### Viterbi-Style Path Optimization

FormantPath can use a Viterbi-like algorithm to find the globally optimal path:

```{r path-finder, eval=FALSE}
fp <- sound$to_formant_path(num_steps_up_down = 3L)

# Find optimal path with custom weights
fp$path_finder(
  q_weight = 1.0,                        # Formant quality
  frequency_change_weight = 1.0,         # Smoothness penalty
  stress_weight = 1.0,                   # Model fit
  ceiling_change_weight = 1.0,           # Consistency bonus
  intensity_modulation_step_size = 5.0,
  window_length = 0.025,
  parameters = c(1, 1, 1, 1, 1),
  powerf = 1.25
)

frm_result <- fp$extract_formant()
```

**Weight meanings**:

- `q_weight`: Prefer higher formant quality scores
- `frequency_change_weight`: Penalize large formant jumps
- `stress_weight`: Prefer lower stress (better model fit)
- `ceiling_change_weight`: Prefer consistent ceiling choice

## Use Cases

### 1. Unknown Speaker Demographics

When you don't know if the speaker is male, female, or child:

```{r use-case-unknown}
# Test wide range of ceilings
fp <- sound$to_formant_path(
  formant_ceiling = 5500,      # Middle ground
  ceiling_step_fraction = 0.10, # ±10% steps (wider range)
  num_steps_up_down = 3L       # 7 candidates: 4500-6800 Hz
)

ceilings <- fp$get_all_ceiling_frequencies()
cat("Testing range:", round(min(ceilings)), "-", 
    round(max(ceilings)), "Hz\n")
cat("Covers: Male (5000), Female (5500), Child (6500+)\n")

frm_result <- fp$extract_formant()
```

### 2. Noisy or Difficult Audio

For poor quality audio, use more candidates and tighter steps:

```{r use-case-noisy}
fp <- sound$to_formant_path(
  formant_ceiling = 5500,
  ceiling_step_fraction = 0.05,  # Finer steps
  num_steps_up_down = 4L,        # More candidates (9 total)
  window_length = 0.030          # Longer window for stability
)

cat("Using", fp$get_number_of_candidates(), "candidates for robustness\n")
```

### 3. Vowel-Specific Analysis

Compare formants across different vowels:

```{r use-case-vowels, eval=FALSE}
# Assuming you have time-aligned TextGrid
textgrid <- TextGrid(system.file("extdata", "test.TextGrid", package = "pladdrr"))

# Extract vowel segments
intervals <- textgrid$get_tier(1)$get_intervals()
vowels <- intervals[intervals$text %in% c("i", "a", "u")]

# Analyze each vowel with FormantPath
results <- lapply(vowels, function(v) {
  # Extract vowel segment
  vowel_sound <- sound$extract_part(v$xmin, v$xmax)
  
  # Robust formant tracking
  fp <- vowel_sound$to_formant_path(num_steps_up_down = 2L)
  frm_result <- fp$extract_formant()
  
  # Get mean F1/F2
  df <- as.data.frame(frm_result)
  list(
    vowel = v$text,
    f1 = mean(df[df$formant == 1, "frequency"], na.rm = TRUE),
    f2 = mean(df[df$formant == 2, "frequency"], na.rm = TRUE)
  )
})

vowel_space <- do.call(rbind, lapply(results, as.data.frame))
print(vowel_space)
```

### 4. Batch Processing Multiple Files

```{r batch-processing, eval=FALSE}
# Process multiple speakers with unknown characteristics
files <- c("speaker1.wav", "speaker2.wav", "speaker3.wav")

formants <- lapply(files, function(file) {
  sound <- Sound(file)
  fp <- sound$to_formant_path(
    formant_ceiling = 5500,
    num_steps_up_down = 2L
  )
  fp$extract_formant()
})

names(formants) <- files
```

## Performance Considerations

### Memory Usage

FormantPath stores multiple formant candidates:

```{r memory-note, eval=FALSE}
# 5 candidates = 5× the memory of standard Formant
fp <- sound$to_formant_path(num_steps_up_down = 2L)  # 5 candidates
# Memory ≈ 5 × size of single Formant object
```

**Tips**:
- Use fewer candidates for long files (`num_steps_up_down = 1`)
- Extract and discard FormantPath after getting optimal Formant
- Process files individually rather than loading all in memory

### Computation Time

FormantPath is slower than standard extraction:

```{r timing, eval=FALSE}
# Standard: ~0.3s for 1s audio
system.time(frm_result <- sound$to_formant_burg())

# FormantPath with 5 candidates: ~1.5s for 1s audio
system.time(fp <- sound$to_formant_path(num_steps_up_down = 2L))
```

Roughly **N× slower** where N = number of candidates.

## Best Practices

### ✓ DO

1. **Use FormantPath when speaker characteristics are unknown**
2. **Start with `num_steps_up_down = 2`** (5 candidates) as default
3. **Extract the formant immediately** to save memory:
   ```r
   frm_result <- sound$to_formant_path(num_steps_up_down = 2L)$extract_formant()
   ```
4. **Compare candidates visually** when results look suspicious
5. **Use wider step fractions (0.10)** for diverse speakers
6. **Document which method you used** for reproducibility

### ✗ DON'T

1. **Don't use FormantPath for all analyses** - standard method is faster for known speakers
2. **Don't use too many candidates** (`num_steps_up_down > 4`) unless necessary
3. **Don't forget to extract** - FormantPath itself isn't directly usable
4. **Don't assume optimal = perfect** - always inspect difficult cases
5. **Don't ignore stress values** - high stress indicates poor fit

## Comparison with Other Tools

### pladdrr FormantPath vs Praat

pladdrr provides direct R access to Praat's FormantPath algorithm:

| Feature | pladdrr | Praat GUI |
|---------|---------|-----------|
| Multiple candidates | ✅ | ✅ |
| Automatic path selection | ✅ | ✅ |
| Manual path override | ✅ | ✅ |
| Data export | ✅ (data.frame) | ✅ (text file) |
| Batch processing | ✅ (R loops) | ❌ (manual) |
| Visualization | ✅ (ggplot2) | ✅ (built-in) |

### vs Parselmouth

Parselmouth does not currently expose FormantPath - only standard `to_formant()`.

## Troubleshooting

### Issue: All candidates look similar

**Cause**: Step fraction too small or range too narrow

**Solution**: Increase `ceiling_step_fraction` to 0.10 or higher

```{r troubleshoot-similar, eval=FALSE}
# Instead of this (too narrow)
fp <- sound$to_formant_path(
  formant_ceiling = 5500,
  ceiling_step_fraction = 0.02,  # ±2% only
  num_steps_up_down = 2L
)

# Use this (better range)
fp <- sound$to_formant_path(
  formant_ceiling = 5500,
  ceiling_step_fraction = 0.10,  # ±10%
  num_steps_up_down = 2L
)
```

### Issue: Extracted formant has gaps (NA values)

**Cause**: No candidate had valid formants at those frames

**Solution**: Increase `max_num_formants` or adjust `formant_ceiling`

```{r troubleshoot-gaps, eval=FALSE}
fp <- sound$to_formant_path(
  max_num_formants = 6,  # Instead of 5
  formant_ceiling = 5500,
  num_steps_up_down = 2L
)
```

### Issue: FormantPath is too slow

**Solution**: Reduce candidates or increase time step

```{r troubleshoot-slow, eval=FALSE}
# Faster: fewer candidates, larger time step
fp <- sound$to_formant_path(
  time_step = 0.010,         # Instead of 0.005
  num_steps_up_down = 1L,    # 3 candidates instead of 5
  formant_ceiling = 5500
)
```

## Summary

**FormantPath is ideal when**:
- Speaker demographics unknown
- Audio quality is poor
- Need maximum robustness
- Comparing diverse speakers

**Use standard formant extraction when**:
- Speaker type is known (male/female)
- Audio quality is good
- Speed is critical
- Memory is limited

**Key Parameters**:
- `formant_ceiling`: Center frequency (5000-5500 for adults)
- `num_steps_up_down`: 1-2 for most cases, 3-4 for difficult audio
- `ceiling_step_fraction`: 0.05 (narrow) to 0.10 (wide)

## Further Reading

- Praat manual: [FormantPath](http://www.fon.hum.uva.nl/praat/manual/FormantPath.html)
- Weenink, D. (2015). "Improved formant frequency measurements of short segments"
- pladdrr documentation: `?to_formant_path`

## Session Info

```{r session-info}
sessionInfo()
```
