---
title: "Getting Started with speaker: Phonetic Analysis in R"
author: "speaker Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with speaker: Phonetic Analysis in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The `speaker` package provides direct access to Praat's phonetic analysis capabilities from R, without requiring Python. It implements core phonetic analysis objects (Sound, Pitch, Formant, Intensity) with a modern R6 object-oriented interface that follows Praat's conventions.

## Installation

```{r eval = FALSE}
# Install from source
install.packages("speaker", type = "source")

# Or install from GitHub
# devtools::install_github("your-username/speaker")
```

```{r}
library(pladdrr)
```

## Creating and Loading Sounds

### Generate Test Sounds

The package includes functions to generate test signals:

```{r}
# Generate a 440 Hz sine wave (A4 note)
sound_a4 <- generate_sine_wave(frequency = 440, duration = 0.5, 
                               amplitude = 0.7, sampling_rate = 44100)

print(sound_a4)
```

```{r}
# Generate white noise
noise <- generate_noise(duration = 0.2, sampling_rate = 44100, 
                       amplitude = 0.3)

print(noise)
```

### Load Audio Files

```{r eval = FALSE}
# Load a WAV file using R6 interface
speech <- Sound$new("path/to/speech.wav")

# Load specific channel from stereo file (1-indexed)
speech_left <- Sound$new("path/to/stereo.wav", channel = 1)
speech_right <- Sound$new("path/to/stereo.wav", channel = 2)
```

## Basic Sound Properties

Extract basic information from sound objects using R6 methods:

```{r}
# Duration in seconds
cat("Duration:", sound_a4$get_duration(), "seconds\n")

# Sampling rate
cat("Sampling rate:", sound_a4$get_sampling_frequency(), "Hz\n")

# Number of samples
cat("Samples:", sound_a4$get_number_of_samples(), "\n")

# Number of channels
cat("Channels:", sound_a4$get_number_of_channels(), "\n")
```

## Sound Statistics

Calculate amplitude statistics:

```{r}
# Individual statistics
cat("Mean amplitude:", sound_mean(sound_a4), "\n")
cat("RMS amplitude:", sound_rms(sound_a4), "\n")
cat("Min amplitude:", sound_min(sound_a4), "\n")
cat("Max amplitude:", sound_max(sound_a4), "\n")

# All statistics at once
stats <- sound_statistics(sound_a4)
print(stats)
```

## Pitch Analysis

Extract fundamental frequency (F0) from speech or sustained tones using R6 methods:

```{r}
# For speech, use typical settings
# pitch <- speech$to_pitch(pitch_floor = 75, pitch_ceiling = 600)

# For our test signal, we need to create a more complex sound
# (pure sine waves may not be detected as voiced)
# In real usage with speech:
# mean_f0 <- pitch$get_mean()
# min_f0 <- pitch$get_minimum()
# max_f0 <- pitch$get_maximum()
```

### Pitch Parameters

```{r eval = FALSE}
# Male voice
pitch_male <- sound$to_pitch(pitch_floor = 50, pitch_ceiling = 300)

# Female voice  
pitch_female <- sound$to_pitch(pitch_floor = 100, pitch_ceiling = 600)

# Child voice
pitch_child <- sound$to_pitch(pitch_floor = 150, pitch_ceiling = 800)
```

### Querying Pitch Values

```{r eval = FALSE}
# Get F0 at specific time point
f0_at_1s <- pitch$get_value_at_time(time = 1.0)

# Get mean F0 over time range
mean_f0_range <- pitch$get_mean(from_time = 0.5, to_time = 1.5)

# Get minimum and maximum
min_f0 <- pitch$get_minimum()
max_f0 <- pitch$get_maximum()
```

## Formant Analysis

Analyze vocal tract resonances (formants) for vowel characterization using R6:

```{r}
# Extract formants (default settings for adult female)
formants <- sound_a4$to_formant_burg(max_frequency = 5500, max_formants = 5)

# R6 objects print nicely
formants
```

### Speaker-Specific Settings

```{r eval = FALSE}
# Adult male
formants_male <- sound$to_formant_burg(max_frequency = 5000, max_formants = 5)

# Adult female
formants_female <- sound$to_formant_burg(max_frequency = 5500, max_formants = 5)

# Child
formants_child <- sound$to_formant_burg(max_frequency = 8000, max_formants = 5)
```

### Querying Formant Values

```{r}
# Get F1 and F2 at specific time (vowel quality)
f1 <- get_formant_at_time(formants, formant_number = 1, time = 0.25)
f2 <- get_formant_at_time(formants, formant_number = 2, time = 0.25)

cat("F1:", round(f1, 1), "Hz\n")
cat("F2:", round(f2, 1), "Hz\n")
```

```{r}
# Get mean formant over time range
mean_f1 <- get_mean_formant(formants, formant_number = 1, 
                           time_range = c(0.1, 0.4))
mean_f2 <- get_mean_formant(formants, formant_number = 2, 
                           time_range = c(0.1, 0.4))

cat("Mean F1:", round(mean_f1, 1), "Hz\n")
cat("Mean F2:", round(mean_f2, 1), "Hz\n")
```

## Intensity Analysis

Measure sound power (loudness) over time using R6:

```{r}
# Extract intensity
intensity <- sound_a4$to_intensity(minimum_pitch = 100)

# R6 objects print nicely
intensity
```

### Intensity Statistics

```{r}
# Mean intensity
mean_db <- intensity$get_mean()
cat("Mean intensity:", round(mean_db, 2), "dB\n")

# Intensity range
min_db <- intensity$get_minimum()
max_db <- intensity$get_maximum()
cat("Intensity range:", round(min_db, 2), "-", round(max_db, 2), "dB\n")

# Standard deviation
sd_db <- intensity$get_standard_deviation()
cat("SD intensity:", round(sd_db, 2), "dB\n")
```

### Intensity at Specific Time

```{r}
# Query intensity at time point
int_at_time <- intensity$get_value_at_time(time = 0.25)
cat("Intensity at 0.25s:", round(int_at_time, 2), "dB\n")

# With interpolation (cubic)
int_interp <- intensity$get_value_at_time(time = 0.25, interpolation = "cubic")
cat("Intensity (interpolated):", round(int_interp, 2), "dB\n")
```

## Working with Data Frames

All analysis objects can be converted to data frames for plotting and further analysis:

```{r}
# Convert to data frame using R6 method
formant_df <- formants$as_data_frame()
head(formant_df)
```

```{r}
intensity_df <- intensity$as_data_frame()
head(intensity_df)
```

## Visualization

The `speaker` package provides comprehensive ggplot2-based visualization functions for all analysis types. For detailed examples of creating publication-quality plots, see `vignette("visualization")`, which covers:

- Voice quality visualization (AVQI, DSI, CPP)
- Formant analysis (vowel spaces, trajectories)
- Pitch and intensity contours
- Spectral analysis (spectrograms, spectra, LTAS)
- TextGrid annotations
- Multi-panel diagnostic reports

## Complete Workflow Example

Here's a complete analysis workflow using R6:

```{r eval = FALSE}
# 1. Load sound
sound <- Sound$new("vowel.wav")

# 2. Extract all analyses
pitch <- sound$to_pitch(pitch_floor = 75, pitch_ceiling = 600)
formants <- sound$to_formant_burg(max_frequency = 5500, max_formants = 5)
intensity <- sound$to_intensity(minimum_pitch = 75)

# 3. Get measurements at vowel midpoint
midpoint <- sound$get_duration() / 2

f0 <- pitch$get_value_at_time(time = midpoint)
f1 <- formants$get_value_at_time(formant_number = 1, time = midpoint)
f2 <- formants$get_value_at_time(formant_number = 2, time = midpoint)
f3 <- formants$get_value_at_time(formant_number = 3, time = midpoint)
int <- intensity$get_value_at_time(time = midpoint)

# 4. Create summary
vowel_data <- data.frame(
  F0 = f0,
  F1 = f1,
  F2 = f2,
  F3 = f3,
  Intensity = int
)

print(vowel_data)
```

## Comparing with Praat Scripts

The package follows Praat's conventions, making it easy to translate Praat scripts:

### Praat Script
```
# Praat
sound = Read from file: "sound.wav"
pitch = To Pitch: 0.01, 75, 600
f0 = Get mean: 0, 0, "Hertz"
```

### speaker Equivalent (R6)
```{r eval = FALSE}
# R with speaker (R6 interface)
sound <- Sound$new("sound.wav")
pitch <- sound$to_pitch(time_step = 0.01, 
                       pitch_floor = 75, pitch_ceiling = 600)
f0 <- pitch$get_mean(from_time = 0, to_time = 0, unit = "hertz")
```

## Advanced Topics

### Custom Analysis Parameters

```{r eval = FALSE}
# Fine-tune formant detection
formants <- sound$to_formant_burg(
  time_step = 0.005,           # 5 ms steps
  max_frequency = 5500,        # Adult female range
  max_formants = 5,            # Track F1-F5
  window_length = 0.025,       # 25 ms window
  pre_emphasis_from = 50       # Pre-emphasis from 50 Hz
)

# Fine-tune intensity
intensity <- sound$to_intensity(
  minimum_pitch = 75,          # Affects window length
  time_step = 0.01,            # 10 ms steps (auto if 0)
  subtract_mean = FALSE        # Absolute intensity in dB SPL
)
```

### Handling NA Values

All analysis functions gracefully handle undefined values:

```{r eval = FALSE}
# Pitch may be NA for unvoiced segments
f0 <- pitch$get_value_at_time(time = 1.0)
if (is.na(f0)) {
  cat("Unvoiced at this time point\n")
} else {
  cat("F0:", f0, "Hz\n")
}

# Mean functions automatically exclude NA
mean_f0 <- get_mean_pitch(pitch)  # NA values excluded
```

## Package Information

```{r}
# Package version
packageVersion("speaker")

# Citation information
citation("speaker")
```

## See Also

- Praat: <https://www.praat.org>
- Package documentation: `help(package = "pladdrr")`
- Function reference: `?extract_pitch`, `?extract_formants`, `?extract_intensity`

## Conclusion

The `speaker` package provides a comprehensive toolkit for phonetic analysis in R, with:

- **4 core analysis objects**: Sound, Pitch, Formant, Intensity
- **45+ functions**: Complete analysis capabilities
- **Praat compatibility**: Easy translation from Praat scripts
- **Clean S3 interface**: Familiar R programming patterns
- **Robust error handling**: Graceful handling of edge cases

For more information, see the individual function documentation and the package README.
