---
title: "Integrated Phonetic Analysis with pladdrr"
author: "pladdrr Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrated Phonetic Analysis with pladdrr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE  # Disable evaluation due to complex dependencies
)
```

## Overview

This vignette demonstrates how to perform integrated phonetic analysis workflows using the **pladdrr** package. We will cover:

- **TextGrid-based segmentation**: Create and manipulate annotations
- **Sound manipulation**: Extract, resample, and normalize audio
- **Acoustic measurements**: F0, formants, intensity, and voice quality
- **Batch processing**: Analyze multiple segments efficiently
- **Data export**: Prepare data for statistical analysis in R

This workflow is typical in phonetic research, where annotated speech data requires systematic acoustic analysis.

## Background

### TextGrid Annotation

TextGrids are Praat's standard format for time-aligned annotations. They consist of:

- **Interval tiers**: Segments with start/end times and labels (e.g., words, phones)
- **Point tiers**: Time points with labels (e.g., landmarks, events)

TextGrids enable:

- Manual or automatic (forced alignment) annotation
- Multi-level linguistic annotation (phones, syllables, words, utterances)
- Integration with acoustic analysis

### The Phonetic Analysis Pipeline

A typical phonetic analysis workflow:

1. **Record** and preprocess audio
2. **Annotate** with TextGrid (manual or via forced alignment)
3. **Segment** audio based on annotations
4. **Extract** acoustic features for each segment
5. **Analyze** patterns statistically

The **pladdrr** package provides all necessary tools for steps 2-5.

## Part 1: Loading and Creating Data

### Creating Synthetic Audio

For demonstration, we create a 5-second audio signal. In real research, you would load existing recordings.

```{r create_audio}
library(pladdrr)

# Create synthetic audio (440 Hz tone)
sound <- Sound$create_tone(
  duration = 5.0,
  frequency = 440,
  sampling_rate = 16000,
  amplitude = 0.5
)

# Inspect audio properties
cat("Duration:", sound$get_duration(), "seconds\n")
cat("Sample rate:", sound$get_sampling_frequency(), "Hz\n")
cat("Channels:", sound$get_number_of_channels(), "\n")
```

### Creating a TextGrid

Create a TextGrid with word and phone annotations:

```{r create_textgrid}
# Create TextGrid with 2 interval tiers
tg <- TextGrid$create(
  tmin = 0,
  tmax = 5,
  tier_names = "words phones",
  point_tiers = ""  # Both are interval tiers
)

# Add word boundaries
tg$insert_boundary(1, 1.0)
tg$insert_boundary(1, 2.5)
tg$insert_boundary(1, 4.0)

# Label word intervals
tg$set_interval_text(1, 1, "silence")
tg$set_interval_text(1, 2, "hello")
tg$set_interval_text(1, 3, "world")
tg$set_interval_text(1, 4, "silence")

# Add phone boundaries (finer granularity)
phone_times <- c(1.0, 1.3, 1.7, 2.0, 2.5, 3.0, 3.5, 4.0)
for (t in phone_times) {
  tg$insert_boundary(2, t)
}

# Label phone intervals
phone_labels <- c("", "h", "E", "l", "oU", "w", "3", "ld", "")
for (i in seq_along(phone_labels)) {
  tg$set_interval_text(2, i, phone_labels[i])
}

cat("TextGrid created with", tg$get_number_of_tiers(), "tiers\n")
```

## Part 2: TextGrid Queries

### Exploring Tier Structure

```{r explore_tiers}
# Get tier information
tier_names <- tg$get_tier_names()
for (i in 1:tg$get_number_of_tiers()) {
  tier_type <- if (tg$tier_is_interval_tier(i)) "IntervalTier" else "PointTier"
  n_items <- tg$get_number_of_intervals(i)
  cat(sprintf("Tier %d: %s (%s) - %d items\n", 
              i, tier_names[i], tier_type, n_items))
}
```

### Extracting Interval Information

```{r query_intervals}
# Get all intervals from the phones tier
phone_tier <- 2
n_intervals <- tg$get_number_of_intervals(phone_tier)

# Create a data frame with interval information
phone_data <- data.frame(
  interval = 1:n_intervals,
  start = numeric(n_intervals),
  end = numeric(n_intervals),
  duration = numeric(n_intervals),
  label = character(n_intervals),
  stringsAsFactors = FALSE
)

for (i in 1:n_intervals) {
  phone_data$start[i] <- tg$get_interval_start_time(phone_tier, i)
  phone_data$end[i] <- tg$get_interval_end_time(phone_tier, i)
  phone_data$duration[i] <- phone_data$end[i] - phone_data$start[i]
  phone_data$label[i] <- tg$get_interval_text(phone_tier, i)
}

print(phone_data)
```

## Part 3: Sound Manipulation

### Extracting Audio Segments

Extract specific portions of audio based on TextGrid intervals:

```{r extract_segments, eval=FALSE}
# Extract the "hello" word (interval 2 of word tier)
hello_start <- tg$get_interval_start_time(1, 2)
hello_end <- tg$get_interval_end_time(1, 2)

hello_sound <- sound$extract_part(
  from_time = hello_start,
  to_time = hello_end,
  preserve_times = FALSE
)

cat("Extracted 'hello':", hello_sound$get_total_duration(), "seconds\n")
```

### Resampling

Change sampling rate for analysis or compatibility:

```{r resample}
# Resample to 8 kHz (e.g., for telephony simulations)
sound_8k <- sound$resample(new_frequency = 8000, precision = 50)
cat("Resampled to", sound_8k$get_sampling_frequency(), "Hz\n")
```

### Intensity Normalization

Normalize loudness across recordings:

```{r normalize}
# Scale to 70 dB SPL
sound_normalized <- sound$scale_intensity(new_intensity = 70)
cat("Normalized to 70 dB\n")
```

### Pre-emphasis Filtering

Apply pre-emphasis for formant analysis:

```{r preemphasis}
# Pre-emphasize from 50 Hz
sound_preemph <- sound$pre_emphasize(from_frequency = 50)
cat("Applied pre-emphasis from 50 Hz\n")
```

## Part 4: Acoustic Analysis

### Fundamental Frequency (F0) Extraction

```{r f0_extraction}
# Extract pitch contour
pitch <- sound$to_pitch(
  time_step = 0.01,
  pitch_floor = 75,
  pitch_ceiling = 600
)

# Get statistics
f0_mean <- pitch$get_mean(from_time = 0, to_time = 0, unit = "hertz")
f0_sd <- pitch$get_standard_deviation(from_time = 0, to_time = 0, unit = "hertz")

cat(sprintf("F0: Mean = %.1f Hz, SD = %.1f Hz\n", f0_mean, f0_sd))
```

### Formant Analysis

```{r formant_analysis}
# Extract formants using Burg's algorithm
formant <- sound$to_formant_burg(
  time_step = 0.01,
  max_formants = 5,
  max_frequency = 5500,
  window_length = 0.025,
  pre_emphasis_from = 50
)

# Measure formants at specific time point (e.g., vowel midpoint)
vowel_time <- 1.5  # Middle of "E" vowel
f1 <- formant$get_value_at_time(formant_number = 1, time = vowel_time, unit = "hertz")
f2 <- formant$get_value_at_time(formant_number = 2, time = vowel_time, unit = "hertz")
f3 <- formant$get_value_at_time(formant_number = 3, time = vowel_time, unit = "hertz")

cat(sprintf("Formants at %.2f s: F1 = %.0f Hz, F2 = %.0f Hz, F3 = %.0f Hz\n",
            vowel_time, f1, f2, f3))
```

### Intensity

```{r intensity}
# Extract intensity contour
intensity <- sound$to_intensity(
  minimum_pitch = 100,
  time_step = 0.01
)

int_mean <- intensity$get_mean(from_time = 0, to_time = 0)
cat(sprintf("Mean intensity: %.1f dB\n", int_mean))
```

### Voice Quality (HNR)

```{r hnr}
# Harmonics-to-Noise Ratio
harmonicity <- sound$to_harmonicity_cc(
  time_step = 0.01,
  min_pitch = 75,
  silence_threshold = 0.1,
  periods_per_window = 1.0
)

hnr_mean <- harmonicity$get_mean(from_time = 0, to_time = 0)
cat(sprintf("Mean HNR: %.1f dB\n", hnr_mean))
```

## Part 5: Batch Processing

Process all vowel segments systematically:

```{r batch_processing}
# Identify vowel intervals (uppercase labels)
vowel_intervals <- which(grepl("[AEIOUY3]", phone_data$label, ignore.case = FALSE))

# Extract acoustic features for each vowel
vowel_features <- data.frame(
  phone = character(),
  start = numeric(),
  end = numeric(),
  duration = numeric(),
  f0_mean = numeric(),
  f1 = numeric(),
  f2 = numeric(),
  f3 = numeric(),
  intensity = numeric(),
  hnr = numeric(),
  stringsAsFactors = FALSE
)

for (idx in vowel_intervals) {
  label <- phone_data$label[idx]
  start <- phone_data$start[idx]
  end <- phone_data$end[idx]
  midpoint <- start + (end - start) / 2
  
  # Extract features at vowel midpoint
  f0 <- pitch$get_value_at_time(time = midpoint, unit = "hertz", interpolate = "linear")
  f1 <- formant$get_value_at_time(1, midpoint, "hertz")
  f2 <- formant$get_value_at_time(2, midpoint, "hertz")
  f3 <- formant$get_value_at_time(3, midpoint, "hertz")
  int <- intensity$get_value_at_time(midpoint, interpolate = "cubic")
  hnr <- harmonicity$get_value_at_time(midpoint, interpolate = "linear")
  
  vowel_features <- rbind(vowel_features, data.frame(
    phone = label,
    start = start,
    end = end,
    duration = end - start,
    f0_mean = if (is.na(f0)) NA else f0,
    f1 = f1,
    f2 = f2,
    f3 = f3,
    intensity = int,
    hnr = hnr
  ))
}

print(vowel_features)
```

## Part 6: Data Export and Visualization

### Export to CSV

```{r export_csv, eval=FALSE}
# Save for statistical analysis
write.csv(vowel_features, "vowel_acoustic_features.csv", row.names = FALSE)
```

### Visualization

The `pladdrr` package provides comprehensive plotting functions for all analysis types. For detailed visualization examples including vowel spaces, formant trajectories, pitch/intensity contours, spectrograms, and voice quality reports, see `vignette("visualization")`.

## Real-World Applications

### 1. Sociolinguistic Research

Analyze vowel variation across speakers, dialects, or social groups:

```r
# Load multiple speakers
speakers <- c("speaker1.wav", "speaker2.wav", "speaker3.wav")
textgrids <- c("speaker1.TextGrid", "speaker2.TextGrid", "speaker3.TextGrid")

# Process all speakers
all_data <- data.frame()
for (i in seq_along(speakers)) {
  sound <- Sound$new(speakers[i])
  tg <- TextGrid$new(textgrids[i])
  # ... extract features ...
  all_data <- rbind(all_data, features)
}

# Statistical analysis
model <- lm(f1 ~ vowel * speaker, data = all_data)
```

### 2. L2 Acquisition Studies

Track vowel development in second language learners:

```r
# Longitudinal data (same speaker, multiple sessions)
sessions <- c("session1.wav", "session2.wav", "session3.wav")

# Measure vowel space area over time
# ... (see Example 9 for vowel space calculations) ...
```

### 3. Clinical Voice Assessment

Multi-dimensional voice profiling:

```r
# Extract comprehensive voice quality metrics
voice_profile <- data.frame(
  f0_mean = pitch$get_mean(),
  f0_sd = pitch$get_standard_deviation(),
  jitter = pointprocess$get_jitter_local(),
  shimmer = pointprocess_sound$get_shimmer_local(),
  hnr = harmonicity$get_mean(),
  intensity_mean = intensity$get_mean()
)

# Add cepstral peak prominence smoothed (CPPS)
# CPPS is highly sensitive to parameter choice - see example for details
cepstrogram <- intensity$to_power_cepstrogram(
  pitch_floor = 60,
  time_step = 0.002
)
voice_profile$cpps <- cepstrogram$get_cpps(
  subtract_tilt = FALSE,  # Match Praat default
  time_averaging_window = 0.001,
  quefrency_averaging_window = 0.0005,
  pitch_floor = 60,
  pitch_ceiling = 330
)
```

**Note:** CPPS values are highly sensitive to parameters. See `system.file("examples", "10_cpps_analysis.R", package = "pladdrr")` for a comprehensive guide on parameter effects and clinical interpretation.

## Best Practices

### 1. Pre-emphasis for Formants

Always pre-emphasize before formant analysis:

```r
sound_preemph <- sound$pre_emphasize(from_frequency = 50)
formant <- sound_preemph$to_formant_burg(...)
```

### 2. Appropriate Analysis Parameters

Choose parameters based on speaker characteristics:

```r
# Adult male
pitch <- sound$to_pitch(pitch_floor = 75, pitch_ceiling = 300)
formant <- sound$to_formant_burg(max_frequency = 5000)

# Adult female
pitch <- sound$to_pitch(pitch_floor = 100, pitch_ceiling = 500)
formant <- sound$to_formant_burg(max_frequency = 5500)

# Child
pitch <- sound$to_pitch(pitch_floor = 200, pitch_ceiling = 700)
formant <- sound$to_formant_burg(max_frequency = 8000)
```

### 3. Handle Missing Values

Acoustic measurements may fail:

```r
f0 <- pitch$get_value_at_time(time, "hertz", interpolate = "linear")
if (is.na(f0)) {
  # Handle voiceless segments or tracking failures
  cat("Warning: No F0 at time", time, "\n")
}
```

### 4. Efficient Batch Processing

For large corpora, use vectorization and sampling:

```r
# Sample large datasets
if (nrow(data) > 10000) {
  data <- data[sample(nrow(data), 1000), ]  # Random sample
}
```

## Summary

The **pladdrr** package enables:

- **Flexible annotation workflows** with TextGrid creation and manipulation
- **Comprehensive sound processing** (segmentation, resampling, filtering)
- **Multi-dimensional acoustic analysis** (prosody, voice quality, spectral features)
- **Seamless R integration** for statistical analysis and visualization

For more examples, see:

- `vignette("vowel-space-analysis")` - F1-F2 analysis and normalization
- `vignette("textgrid-workflows")` - Advanced TextGrid manipulation
- `inst/examples/` - Complete workflow scripts

## References

- Boersma, P., & Weenink, D. (2023). *Praat: doing phonetics by computer*. http://www.praat.org/
- Thomas, E. R., & Kendall, T. (2007). NORM: The vowel normalization and plotting suite. http://ncslaap.lib.ncsu.edu/tools/norm/
