---
title: "Migration Guide: From Parselmouth (Python) to pladdrr"
author: "Fredrik Nylén"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migration Guide: From Parselmouth (Python) to pladdrr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This guide helps Python users familiar with Parselmouth transition to pladdrr for R. Both packages provide access to Praat's functionality, but pladdrr offers a more direct, object-oriented interface.

## Key Differences

### Architecture

**Parselmouth (Python):**
- Generic `praat.call()` dispatcher
- String-based command names
- Python wrapper around Praat C++ code

**pladdrr (R):**
- Direct R6 method calls
- Type-safe parameters
- Direct C++ binding (Rcpp)

### Syntax Comparison

| Feature | Parselmouth | pladdrr |
|---------|-------------|---------|
| Object creation | `pm.Sound(file)` | `Sound$new(file)` |
| Method calls | `pm.praat.call(obj, "Command", args)` | `obj$command(args)` |
| Parameter style | Positional | Named |
| Autocomplete | Limited | Full IDE support |

## Common Operations

### Loading Sound Files

**Parselmouth:**
```python
import parselmouth as pm

sound = pm.Sound("audio.wav")
```

**pladdrr:**
```{r}
library(pladdrr)

sound <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))
```

### Pitch Extraction

**Parselmouth:**
```python
import parselmouth as pm

sound = pm.Sound("audio.wav")
pitch = pm.praat.call(sound, "To Pitch", 0.01, 75, 600)
mean_f0 = pm.praat.call(pitch, "Get mean", 0, 0, "Hertz")
```

**pladdrr:**
```{r}
sound <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))
pitch <- sound$to_pitch(time_step = 0.01, pitch_floor = 75, pitch_ceiling = 600)
mean_f0 <- pitch$get_mean(from_time = 0, to_time = 0, unit = "hertz")
```

### Formant Analysis

**Parselmouth:**
```python
sound = pm.Sound("vowel.wav")
formant = pm.praat.call(sound, "To Formant (burg)", 0.01, 5, 5500, 0.025, 50)
f1 = pm.praat.call(formant, "Get value at time", 1, 0.5, "Hertz", "Linear")
f2 = pm.praat.call(formant, "Get value at time", 2, 0.5, "Hertz", "Linear")
```

**pladdrr:**
```{r}
sound <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))
formant <- sound$to_formant_burg(
  time_step = 0.01,
  max_number_of_formants = 5,
  maximum_formant = 5500,
  window_length = 0.025,
  pre_emphasis_from = 50
)
f1 <- formant$get_value_at_time(formant_number = 1, time = 0.5, unit = "hertz")
f2 <- formant$get_value_at_time(formant_number = 2, time = 0.5, unit = "hertz")
```

### Intensity Measurements

**Parselmouth:**
```python
sound = pm.Sound("audio.wav")
intensity = pm.praat.call(sound, "To Intensity", 100, 0.01, True)
mean_int = pm.praat.call(intensity, "Get mean", 0, 0, "energy")
```

**pladdrr:**
```{r}
sound <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))
intensity <- sound$to_intensity(minimum_pitch = 100, time_step = 0.01, subtract_mean = TRUE)
mean_int <- intensity$get_mean(from_time = 0, to_time = 0, averaging_method = "energy")
```

### Spectral Analysis

**Parselmouth:**
```python
sound = pm.Sound("audio.wav")
spectrum = pm.praat.call(sound, "To Spectrum", True)
cog = pm.praat.call(spectrum, "Get centre of gravity", 2.0)
```

**pladdrr:**
```{r}
sound <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))
spectrum <- sound$to_spectrum(fast = TRUE)
cog <- spectrum$get_centre_of_gravity(power = 2.0)
```

## Batch Processing

### Parselmouth Approach

```python
import parselmouth as pm
import os
import pandas as pd

results = []
for filename in os.listdir('.'):
    if filename.endswith('.wav'):
        sound = pm.Sound(filename)
        pitch = pm.praat.call(sound, "To Pitch", 0.01, 75, 600)
        mean_f0 = pm.praat.call(pitch, "Get mean", 0, 0, "Hertz")
        
        results.append({
            'file': filename,
            'mean_f0': mean_f0
        })

df = pd.DataFrame(results)
df.to_csv('results.csv', index=False)
```

### pladdrr Approach

```{r}
library(pladdrr)

files <- list.files(system.file("extdata", package = "pladdrr"),
                    pattern = "\\.wav$", full.names = TRUE)

results <- lapply(files, function(filepath) {
  sound <- Sound$new(filepath)
  pitch <- sound$to_pitch(time_step = 0.01, pitch_floor = 75, pitch_ceiling = 600)
  mean_f0 <- pitch$get_mean(from_time = 0, to_time = 0, unit = "hertz")
  
  data.frame(
    file = basename(filepath),
    mean_f0 = mean_f0
  )
})

results_df <- do.call(rbind, results)
write.csv(results_df, file.path(tempdir(), "results.csv"), row.names = FALSE)
```

## Working with Data Frames

### Extracting Time Series Data

**Parselmouth:**
```python
sound = pm.Sound("audio.wav")
pitch = pm.praat.call(sound, "To Pitch", 0.01, 75, 600)

# Extract values manually
times = []
frequencies = []
for i in range(pitch.n_frames):
    time = pitch.xs[i]
    f0 = pm.praat.call(pitch, "Get value at time", time, "Hertz", "Linear")
    times.append(time)
    frequencies.append(f0)

df = pd.DataFrame({'time': times, 'frequency': frequencies})
```

**pladdrr:**
```{r}
sound <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))
pitch <- sound$to_pitch(time_step = 0.01, pitch_floor = 75, pitch_ceiling = 600)

# Direct conversion to data frame
pitch_data <- pitch$as_data_frame()
# Returns data.frame with 'time' and 'frequency' columns
```

## Visualization

### Parselmouth with matplotlib

```python
import matplotlib.pyplot as plt

sound = pm.Sound("audio.wav")
pitch = pm.praat.call(sound, "To Pitch", 0.01, 75, 600)

# Manual extraction
times = [pitch.xs[i] for i in range(pitch.n_frames)]
f0s = [pm.praat.call(pitch, "Get value at time", t, "Hertz", "Linear") 
       for t in times]

plt.figure(figsize=(10, 4))
plt.plot(times, f0s)
plt.xlabel('Time (s)')
plt.ylabel('Frequency (Hz)')
plt.title('Pitch Contour')
plt.show()
```

### pladdrr with ggplot2

```{r}
library(pladdrr)
library(ggplot2)

sound <- Sound$new(system.file("extdata", "test.wav", package = "pladdrr"))
pitch <- sound$to_pitch(time_step = 0.01, pitch_floor = 75, pitch_ceiling = 600)
pitch_data <- pitch$as_data_frame()

ggplot(pitch_data, aes(x = time, y = frequency)) +
  geom_line() +
  labs(title = "Pitch Contour", x = "Time (s)", y = "Frequency (Hz)") +
  theme_minimal()
```

## Advanced Workflows

### Voice Analysis Pipeline

**Parselmouth:**
```python
def analyze_voice(filename):
    sound = pm.Sound(filename)
    
    pitch = pm.praat.call(sound, "To Pitch", 0.01, 75, 600)
    mean_f0 = pm.praat.call(pitch, "Get mean", 0, 0, "Hertz")
    
    harmonicity = pm.praat.call(sound, "To Harmonicity (cc)", 0.01, 75, 0.1, 1.0)
    mean_hnr = pm.praat.call(harmonicity, "Get mean", 0, 0)
    
    pointprocess = pm.praat.call(sound, "To PointProcess (periodic, cc)", 75, 600)
    jitter = pm.praat.call(pointprocess, "Get jitter (local)", 0, 0, 0.0001, 0.02, 1.3)
    
    return {
        'mean_f0': mean_f0,
        'mean_hnr': mean_hnr,
        'jitter': jitter
    }
```

**pladdrr:**
```{r}
analyze_voice <- function(filename) {
  sound <- Sound$new(filename)
  
  pitch <- sound$to_pitch(time_step = 0.01, pitch_floor = 75, pitch_ceiling = 600)
  mean_f0 <- pitch$get_mean(from_time = 0, to_time = 0, unit = "hertz")
  
  harmonicity <- sound$to_harmonicity_cc(time_step = 0.01, min_pitch = 75,
                                         silence_threshold = 0.1, periods_per_window = 1.0)
  mean_hnr <- harmonicity$get_mean(from_time = 0, to_time = 0)
  
  pointprocess <- sound$to_pointprocess_periodic_cc(pitch_floor = 75, pitch_ceiling = 600)
  jitter <- pointprocess$get_jitter_local(from_time = 0, to_time = 0,
                                          period_floor = 0.0001, period_ceiling = 0.02,
                                          max_period_factor = 1.3)
  
  list(
    mean_f0 = mean_f0,
    mean_hnr = mean_hnr,
    jitter = jitter
  )
}
```

## Integration with Data Science Tools

### Parselmouth + pandas/numpy

```python
import numpy as np
import pandas as pd

files = ['file1.wav', 'file2.wav', 'file3.wav']
results = [analyze_voice(f) for f in files]
df = pd.DataFrame(results)
df['speaker'] = ['A', 'B', 'C']
```

### pladdrr + tidyverse

```{r}
# Requires the dplyr, purrr, and tidyr packages (not pladdrr dependencies;
# install separately if needed)
library(pladdrr)
library(dplyr)
library(purrr)
library(tidyr)

wav <- system.file("extdata", "test.wav", package = "pladdrr")
results <- tibble(
  file = c(wav, wav, wav),
  speaker = c('A', 'B', 'C')
) %>%
  mutate(
    analysis = map(file, analyze_voice)
  ) %>%
  unnest_wider(analysis)
```

## Advantages of pladdrr

1. **No Python Dependency**: Pure R package
2. **Better IDE Support**: Full autocomplete and documentation
3. **Type Safety**: Named parameters prevent errors
4. **R Ecosystem**: Integration with tidyverse, ggplot2, etc.
5. **Self-Documenting**: Method names describe functionality

## When to Use Each

### Use Parselmouth if:
- You're already working in Python
- You need Python-specific libraries (scipy, scikit-learn)
- Your team uses Python

### Use pladdrr if:
- You're working in R
- You need R-native tooling for statistics and visualization
- You need advanced visualizations (ggplot2)
- You prefer type-safe, self-documenting code

## Common Issues and Solutions

### Issue 1: String-based commands

**Problem**: In Parselmouth, you must remember exact command strings.

**Solution**: pladdrr provides method autocomplete in RStudio/VS Code.

### Issue 2: Positional parameters

**Problem**: Easy to mix up parameter order in Parselmouth.

**Solution**: pladdrr uses named parameters for clarity.

### Issue 3: Documentation

**Problem**: Parselmouth refers to Praat documentation.

**Solution**: pladdrr has R help docs: `?Sound`, `?Pitch`, etc.

## Getting Help

- Package documentation: `help(package = "pladdrr")`
- Vignettes: `vignette(package = "pladdrr")`
- Maintainer: fredrik.nylen@umu.se
- Compare with Praat manual: https://www.fon.hum.uva.nl/praat/manual/

## Conclusion

Both Parselmouth and pladdrr provide access to Praat's functionality.
pladdrr differs in offering a direct, object-oriented (R6) interface with
named parameters, instead of a string-based `praat.call()` dispatcher, and
integrates with the R ecosystem (tidyverse, ggplot2).

The transition from Parselmouth to pladdrr mainly involves replacing `praat.call()` calls with direct method calls, using the naming conventions shown above.
