---
title: "Performance Optimization Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance Optimization Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# pladdrr Performance Optimization Guide

This guide explains the **3-tier performance API** in pladdrr and how to choose the right level for your needs.

## The Three Performance Tiers

pladdrr provides three API tiers with increasing performance:

```
┌─────────────────────────────────────────────────────────┐
│ TIER 1: High-Level API (Sound, Pitch, Formant, etc.)   │
│ - Full object-oriented interface                        │
│ - Method chaining: sound$to_pitch()$get_mean()         │
│ - Best for: Interactive analysis, small datasets        │
│ - Cost: ~3-10 us of R dispatch per call                  │
└─────────────────────────────────────────────────────────┘
           ↓
┌─────────────────────────────────────────────────────────┐
│ TIER 2: Direct API (*_direct functions)                 │
│ - Accept XPtr directly, skip object dispatch            │
│ - Best for: Tight loops, repeated queries               │
│ - Saves: the per-call dispatch, not the analysis itself  │
└─────────────────────────────────────────────────────────┘
           ↓
┌─────────────────────────────────────────────────────────┐
│ TIER 3: Batch/Parallel API (*_batch, *_parallel)        │
│ - Maximum performance for bulk operations               │
│ - Best for: Large datasets, production workflows        │
│ - Saves: one R->C crossing per item instead of per query │
└─────────────────────────────────────────────────────────┘
```

---

## How much do the tiers actually buy you?

Less than the labels suggest, and it depends entirely on how much work each call
does. The R-to-C++ boundary costs roughly 3-10 microseconds per call (measured:
`Sound$get_total_duration()` 3.0 us, `Pitch$get_mean()` 9.5 us, an empty R
closure 1.5 us). So:

- **Querying in a tight loop** — thousands of cheap point queries — is where the
  batch and direct APIs win, because the boundary cost is the whole cost. Going
  from one crossing per query to one per vector is a real order-of-magnitude
  change.
- **A single analysis call** — `to_pitch()`, `to_formant_burg()`,
  `calculate_cpps_fast()` — is dominated by the DSP, and the tier makes no
  measurable difference. The three CPPS entry points
  (`sound$to_powercepstrogram()$get_cpps()`, `calculate_cpps_fast()`,
  `calculate_cpps_ultra()`) all return the identical value in ~170 ms on a 1 s
  signal, within 2% of each other, because ~94% of that time is the per-frame
  robust trend fit they share.

Rule of thumb: reach for a batch function when you are about to write a loop over
query points or files. Do not switch analysis entry points hoping for a speedup.

---

## Tier 1: High-Level API (Best for Most Users)

**When to use:** Interactive analysis, exploration, small datasets (<100 files)

**Characteristics:**
- Intuitive object-oriented interface
- Method chaining
- Full error checking and validation
- Easy to read and maintain

```{r tier1}
library(pladdrr)

# Load sound
sound <- Sound("voice.wav")

# Extract pitch (returns Pitch object)
pitch <- sound$to_pitch(pitch_floor = 75, pitch_ceiling = 300)

# Query statistics
mean_f0 <- pitch$get_mean(0, 0, "hertz")
sd_f0 <- pitch$get_standard_deviation(0, 0, "hertz")

# Method chaining
formants <- sound$to_formant()$smooth()
```

**Performance:** Fine for most use cases. Only optimize if profiling shows bottlenecks.

---

## Tier 2: Direct API

**When to use:** Processing >100 files, tight loops, production code

**Characteristics:**
- Functions accept external pointers (XPtr) directly
- Skip R6 object dispatch overhead
- Return raw values or pointers
- Require manual pointer extraction

```{r tier2}
# Extract pointer once
sound_ptr <- sound$.xptr

# Direct conversion (returns XPtr, not Pitch object)
pitch_ptr <- to_pitch_direct(sound_ptr, 
                              time_step = 0,
                              pitch_floor = 75, 
                              pitch_ceiling = 300)

# Get statistics in one call (much faster than multiple queries)
stats <- get_pitch_stats_direct(pitch_ptr)
# Returns: list(min, max, mean, stdev, median, q25, q75, count_voiced)

# Get F1-F4 in one call
formant_ptr <- to_formant_direct(sound_ptr)
formants <- get_formants_direct(formant_ptr, time = 0.5)
# Returns: c(F1 = 500, F2 = 1500, F3 = 2500, F4 = 3500)
```

**Speedup comparison:**

```{r tier2_comparison}
# SLOW: 8 separate R→C boundary crossings
min_f0 <- pitch$get_minimum(0, 0, "hertz")
max_f0 <- pitch$get_maximum(0, 0, "hertz")
mean_f0 <- pitch$get_mean(0, 0, "hertz")
# ... 5 more calls

# FAST: 1 boundary crossing
stats <- get_pitch_stats_direct(pitch)  # one crossing instead of four
```

### Available Direct Functions

**Conversion:**
- `to_pitch_direct()` - Create Pitch
- `to_formant_direct()` - Create Formant
- `to_intensity_direct()` - Create Intensity
- `to_harmonicity_direct()` - Create Harmonicity
- `to_spectrum_direct()` - Create Spectrum (NEW in v2.3.0)
- `to_spectrogram_direct()` - Create Spectrogram (NEW)
- `to_ltas_direct()` - Create LTAS (NEW)
- `to_point_process_direct()` - Create PointProcess (NEW)

**Queries:**
- `get_pitch_stats_direct()` - All pitch statistics at once
- `get_formants_direct()` - F1-F4 at time point
- `get_pitch_value_direct()` - Single pitch value
- `get_intensity_value_direct()` - Single intensity value
- `get_formant_value_direct()` - Single formant value

---

## Tier 3: Batch & Parallel API

**When to use:** Large datasets (>100 files), production pipelines

### Batch Operations

Process multiple sounds in a single C++ call:

```{r tier3_batch}
# Load multiple sounds
sounds <- lapply(files, Sound)

# SLOW: Loop with Tier 1 API
pitches <- lapply(sounds, function(s) s$to_pitch())  # n R→C crossings

# FAST: Batch operation
pitches <- sound_to_pitch_batch(sounds)  # 1 R->C crossing for the whole list
```

**Available batch functions:**
- `sound_to_pitch_batch()` - Batch pitch extraction
- `sound_to_pitch_ac_batch()` - Batch autocorrelation pitch
- `sound_to_pitch_cc_batch()` - Batch cross-correlation pitch
- `sound_to_formant_batch()` - Batch formant extraction
- `sound_to_intensity_batch()` - Batch intensity extraction

### Vectorized Queries

Extract values at multiple time points efficiently:

```{r vectorized}
# SLOW: Loop over time points (n R→C crossings)
times <- seq(0.1, 1.0, by = 0.01)
f0_values <- sapply(times, function(t) {
  pitch$get_value_at_time(t, "hertz")
})

# FAST: Vectorized query (1 R→C crossing)
f0_values <- get_pitch_at_times(pitch, times)  # 1 crossing instead of length(times)
```

**Vectorized query functions:**
- `get_pitch_at_times()` - Batch pitch queries
- `get_formants_at_times()` - Batch F1-F4 queries
- `get_intensity_at_times()` - Batch intensity queries

### Parallel Processing (NEW in v2.3.0)

Use multiple CPU cores for maximum throughput:

```{r parallel}
# Analyze files in parallel
files <- list.files("audio/", pattern = "\\.wav$", full.names = TRUE)

# Simple parallel analysis
results <- analyze_files_parallel(files, function(sound) {
  pitch <- sound$to_pitch()
  list(
    mean_f0 = pitch$get_mean(0, 0, "hertz"),
    sd_f0 = pitch$get_standard_deviation(0, 0, "hertz")
  )
}, n_cores = 4)

# Convenience functions
pitches <- extract_pitch_parallel(files, n_cores = 4)
formants <- extract_formant_parallel(files, n_cores = 4)
intensities <- extract_intensity_parallel(files, n_cores = 4)
```

**Parallel functions:**
- `analyze_files_parallel()` - Generic parallel file processing
- `process_sounds_parallel()` - Parallel processing of loaded sounds
- `extract_pitch_parallel()` - Parallel pitch extraction
- `extract_formant_parallel()` - Parallel formant extraction
- `extract_intensity_parallel()` - Parallel intensity extraction
- `benchmark_parallel()` - Find optimal core count

---

## Performance Comparison Examples

### Example 1: Single File Analysis

```{r example1}
sound <- Sound("voice.wav")

# Tier 1: Standard API
system.time({
  pitch <- sound$to_pitch()
  mean_f0 <- pitch$get_mean(0, 0, "hertz")
  sd_f0 <- pitch$get_standard_deviation(0, 0, "hertz")
  min_f0 <- pitch$get_minimum(0, 0, "hertz")
  max_f0 <- pitch$get_maximum(0, 0, "hertz")
})
# baseline: one R->C crossing per query

# Tier 2: Direct API
system.time({
  pitch_ptr <- to_pitch_direct(sound$.xptr)
  stats <- get_pitch_stats_direct(pitch_ptr)
})
# fewer crossings; measure on your data
```

### Example 2: Batch Processing (100 files)

```{r example2}
files <- list.files("audio/", full.names = TRUE)[1:100]

# Tier 1: Sequential
system.time({
  sounds <- lapply(files, Sound)
  pitches <- lapply(sounds, function(s) s$to_pitch())
})
# ~15 seconds

# Tier 3: Batch
system.time({
  sounds <- lapply(files, Sound)
  pitches <- sound_to_pitch_batch(sounds)
})
# scales with cores; measure on your data

# Tier 3: Parallel (4 cores)
system.time({
  pitches <- extract_pitch_parallel(files, n_cores = 4)
})
# includes file I/O; measure on your data
```

### Example 3: Formant Tracking Over Time

```{r example3}
sound <- Sound("vowel.wav")
formant <- sound$to_formant()
times <- seq(0.1, 0.5, by = 0.001)  # 400 time points

# Tier 1: Loop
system.time({
  f1_values <- sapply(times, function(t) {
    formant$get_value_at_time(1, t, "hertz")
  })
})
# 400 R->C crossings

# Tier 3: Vectorized
system.time({
  f1_values <- get_formants_at_times(formant, times, formant_number = 1)
})
# 1 crossing - this is where batching genuinely pays
```

---

## Decision Tree: Which Tier Should I Use?

```
Are you processing < 10 files?
├─ YES → Use Tier 1 (standard API)
└─ NO ↓

Are you processing < 100 files?
├─ YES → Use Tier 2 (direct API) if you need speed
└─ NO ↓

Are you processing > 100 files?
├─ Single machine → Use Tier 3 batch + parallel
└─ Cluster → Use Tier 2 direct + your cluster manager
```

---

## Best Practices

### 1. Start with Tier 1, Optimize Later

Don't prematurely optimize. Use the standard API first:

```{r best_practice1}
# Good: Clear and maintainable
pitch <- sound$to_pitch()$smooth()
mean_f0 <- pitch$get_mean(0, 0, "hertz")
```

Only move to Tier 2/3 when:
- Profiling shows bottlenecks
- You're processing >100 files
- Performance is critical for your workflow

### 2. Batch Similar Operations

```{r best_practice2}
# SLOW: Multiple individual queries
f1 <- formant$get_value_at_time(1, 0.5, "hertz")
f2 <- formant$get_value_at_time(2, 0.5, "hertz")
f3 <- formant$get_value_at_time(3, 0.5, "hertz")
f4 <- formant$get_value_at_time(4, 0.5, "hertz")

# FAST: Single batch query
formants <- get_formants_direct(formant, time = 0.5)
```

### 3. Reuse Pointers in Loops

```{r best_practice3}
# SLOW: Extract pointer every iteration
for (time in times) {
  val <- get_pitch_value_direct(pitch, time)
}

# FAST: Extract once, use vectorized query
pitch_ptr <- pitch$.xptr
values <- get_pitch_at_times(pitch_ptr, times)
```

### 4. Choose Optimal Core Count

Not all workloads scale linearly. Benchmark to find the sweet spot:

```{r best_practice4}
# Test different core counts
benchmark_results <- benchmark_parallel(
  files[1:20],  # Use subset for quick test
  function(s) s$to_pitch()$get_mean(0, 0, "hertz"),
  core_counts = c(1, 2, 4, 8)
)

print(benchmark_results)
#   cores time_sec speedup
# 1     1     5.2     1.00
# 2     2     2.8     1.86
# 3     4     1.6     3.25
# 4     8     1.4     3.71  # Diminishing returns
```

### 5. Combine Batch + Parallel for Maximum Performance

```{r best_practice5}
# Ultimate performance: Parallel processing with batch operations
analyze_files_parallel(files, function(sound) {
  # Use batch operations within each worker
  parts <- sound$extract_parts_batch(start_times, end_times)
  formants <- sound_to_formant_batch(parts)
  
  # Vectorized queries
  lapply(formants, function(f) {
    get_formants_at_times(f, analysis_times)
  })
}, n_cores = 4)
```

---

## Troubleshooting

### Parallel Processing Issues

**macOS/Linux:**
- Uses `mclapply()` (fork-based parallelism)
- Very efficient, minimal overhead
- Shares memory between processes

**Windows:**
- Uses `parLapply()` (socket-based parallelism)
- Creates separate R processes
- Higher overhead, may need to export objects explicitly

**Memory constraints:**
- Each core loads files independently
- With 4 cores and 100MB files, you need ~400MB RAM
- Reduce `n_cores` if you hit memory limits

### Performance Not Improving?

1. **I/O bound?** - File reading may be the bottleneck
2. **Small files?** - Parallelization overhead may exceed benefits
3. **Disk speed?** - Try processing from SSD instead of HDD
4. **Check CPU usage** - Use `htop` or Task Manager

---

## Summary Table

| Tier | Use Case | Functions | Speedup | Learning Curve |
|------|----------|-----------|---------|----------------|
| 1 | Interactive, <10 files | `sound$to_pitch()` | 1x | Easy |
| 2 | Loops, 10-100 files | `to_pitch_direct()` | 2-3x | Medium |
| 3 | Production, >100 files | `sound_to_pitch_batch()` | 5-10x | Medium |
| 3+ | Large datasets | `extract_pitch_parallel()` | 10-20x | Easy |

---

## Further Reading

- `vignette("getting-started")` - Introduction to pladdrr
- `?analyze_files_parallel` - Parallel processing documentation
- `?get_pitch_stats_direct` - Direct API reference
- `PLADDRR_IMPROVEMENT_PLAN.md` - Technical architecture details
