---
title: "Performance Optimization Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance Optimization Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# pladdrr Performance Optimization Guide

This guide explains the **3-tier performance API** in pladdrr and how to choose the right level for your needs.

## The Three Performance Tiers

pladdrr provides three API tiers, each trading interface simplicity for lower
per-call overhead:

```
┌─────────────────────────────────────────────────────────┐
│ TIER 1: High-Level API (Sound, Pitch, Formant, etc.)   │
│ - Full object-oriented interface                        │
│ - Method chaining: sound$to_pitch()$get_mean()         │
│ - Best for: Interactive analysis, small datasets        │
└─────────────────────────────────────────────────────────┘
           ↓
┌─────────────────────────────────────────────────────────┐
│ TIER 2: Direct API (*_direct functions)                 │
│ - Accept XPtr directly, skip object dispatch            │
│ - Best for: Tight loops, repeated queries                │
│ - Reduces per-call R dispatch overhead                  │
└─────────────────────────────────────────────────────────┘
           ↓
┌─────────────────────────────────────────────────────────┐
│ TIER 3: Batch/Parallel API (*_batch, *_parallel)        │
│ - Designed for bulk operations                          │
│ - Best for: Large datasets, production workflows        │
│ - Replaces many R->C crossings with one per call        │
└─────────────────────────────────────────────────────────┘
```

---

## When do the tiers actually matter?

It depends entirely on how much work each call does relative to the R-to-C++
boundary it crosses.

- **Querying in a tight loop** — thousands of cheap point queries — is where
  the batch and direct APIs are worth reaching for, because the per-call
  dispatch overhead is a larger share of the total cost. Going from one
  crossing per query to one crossing per vector changes the shape of that
  cost.
- **A single analysis call** — `to_pitch()`, `to_formant_burg()`,
  `calculate_cpps_fast()` — is dominated by the underlying DSP work, and the
  tier you call it through makes little practical difference. The different
  CPPS entry points (`sound$to_powercepstrogram()$get_cpps()`,
  `calculate_cpps_fast()`, `calculate_cpps_ultra()`) all compute the same
  value using the same core algorithm.

Rule of thumb: reach for a batch or direct function when you are about to
write a loop over query points or files. Switching analysis entry points for
a single call is not expected to help — measure your own workload with
`system.time()` if you're unsure.

---

## Tier 1: High-Level API (Best for Most Users)

**When to use:** Interactive analysis, exploration, small datasets (<100 files)

**Characteristics:**
- Intuitive object-oriented interface
- Method chaining
- Full error checking and validation
- Easy to read and maintain

```{r tier1}
library(pladdrr)

wav_file <- system.file("extdata", "test.wav", package = "pladdrr")

# Load sound
sound <- Sound(wav_file)

# Extract pitch (returns Pitch object)
pitch <- sound$to_pitch(pitch_floor = 75, pitch_ceiling = 300)

# Query statistics
mean_f0 <- pitch$get_mean(0, 0, "hertz")
sd_f0 <- pitch$get_standard_deviation(0, 0, "hertz")

# Method chaining
mean_f1 <- sound$to_formant()$get_mean(1, 0, 0, "hertz")
```

Adequate for most use cases. Only reach for Tier 2/3 if profiling your own
code shows a bottleneck.

---

## Tier 2: Direct API

**When to use:** Processing >100 files, tight loops, production code

**Characteristics:**
- Functions accept external pointers (XPtr) directly
- Skip R6 object dispatch overhead
- Return raw values or pointers
- Require manual pointer extraction

```{r tier2}
# Extract pointer once
sound_ptr <- sound$.xptr

# Direct conversion (returns XPtr, not Pitch object)
pitch_ptr <- to_pitch_direct(sound_ptr,
                              time_step = 0,
                              pitch_floor = 75,
                              pitch_ceiling = 300)

# Get statistics in one call
stats <- get_pitch_stats_direct(pitch_ptr)
# Returns: list(min, max, mean, stdev, median, q25, q75, count_voiced)

# Get F1-F4 in one call
formant_ptr <- to_formant_direct(sound_ptr)
formants <- get_formants_direct(formant_ptr, time = 0.5)
# Returns: c(F1 = 500, F2 = 1500, F3 = 2500, F4 = 3500)
```

### Fewer R→C boundary crossings

```{r tier2_comparison}
# Tier 1 approach: one boundary crossing per query
min_f0 <- pitch$get_minimum(0, 0, "hertz")
max_f0 <- pitch$get_maximum(0, 0, "hertz")
mean_f0 <- pitch$get_mean(0, 0, "hertz")
# ... 5 more calls

# Tier 2 approach: one boundary crossing for all statistics
stats <- get_pitch_stats_direct(pitch)
```

### Available Direct Functions

**Conversion:**
- `to_pitch_direct()` - Create Pitch
- `to_formant_direct()` - Create Formant
- `to_intensity_direct()` - Create Intensity
- `to_harmonicity_direct()` - Create Harmonicity
- `to_spectrum_direct()` - Create Spectrum
- `to_spectrogram_direct()` - Create Spectrogram
- `to_ltas_direct()` - Create LTAS
- `to_point_process_direct()` - Create PointProcess

**Queries:**
- `get_pitch_stats_direct()` - All pitch statistics at once
- `get_formants_direct()` - F1-F4 at time point
- `get_pitch_value_direct()` - Single pitch value
- `get_intensity_value_direct()` - Single intensity value
- `get_formant_value_direct()` - Single formant value

---

## Tier 3: Batch & Parallel API

**When to use:** Large datasets (>100 files), production pipelines

### Batch Operations

Process multiple sounds in a single C++ call:

```{r tier3_batch}
# In practice `files` would be a vector of paths to distinct recordings, e.g.
# list.files("audio/", pattern = "\\.wav$", full.names = TRUE). This example
# repeats the bundled test file to keep the vignette self-contained.
files <- rep(wav_file, 3)

# Load multiple sounds
sounds <- lapply(files, Sound)

# Tier 1 approach: one R->C crossing per sound
pitches <- lapply(sounds, function(s) s$to_pitch())

# Tier 3 approach: one R->C crossing for the whole list
pitches <- sound_to_pitch_batch(sounds)
```

**Available batch functions:**
- `sound_to_pitch_batch()` - Batch pitch extraction
- `sound_to_pitch_ac_batch()` - Batch autocorrelation pitch
- `sound_to_pitch_cc_batch()` - Batch cross-correlation pitch
- `sound_to_formant_batch()` - Batch formant extraction
- `sound_to_intensity_batch()` - Batch intensity extraction

### Vectorized Queries

Extract values at multiple time points in one call:

```{r vectorized}
# Tier 1 approach: one R->C crossing per time point
times <- seq(0.1, 1.0, by = 0.01)
f0_values <- sapply(times, function(t) {
  pitch$get_value_at_time(t, "hertz")
})

# Tier 3 approach: one R->C crossing for the whole vector
f0_values <- get_pitch_at_times(pitch, times)
```

**Vectorized query functions:**
- `get_pitch_at_times()` - Batch pitch queries
- `get_formants_at_times()` - Batch F1-F4 queries
- `get_intensity_at_times()` - Batch intensity queries

### Parallel Processing

Distribute file-level work across multiple CPU cores:

```{r parallel}
# Analyze files in parallel. In practice `files` is a vector of paths to
# distinct recordings, e.g. list.files("audio/", pattern = "\\.wav$",
# full.names = TRUE); this example repeats the bundled test file.
files <- rep(wav_file, 4)

# Simple parallel analysis
results <- analyze_files_parallel(files, function(sound) {
  pitch <- sound$to_pitch()
  list(
    mean_f0 = pitch$get_mean(0, 0, "hertz"),
    sd_f0 = pitch$get_standard_deviation(0, 0, "hertz")
  )
}, n_cores = 2)

# Convenience functions
pitches <- extract_pitch_parallel(files, n_cores = 2)
formants <- extract_formant_parallel(files, n_cores = 2)
intensities <- extract_intensity_parallel(files, n_cores = 2)
```

**Parallel functions:**
- `analyze_files_parallel()` - Generic parallel file processing
- `process_sounds_parallel()` - Parallel processing of loaded sounds
- `extract_pitch_parallel()` - Parallel pitch extraction
- `extract_formant_parallel()` - Parallel formant extraction
- `extract_intensity_parallel()` - Parallel intensity extraction
- `benchmark_parallel()` - Measure your own workload across core counts

---

## Tier Comparison Examples

The examples below use `system.time()` so you can measure each approach
against your own files and hardware — the numbers vary too much by machine,
file size, and workload to be worth quoting here.

### Example 1: Single File Analysis

```{r example1}
sound <- Sound(wav_file)

# Tier 1: Standard API
system.time({
  pitch <- sound$to_pitch()
  mean_f0 <- pitch$get_mean(0, 0, "hertz")
  sd_f0 <- pitch$get_standard_deviation(0, 0, "hertz")
  min_f0 <- pitch$get_minimum(0, 0, "hertz")
  max_f0 <- pitch$get_maximum(0, 0, "hertz")
})
# measure on your own data

# Tier 2: Direct API
system.time({
  pitch_ptr <- to_pitch_direct(sound$.xptr)
  stats <- get_pitch_stats_direct(pitch_ptr)
})
# fewer crossings; measure on your own data
```

### Example 2: Batch Processing

```{r example2}
# In practice `files` is a vector of paths to distinct recordings, e.g.
# list.files("audio/", full.names = TRUE)[1:100]; this example repeats the
# bundled test file to keep the vignette self-contained and fast to build.
files <- rep(wav_file, 20)

# Tier 1: Sequential
system.time({
  sounds <- lapply(files, Sound)
  pitches <- lapply(sounds, function(s) s$to_pitch())
})
# measure on your own data

# Tier 3: Batch
system.time({
  sounds <- lapply(files, Sound)
  pitches <- sound_to_pitch_batch(sounds)
})
# scales with cores; measure on your own data

# Tier 3: Parallel (2 cores)
system.time({
  pitches <- extract_pitch_parallel(files, n_cores = 2)
})
# includes file I/O; measure on your own data
```

### Example 3: Formant Tracking Over Time

```{r example3}
sound <- Sound(wav_file)
formant <- sound$to_formant()
times <- seq(0.1, 0.5, by = 0.001)

# Tier 1: Loop
system.time({
  f1_values <- sapply(times, function(t) {
    formant$get_value_at_time(1, t, "hertz")
  })
})
# one R->C crossing per time point

# Tier 3: Vectorized
system.time({
  f1_values <- get_formants_at_times(formant, times, formant_numbers = 1)
})
# one R->C crossing for the whole vector — this is where batching pays off
```

---

## Decision Tree: Which Tier Should I Use?

```
Are you processing < 10 files?
├─ YES → Use Tier 1 (standard API)
└─ NO ↓

Are you processing < 100 files?
├─ YES → Use Tier 2 (direct API) if profiling shows it helps
└─ NO ↓

Are you processing > 100 files?
├─ Single machine → Use Tier 3 batch + parallel
└─ Cluster → Use Tier 2 direct + your cluster manager
```

---

## Best Practices

### 1. Start with Tier 1, Optimize Later

Don't prematurely optimize. Use the standard API first:

```{r best_practice1}
# Good: Clear and maintainable
pitch <- sound$to_pitch()$smooth()
mean_f0 <- pitch$get_mean(0, 0, "hertz")
```

Only move to Tier 2/3 when:
- Profiling your own code shows a bottleneck
- You're processing >100 files
- Runtime matters for your workflow

### 2. Batch Similar Operations

```{r best_practice2}
# Tier 1: Multiple individual queries
f1 <- formant$get_value_at_time(1, 0.5, "hertz")
f2 <- formant$get_value_at_time(2, 0.5, "hertz")
f3 <- formant$get_value_at_time(3, 0.5, "hertz")
f4 <- formant$get_value_at_time(4, 0.5, "hertz")

# Tier 2: Single batch query
formants <- get_formants_direct(formant, time = 0.5)
```

### 3. Reuse Pointers in Loops

```{r best_practice3}
# Avoid: Repeated single-value queries in a loop
for (time in times) {
  val <- get_pitch_value_direct(pitch, time)
}

# Prefer: One call for the whole vector of times
# (get_pitch_at_times() takes the Pitch object itself, not an XPtr — it
# extracts the pointer internally)
values <- get_pitch_at_times(pitch, times)
```

### 4. Choose a Core Count That Fits Your Workload

Not all workloads scale linearly with core count, and small files can make
parallelization overhead outweigh any benefit. Use `benchmark_parallel()` to
test a range of core counts on a representative subset of your own files
before committing to a setting for a full run:

```{r best_practice4}
# Test different core counts on a subset
benchmark_results <- benchmark_parallel(
  files[1:10],  # Use subset for a quick test
  function(s) s$to_pitch()$get_mean(0, 0, "hertz"),
  core_counts = c(1, 2)  # extend to c(1, 2, 4, 8) etc. up to your machine's cores
)

print(benchmark_results)
# Inspect the returned table to see where returns diminish on your machine
```

### 5. Combine Batch and Parallel Processing

```{r best_practice5}
# Define the per-file constants *inside* the worker function rather than
# capturing them from the enclosing environment. On Windows (and on macOS,
# where analyze_files_parallel() also uses a PSOCK cluster to avoid known
# fork/event-loop issues), each worker is a separate R process that only
# receives the function and file path — free variables referenced from the
# calling environment are not automatically exported and would error with
# "object not found" unless passed via clusterExport() or defined locally
# like this.
analyze_files_parallel(files, function(sound) {
  start_times <- c(0.1, 0.3)
  end_times <- c(0.2, 0.4)
  analysis_times <- seq(0.01, 0.05, by = 0.01)

  # Use batch operations within each worker
  parts <- sound$extract_parts_batch(start_times, end_times)
  formants <- sound_to_formant_batch(parts)

  # Vectorized queries
  lapply(formants, function(f) {
    get_formants_at_times(f, analysis_times)
  })
}, n_cores = 2)
```

---

## Troubleshooting

### Parallel Processing Issues

**macOS/Linux:**
- Uses `mclapply()` (fork-based parallelism)
- Shares memory between processes

**Windows:**
- Uses `parLapply()` (socket-based parallelism)
- Creates separate R processes
- May need to export objects explicitly

**Memory constraints:**
- Each core loads files independently
- With 4 cores and 100MB files, plan for roughly 400MB RAM
- Reduce `n_cores` if you hit memory limits

### Runtime Not Improving?

If parallelizing or batching your own workload isn't helping, check:

1. **I/O bound?** - File reading may be the bottleneck, not analysis
2. **Small files?** - Parallelization overhead can exceed any benefit
3. **Disk speed?** - Try processing from SSD instead of HDD
4. **CPU usage** - Use `htop` or Task Manager to see whether cores are saturated

---

## Summary Table

| Tier | Use Case | Functions | Learning Curve |
|------|----------|-----------|----------------|
| 1 | Interactive, <10 files | `sound$to_pitch()` | Easy |
| 2 | Loops, 10-100 files | `to_pitch_direct()` | Medium |
| 3 | Production, >100 files | `sound_to_pitch_batch()` | Medium |
| 3+ | Large datasets | `extract_pitch_parallel()` | Easy |

---

## Further Reading

- `vignette("getting-started")` - Introduction to pladdrr
- `?analyze_files_parallel` - Parallel processing documentation
- `?get_pitch_stats_direct` - Direct API reference
