---
title: "Praat Interpreter: Variables and Expressions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Praat Interpreter: Variables and Expressions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pladdrr)
```

## Introduction

The `pladdrr` package provides complete access to Praat's interpreter system, enabling you to:

- Evaluate Praat expressions directly from R
- Manage persistent variables across expressions
- Work with all 5 Praat variable types (numeric, string, vector, matrix, string array)
- Query the global Praat object list

This vignette demonstrates how to use the interpreter for computational tasks without requiring Praat scripts.

## Standalone Expression Evaluation

The simplest way to use the interpreter is through standalone expression evaluation functions. These create temporary interpreters and return results immediately.

### Numeric Expressions

```{r numeric-expressions}
# Basic arithmetic
praat_eval_numeric("2 + 2")
praat_eval_numeric("sqrt(16) + 2^3")

# Trigonometric functions
praat_eval_numeric("sin(pi/2)")
praat_eval_numeric("cos(0)")

# Complex calculations
praat_eval_numeric("log10(1000)")
```

### String Expressions

```{r string-expressions}
# String concatenation
praat_eval_string('"Hello" + " " + "World"')

# String functions
praat_eval_string('left$("testing", 4)')
praat_eval_string('string$(42)')
```

### Vector Expressions

```{r vector-expressions}
# Create vectors
vec <- praat_eval_vector("{ 1, 2, 3, 4, 5 }")
vec

# Vector with decimals
vec2 <- praat_eval_vector("{ 1.1, 2.2, 3.3 }")
vec2
```

### Matrix Expressions

```{r matrix-expressions}
# Create 2x2 matrix
mat <- praat_eval_matrix("{{ 1, 2 }, { 3, 4 }}")
mat

# Matrix dimensions are preserved
dim(mat)
```

### String Array Expressions

```{r string-array-expressions}
# Create string arrays
names <- praat_eval_string_array('{ "Alice", "Bob", "Charlie" }')
names
```

## Persistent Interpreter with State

For more complex workflows, use the `PraatInterpreter` R6 class which maintains state between operations.

### Creating an Interpreter

```{r create-interpreter}
# Create a persistent interpreter instance
interp <- PraatInterpreter$new()
```

### Setting Variables

The interpreter automatically detects variable types and adds appropriate Praat suffixes:

```{r set-variables}
# Numeric variables (no suffix)
interp$set_variable("x", 42)
interp$set_variable("pi_value", 3.14159)

# String variables ($ suffix auto-added)
interp$set_variable("name", "John")
interp$set_variable("greeting", "Hello")

# Vector variables (# suffix auto-added)
interp$set_variable("data", c(1.5, 2.5, 3.5, 4.5))

# Matrix variables (## suffix auto-added)
interp$set_variable("matrix", matrix(1:6, nrow = 2, ncol = 3))

# String array variables ($# suffix auto-added)
interp$set_variable("names", c("alpha", "beta", "gamma"))
```

### Getting Variables

Retrieve variables using Praat suffix notation:

```{r get-variables}
# Get numeric variable
interp$get_variable("x")

# Get string variable (must include $ suffix)
interp$get_variable("name$")

# Get vector (# suffix)
interp$get_variable("data#")

# Get matrix (## suffix)
interp$get_variable("matrix##")

# Get string array ($# suffix)
interp$get_variable("names$#")
```

### Using Variables in Expressions

Variables can be used directly in expressions:

```{r variables-in-expressions}
# Set some variables
interp$set_variable("a", 10)
interp$set_variable("b", 20)

# Use in arithmetic
result <- interp$eval("a + b")
result

result <- interp$eval("a * b")
result

# String operations with variables
interp$set_variable("first", "hello")
interp$set_variable("second", "world")

result <- interp$eval('first$ + " " + second$')
result
```

### Vector and Matrix Operations

```{r vector-matrix-ops}
# Create a vector
interp$set_variable("vec", c(1, 2, 3, 4, 5))

# Get vector size
size <- interp$eval("size(vec#)")
size

# Sum vector
total <- interp$eval("sum(vec#)")
total

# Mean of vector
mean_val <- interp$eval("mean(vec#)")
mean_val

# Create a matrix
interp$set_variable("mat", matrix(1:4, nrow = 2, ncol = 2))

# Query matrix dimensions
rows <- interp$eval("numberOfRows(mat##)")
cols <- interp$eval("numberOfColumns(mat##)")
c(rows = rows, cols = cols)
```

## Praat Variable Type Reference

Praat uses suffix notation to distinguish variable types:

| R Type | Praat Suffix | Example Variable | Set | Get |
|--------|--------------|------------------|-----|-----|
| Numeric scalar | (none) | `x` | `set_variable("x", 42)` | `get_variable("x")` |
| String | `$` | `name$` | `set_variable("name", "test")` | `get_variable("name$")` |
| Numeric vector | `#` | `data#` | `set_variable("data", c(1,2,3))` | `get_variable("data#")` |
| Numeric matrix | `##` | `mat##` | `set_variable("mat", matrix(...))` | `get_variable("mat##")` |
| String array | `$#` | `names$#` | `set_variable("names", c("a","b"))` | `get_variable("names$#")` |

**Key Points:**

- When **setting** variables, omit the suffix - it's added automatically
- When **getting** variables, include the appropriate suffix
- Use variables **in expressions** with their full suffix notation

## Object Management

Query Praat's global object list (useful when objects are created through other means):

```{r object-management}
# Get count of objects
count <- praat_object_count()
count

# List all objects (returns data frame)
objects <- praat_list_objects()
objects

# Also available via R6 instance
count2 <- interp$object_count()
objects2 <- interp$list_objects()
```

**Note:** In library mode (without GUI), object creation commands like `Create Sound...` are not available. The object management functions are ready for use when GUI actions are enabled or when working with objects created through pladdrr's direct API (e.g., `sound_load()`, `pitch_create()`, etc.).

## Practical Example: Statistical Calculations

```{r practical-example}
# Create an interpreter for a statistical workflow
stats <- PraatInterpreter$new()

# Set up data
stats$set_variable("values", c(12.5, 15.3, 11.8, 14.2, 13.7, 16.1))

# Calculate statistics using Praat functions
n <- stats$eval("size(values#)")
mean_val <- stats$eval("mean(values#)")
stdev <- stats$eval("stdev(values#)")

# Results
cat("Sample size:", n, "\n")
cat("Mean:", mean_val, "\n")
cat("Std Dev:", stdev, "\n")

# Calculate z-scores manually
stats$set_variable("mean", mean_val)
stats$set_variable("sd", stdev)

# For each value, could calculate (x - mean) / sd
# This demonstrates variable persistence across multiple operations
```

## Performance Considerations

- **Standalone functions** (`praat_eval_*`) create temporary interpreters - use for one-off calculations
- **Persistent interpreters** (`PraatInterpreter$new()`) reuse state - use for workflows with multiple related operations
- **Variable overhead** is minimal
- **Type conversion** between R and Praat is automatic and efficient

## Best Practices

1. **Use persistent interpreters for workflows**: When you need multiple related calculations, create one interpreter instance and reuse it

2. **Name variables clearly**: Since Praat and R variables are separate, use descriptive names to avoid confusion

3. **Remember suffix notation**: When getting variables or using them in expressions, always include the appropriate suffix

4. **Leverage type auto-detection**: Let `set_variable()` automatically determine the correct Praat type from your R object

5. **Check object types**: Use `class()`, `dim()`, etc. to verify R objects before setting as Praat variables

## Summary

The Praat interpreter integration provides:

- Access to Praat's expression evaluation
- All 5 variable types (numeric, string, vector, matrix, string array)
- Persistent state management via `PraatInterpreter`
- Automatic type detection and conversion between R and Praat
- Object list queries
- R6 and standalone functional APIs

The interpreter can be combined with pladdrr's other API functions for phonetic analysis workflows.

## See Also

- `vignette("getting-started")` - Introduction to pladdrr
- `vignette("formant-analysis")` - Formant extraction workflows
- `?PraatInterpreter` - R6 class documentation
- `?praat_eval_numeric` - Expression evaluation functions
