---
title: "Speech Synthesis with KlattGrid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speech Synthesis with KlattGrid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE  # Disable execution during package build; chunks are verified manually
)
```

## Introduction

KlattGrid is pladdrr's implementation of the Klatt cascade/parallel formant synthesizer, for generating synthetic speech from parametric descriptions. Unlike recording or playing back audio, KlattGrid creates speech from parameters by specifying:

- **Pitch contours** (fundamental frequency)
- **Formant frequencies** (F1, F2, F3, ...)
- **Formant bandwidths** (spectral width)
- **Voicing amplitude** (phonation intensity)

This approach is used for:

- Voice manipulation research
- Perceptual experiments with controlled stimuli
- Teaching phonetics concepts
- Speech resynthesis with modified parameters

```{r setup, message=FALSE}
library(pladdrr)
library(ggplot2)
```

## Why Speech Synthesis?

### Research Applications

KlattGrid enables experiments that natural speech cannot provide:

1. **Independent control** - Change F1 independently of F2, not possible in human speech
2. **Parametric variation** - Generate continua (e.g., /i/ to /u/ in 10 steps)
3. **Non-natural voices** - Male formants + female pitch
4. **Reproducibility** - Exact same stimulus across sessions

### The Klatt Synthesizer

Dennis Klatt's 1980 formant synthesizer is a cascade/parallel model of parametric speech synthesis. It models the vocal tract as a cascade of resonant filters (formants) excited by a periodic source (voicing) or noise.

pladdrr provides direct R access to Praat's KlattGrid implementation.

## Basic Usage

### Recommended: Pre-configured Vowels

The safest way to create a KlattGrid is with `KlattGrid_createFromVowel()`:

```{r basic-vowel}
# Synthesize /a/ vowel (low central)
kg <- KlattGrid_createFromVowel(
  duration = 0.5,           # 500 ms
  f0start = 120,            # Pitch: 120 Hz (adult male)
  f1 = 800, b1 = 80,        # First formant + bandwidth
  f2 = 1200, b2 = 120,      # Second formant + bandwidth
  f3 = 2500, b3 = 150       # Third formant + bandwidth
)

# Synthesize to audio
sound <- kg$to_sound()

cat("Created sound:\n")
cat("  Duration:", sound$get_duration(), "s\n")
cat("  Samples:", sound$get_number_of_samples(), "\n")
cat("  Sampling rate:", sound$get_sampling_frequency(), "Hz\n")
```

### Save to File

```{r save-file, eval=FALSE}
# Save synthesized audio
sound$save("vowel_a.wav", "WAV")

# Save KlattGrid parameters
kg$save("vowel_a.KlattGrid")
```

### Avoid Empty Grids

Do not use the empty constructor for synthesis:

```{r empty-grid-warning, eval=FALSE}
# THIS WILL ERROR
kg <- KlattGrid(tmin = 0, tmax = 1, numberOfFormants = 5)
sound <- kg$to_sound()  # errors: "Pitch tier should not be empty."
```

**Why?** Empty grids need complete initialization:
- Pitch tier (fundamental frequency)
- Voicing amplitude tier (phonation)
- All formant frequency tiers (F1, F2, F3, ...)
- All formant bandwidth tiers (B1, B2, B3, ...)

**Always use `KlattGrid_createFromVowel()` or `KlattGrid_createExample()` instead.**

## Vowel Synthesis

### The Vowel Triangle

The three corner vowels (/i/, /a/, /u/) define the vowel space:

```{r vowel-triangle}
# /i/ - high front vowel (English "beet")
kg_i <- KlattGrid_createFromVowel(
  duration = 0.3, f0start = 200,
  f1 = 280, b1 = 50,
  f2 = 2250, b2 = 100,
  f3 = 2890, b3 = 120
)

# /a/ - low central vowel (English "father")
kg_a <- KlattGrid_createFromVowel(
  duration = 0.3, f0start = 200,
  f1 = 730, b1 = 80,
  f2 = 1090, b2 = 120,
  f3 = 2440, b3 = 140
)

# /u/ - high back vowel (English "boot")
kg_u <- KlattGrid_createFromVowel(
  duration = 0.3, f0start = 200,
  f1 = 310, b1 = 60,
  f2 = 870, b2 = 90,
  f3 = 2250, b3 = 140
)

# Synthesize all three
sound_i <- kg_i$to_sound()
sound_a <- kg_a$to_sound()
sound_u <- kg_u$to_sound()

cat("Vowel triangle synthesized:\n")
cat("  /i/: F1=", 280, " F2=", 2250, " Hz (high front)\n", sep="")
cat("  /a/: F1=", 730, " F2=", 1090, " Hz (low central)\n", sep="")
cat("  /u/: F1=", 310, " F2=", 870, " Hz (high back)\n", sep="")
```

### Formant Values for Common Vowels

Adult male (~120 Hz pitch):

| Vowel | IPA | F1 (Hz) | F2 (Hz) | F3 (Hz) | Example Word |
|-------|-----|---------|---------|---------|--------------|
| beet  | /i/ | 280     | 2250    | 2890    | "beet"       |
| bit   | /ɪ/ | 400     | 2000    | 2550    | "bit"        |
| bait  | /e/ | 550     | 1770    | 2490    | "bait"       |
| bet   | /ɛ/ | 660     | 1720    | 2410    | "bet"        |
| bat   | /æ/ | 860     | 1760    | 2390    | "bat"        |
| father| /a/ | 730     | 1090    | 2440    | "father"     |
| bought| /ɔ/ | 570     | 840     | 2410    | "bought"     |
| boat  | /o/ | 500     | 700     | 2410    | "boat"       |
| book  | /ʊ/ | 440     | 1020    | 2240    | "book"       |
| boot  | /u/ | 310     | 870     | 2250    | "boot"       |

### Voice Types

Different speakers require different formant and pitch values:

```{r voice-types}
# Male voice (low pitch, lower formants)
kg_male <- KlattGrid_createFromVowel(
  duration = 0.5, f0start = 100,
  f1 = 750, b1 = 80,
  f2 = 1100, b2 = 120,
  f3 = 2400, b3 = 150
)

# Female voice (higher pitch, higher formants)
kg_female <- KlattGrid_createFromVowel(
  duration = 0.5, f0start = 220,
  f1 = 850, b1 = 70,
  f2 = 1300, b2 = 110,
  f3 = 2800, b3 = 140
)

# Child voice (very high pitch)
kg_child <- KlattGrid_createFromVowel(
  duration = 0.4, f0start = 300,
  f1 = 900, b1 = 60,
  f2 = 1400, b2 = 100,
  f3 = 3000, b3 = 130
)

cat("Pitch ranges:\n")
cat("  Male:   80-180 Hz (typical ~100-120 Hz)\n")
cat("  Female: 160-250 Hz (typical ~200-220 Hz)\n")
cat("  Child:  250-400 Hz (typical ~300 Hz)\n")
```

## Pitch Contours

### Dynamic Intonation

Add pitch points to create dynamic intonation patterns:

```{r pitch-contour}
# Start with base vowel
kg <- KlattGrid_createFromVowel(
  duration = 0.6, f0start = 120,
  f1 = 500, b1 = 50,
  f2 = 1500, b2 = 100,
  f3 = 2500, b3 = 150
)

# Add rising intonation (question)
kg$add_pitch_point(0.2, 110)  # Slight dip
kg$add_pitch_point(0.4, 105)  # Lower
kg$add_pitch_point(0.6, 150)  # Rise at end

sound <- kg$to_sound()
pitch <- sound$to_pitch()

# Visualize pitch contour
df <- as.data.frame(pitch)
df <- df[!is.na(df$frequency), ]

ggplot(df, aes(time, frequency)) +
  geom_line(color = "#E41A1C", linewidth = 1) +
  geom_point(color = "#E41A1C", size = 2, alpha = 0.5) +
  labs(
    title = "Question Intonation Pattern",
    subtitle = "Rising pitch contour: 120 → 105 → 150 Hz",
    x = "Time (s)",
    y = "Frequency (Hz)"
  ) +
  theme_minimal()
```

### Common Pitch Patterns

```{r pitch-patterns, eval=FALSE}
# Flat pitch (citation form)
kg_flat <- KlattGrid_createFromVowel(
  duration = 0.5, f0start = 120,
  f1 = 500, b1 = 50, f2 = 1500, b2 = 100, f3 = 2500, b3 = 150
)
# No additional points needed

# Falling pitch (statement)
kg_fall <- KlattGrid_createFromVowel(
  duration = 0.5, f0start = 150,
  f1 = 500, b1 = 50, f2 = 1500, b2 = 100, f3 = 2500, b3 = 150
)
kg_fall$add_pitch_point(0.5, 100)  # Fall from 150 to 100 Hz

# Rising pitch (question)
kg_rise <- KlattGrid_createFromVowel(
  duration = 0.5, f0start = 100,
  f1 = 500, b1 = 50, f2 = 1500, b2 = 100, f3 = 2500, b3 = 150
)
kg_rise$add_pitch_point(0.5, 180)  # Rise from 100 to 180 Hz
```

## Formant Transitions

### Diphthongs

Create smooth vowel-to-vowel transitions:

```{r diphthong, eval=FALSE}
# /ai/ diphthong: /a/ → /i/ (as in "buy")
kg <- KlattGrid_createFromVowel(
  duration = 0.5, f0start = 150,
  f1 = 730, b1 = 80,      # Start: /a/
  f2 = 1090, b2 = 120,
  f3 = 2440, b3 = 140
)

# Transition to /i/ at end (formantType 1 = oral formants)
kg$add_formant_point(1, 1, 0.5, 280)   # F1: 730 → 280 Hz
kg$add_formant_point(1, 2, 0.5, 2250)  # F2: 1090 → 2250 Hz
kg$add_formant_point(1, 3, 0.5, 2890)  # F3: 2440 → 2890 Hz

sound <- kg$to_sound()

# Extract formants to visualize transition
frm_result <- sound$to_formant_burg(max_frequency = 5500)
df <- as.data.frame(frm_result)
df <- df[df$formant %in% 1:3, ]
names(df)[names(df) == "formant"] <- "formant_number"

ggplot(df, aes(time, frequency, color = factor(formant_number))) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_color_manual(
    values = c("1" = "#E41A1C", "2" = "#377EB8", "3" = "#4DAF4A"),
    labels = c("F1", "F2", "F3")
  ) +
  labs(
    title = "Diphthong /ai/ Formant Transitions",
    subtitle = "/a/ → /i/ transition over 500 ms",
    x = "Time (s)",
    y = "Frequency (Hz)",
    color = "Formant"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

### Vowel Continuum

Generate perceptual continua for experiments:

```{r vowel-continuum, eval=FALSE}
# Create /i/ to /u/ continuum (10 steps)
f1_seq <- seq(280, 310, length.out = 10)  # F1: 280 (/i/) → 310 (/u/)
f2_seq <- seq(2250, 870, length.out = 10)  # F2: 2250 → 870

vowel_continuum <- lapply(1:10, function(i) {
  kg <- KlattGrid_createFromVowel(
    duration = 0.3, f0start = 200,
    f1 = f1_seq[i], b1 = 50,
    f2 = f2_seq[i], b2 = 100,
    f3 = 2500, b3 = 150
  )
  kg$to_sound()
})

# Save each step
for (i in 1:10) {
  vowel_continuum[[i]]$save(sprintf("continuum_%02d.wav", i), "WAV")
}
```

## Visualizing Synthesis

### Spectrogram of Synthetic Vowel

```{r spectrogram, eval=FALSE, fig.width=8, fig.height=5}
# Synthesize /a/ vowel
kg <- KlattGrid_createFromVowel(
  duration = 0.5, f0start = 120,
  f1 = 730, b1 = 80,
  f2 = 1090, b2 = 120,
  f3 = 2440, b3 = 140
)

sound <- kg$to_sound()

# Create spectrogram
spec <- sound$to_spectrogram(
  window_length = 0.005,
  max_frequency = 5000
)

# Extract formants for overlay
frm_result <- sound$to_formant_burg(max_frequency = 5500)
formant_df <- as.data.frame(frm_result)
formant_df <- formant_df[formant_df$formant %in% 1:3, ]
formant_df$formant_num <- formant_df$formant

# Plot
autoplot(spec) +
  geom_line(
    data = formant_df,
    aes(time, frequency, color = factor(formant_num)),
    linewidth = 1.5, alpha = 0.8, inherit.aes = FALSE
  ) +
  scale_color_manual(
    values = c("1" = "red", "2" = "yellow", "3" = "white"),
    labels = c("F1", "F2", "F3")
  ) +
  labs(
    title = "Synthetic /a/ Vowel Spectrogram",
    subtitle = "F1=730, F2=1090, F3=2440 Hz (male voice, 120 Hz pitch)",
    color = "Formant"
  ) +
  theme(legend.position = "right")
```

### Comparing Real vs Synthetic

```{r compare-real-synthetic, fig.width=8, fig.height=6}
# Load real speech
sound_real <- Sound(system.file("extdata", "test.wav", package = "pladdrr"))

# Extract mean formants from real speech
formant_real <- sound_real$to_formant_burg()
df_real <- as.data.frame(formant_real)

f1_mean <- mean(df_real$frequency[df_real$formant == 1], na.rm = TRUE)
f2_mean <- mean(df_real$frequency[df_real$formant == 2], na.rm = TRUE)
f3_mean <- mean(df_real$frequency[df_real$formant == 3], na.rm = TRUE)

cat("Extracted formants from real speech:\n")
cat(sprintf("  F1: %.0f Hz\n", f1_mean))
cat(sprintf("  F2: %.0f Hz\n", f2_mean))
cat(sprintf("  F3: %.0f Hz\n", f3_mean))

# Resynthesize with extracted formants
kg_synth <- KlattGrid_createFromVowel(
  duration = min(sound_real$get_duration(), 1.0),
  f0start = 120,
  f1 = f1_mean, b1 = 80,
  f2 = f2_mean, b2 = 120,
  f3 = f3_mean, b3 = 150
)

sound_synth <- kg_synth$to_sound()

# Compare spectrograms side-by-side
spec_real <- sound_real$to_spectrogram(max_frequency = 5000)
spec_synth <- sound_synth$to_spectrogram(max_frequency = 5000)

cat("\nSynthetic sound matches real speech formant averages\n")
```

## Advanced Features

### Voice Quality Control

Manipulate voicing amplitude for breathiness or creaky voice:

```{r voice-quality, eval=FALSE}
kg <- KlattGrid_createFromVowel(
  duration = 0.5, f0start = 120,
  f1 = 800, b1 = 80, f2 = 1200, b2 = 120, f3 = 2500, b3 = 150
)

# Add breathiness (fade voicing)
kg$add_voicing_amplitude_point(0.0, 60)  # Start normal
kg$add_voicing_amplitude_point(0.5, 30)  # End breathy

sound <- kg$to_sound()
```

### Complex Synthesis Example

Use the pre-configured example for complex synthesis:

```{r complex-example}
# Pre-configured grid with multiple tiers
kg_complex <- KlattGrid_createExample()

# Synthesize full signal
sound_complex <- kg_complex$to_sound()

cat("Complex KlattGrid example:\n")
cat("  Duration:", sound_complex$get_duration(), "s\n")
cat("  Uses multiple pitch points, formant tiers, and amplitude control\n")
```

## Analysis → Synthesis Workflow

### Complete Pipeline

```{r analysis-synthesis-workflow}
# 1. Load real speech
sound_orig <- Sound(system.file("extdata", "test.wav", package = "pladdrr"))

# 2. Extract formants (using FormantPath for robustness)
fp <- sound_orig$to_formant_path(num_steps_up_down = 2L)
frm_result <- fp$extract_formant()

# 3. Get mean formant values
df <- as.data.frame(frm_result)
f1_mean <- mean(df$frequency[df$formant == 1], na.rm = TRUE)
f2_mean <- mean(df$frequency[df$formant == 2], na.rm = TRUE)
f3_mean <- mean(df$frequency[df$formant == 3], na.rm = TRUE)

# 4. Extract pitch
pitch <- sound_orig$to_pitch()
f0_mean <- mean(as.data.frame(pitch)$frequency, na.rm = TRUE)

cat("Extracted parameters:\n")
cat(sprintf("  F0: %.0f Hz\n", f0_mean))
cat(sprintf("  F1: %.0f Hz\n", f1_mean))
cat(sprintf("  F2: %.0f Hz\n", f2_mean))
cat(sprintf("  F3: %.0f Hz\n", f3_mean))

# 5. Resynthesize
kg_resynth <- KlattGrid_createFromVowel(
  duration = min(sound_orig$get_duration(), 1.0),
  f0start = f0_mean,
  f1 = f1_mean, b1 = 80,
  f2 = f2_mean, b2 = 120,
  f3 = f3_mean, b3 = 150
)

sound_resynth <- kg_resynth$to_sound()

cat("\nResynthesis complete - compare original vs synthetic\n")
```

## Use Cases

### 1. Perceptual Experiments

Control stimuli precisely for categorical perception studies:

```{r perceptual-experiment, eval=FALSE}
# Voice onset time (VOT) continuum simulation
# Generate /ba/ to /pa/ continuum by manipulating voice onset

vot_values <- seq(0, 60, by = 10)  # 0-60 ms VOT

stimuli <- lapply(vot_values, function(vot_ms) {
  # Voicing delay = VOT
  kg <- KlattGrid_createFromVowel(
    duration = 0.5, f0start = 120,
    f1 = 730, b1 = 80, f2 = 1090, b2 = 120, f3 = 2440, b3 = 140
  )
  
  # Manipulate voicing onset (simplified)
  if (vot_ms > 0) {
    kg$add_voicing_amplitude_point(0, 0)
    kg$add_voicing_amplitude_point(vot_ms / 1000, 60)
  }
  
  kg$to_sound()
})
```

### 2. Voice Morphing

Create intermediate voices between two speakers:

```{r voice-morphing, eval=FALSE}
# Male voice parameters
f1_male <- 730
f2_male <- 1090
f0_male <- 100

# Female voice parameters
f1_female <- 850
f2_female <- 1300
f0_female <- 220

# Morph continuum (10 steps)
morph_steps <- seq(0, 1, length.out = 10)

morphed_voices <- lapply(morph_steps, function(alpha) {
  # Linear interpolation
  f1 <- f1_male + alpha * (f1_female - f1_male)
  f2 <- f2_male + alpha * (f2_female - f2_male)
  f0 <- f0_male + alpha * (f0_female - f0_male)
  
  kg <- KlattGrid_createFromVowel(
    duration = 0.5, f0start = f0,
    f1 = f1, b1 = 80, f2 = f2, b2 = 120, f3 = 2500, b3 = 150
  )
  
  kg$to_sound()
})
```

### 3. Teaching Phonetics

Demonstrate formant-vowel relationships interactively:

```{r teaching-demo, eval=FALSE}
# Function to synthesize any vowel from F1/F2
synthesize_vowel <- function(f1, f2, label = "") {
  kg <- KlattGrid_createFromVowel(
    duration = 0.5, f0start = 120,
    f1 = f1, b1 = 80, f2 = f2, b2 = 120, f3 = 2500, b3 = 150
  )
  
  sound <- kg$to_sound()
  filename <- sprintf("vowel_%s_F1%d_F2%d.wav", label, f1, f2)
  sound$save(filename, "WAV")
  
  cat(sprintf("Created %s: F1=%d, F2=%d Hz\n", label, f1, f2))
}

# Generate examples
synthesize_vowel(280, 2250, "i")
synthesize_vowel(730, 1090, "a")
synthesize_vowel(310, 870, "u")
```

## Known Limitations

### 1. Empty Grid Errors on Synthesis

**Issue**: Direct synthesis from an empty `KlattGrid()` errors (`"Pitch tier should not be empty."`)

**Workaround**: Always use `KlattGrid_createFromVowel()` or `KlattGrid_createExample()`

**Example of safe usage:**
```{r safe-usage, eval=FALSE}
# safe
kg <- KlattGrid_createFromVowel(
  duration = 0.5, f0start = 120,
  f1 = 800, b1 = 80, f2 = 1200, b2 = 120, f3 = 2500, b3 = 150
)

# safe
kg <- KlattGrid_createExample()

# unsafe - errors, uninitialized tiers
kg <- KlattGrid(tmin = 0, tmax = 1, numberOfFormants = 5)
sound <- kg$to_sound()  # errors
```

### 2. Limited Tier Manipulation

**Current limitations:**
- Cannot query tier contents directly
- Cannot remove points after adding
- No direct formant amplitude control

**Best practice:** Create new grid with desired parameters rather than modifying existing grid

### 3. Formant Constraints

**Physical constraints that must be maintained:**
- F1 < F2 < F3 (formants increase monotonically)
- Bandwidths > 0 Hz (zero bandwidth causes artifacts)
- Pitch in range 50-400 Hz (outside this sounds unnatural)

## Best Practices

### Do

1. **Always use `KlattGrid_createFromVowel()`** for vowel synthesis
2. **Start with known formant values** from the reference table
3. **Adjust formants gradually** (±50-100 Hz steps)
4. **Keep bandwidths proportional**: B1 < B2 < B3 (typically 50-80-120 Hz)
5. **Match pitch to speaker type**: male ~100-120 Hz, female ~200-220 Hz
6. **Test with short durations** (0.3-0.5s) before scaling up
7. **Export to WAV** for analysis in other tools

### Don't

1. **Don't use empty `KlattGrid()` constructor** (errors on synthesis)
2. **Don't violate F1 < F2 < F3** (unnatural synthesis)
3. **Don't use zero bandwidth** (causes artifacts)
4. **Don't expect consonant synthesis** (formants only, no articulation model)
5. **Don't modify existing grids** (create new ones instead)

## Troubleshooting

### Error on `to_sound()`

**Cause**: Incomplete grid initialization

**Solution**: Use `KlattGrid_createFromVowel()` instead of empty grid

### Unnatural sounding synthesis

**Cause**: Formant values out of range or wrong bandwidths

**Solution**: 
- Check F1 < F2 < F3 relationship
- Use bandwidth ratio B1:B2:B3 ≈ 1:1.5:2
- Verify pitch is in 80-300 Hz range

### No sound output

**Cause**: Voicing amplitude too low or pitch out of range

**Solution**: Ensure `f0start` is 80-300 Hz and voicing amplitude > 40 dB

## Comparison with Other Tools

### pladdrr vs Praat GUI

| Feature | pladdrr KlattGrid | Praat GUI |
|---------|-------------------|-----------|
| Vowel synthesis | `createFromVowel()` | Manual setup |
| Pitch contours | `add_pitch_point()` | PitchTier editor |
| Batch synthesis | R loops | Manual, one at a time |
| Parametric experiments | Scriptable continua | Repeated manual steps |
| Visualization | ggplot2 | Built-in |

### vs Parselmouth

Parselmouth does not currently expose KlattGrid - pladdrr is the only R package with direct Klatt synthesis support.

### vs Other R Synthesis

Existing R packages (e.g., `phonTools`) use simpler synthesis methods. pladdrr's KlattGrid wraps Praat's own cascade/parallel Klatt synthesizer C++ code directly, rather than reimplementing it.

## Summary

**KlattGrid is used for:**
- Perceptual experiments requiring controlled stimuli
- Voice manipulation research
- Teaching phonetic concepts
- Speech resynthesis with modified parameters

**Key functions:**
- `KlattGrid_createFromVowel()` - Safe vowel synthesis
- `add_pitch_point()` - Dynamic intonation
- `add_formant_point()` - Formant transitions
- `to_sound()` - Generate audio

**Typical workflow:**
1. Define formant parameters (F1/F2/F3, B1/B2/B3)
2. Set pitch contour (f0start + optional points)
3. Create grid with `createFromVowel()`
4. Synthesize with `to_sound()`
5. Analyze or export

## Further Reading

- Klatt, D. H. (1980). "Software for a cascade/parallel formant synthesizer". *JASA*, 67(3), 971-995.
- Praat manual: [KlattGrid](https://www.fon.hum.uva.nl/praat/manual/KlattGrid.html)
- pladdrr documentation: `?KlattGrid_createFromVowel`
- See also: `vignette("formantpath-robust-tracking")` for formant extraction

## Session Info

```{r session-info}
sessionInfo()
```
