---
title: "TextGrid Workflows with pladdrr"
author: "pladdrr Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TextGrid Workflows with pladdrr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE  # Disable evaluation due to complex dependencies
)
```

## Overview

TextGrids are Praat's standard format for time-aligned speech annotation. This vignette demonstrates:

- **Creating TextGrids** from scratch or from existing files
- **Querying annotations** (intervals, points, boundaries)
- **Modifying annotations** (insert, delete, relabel)
- **Tier management** (add, remove, rename)
- **Large-scale corpus processing** with performance optimization
- **Data export** to R data frames and CSV
- **Integration with forced alignment** tools

TextGrids are central to phonetic research, enabling systematic analysis of annotated speech data.

## Background

### TextGrid Structure

A TextGrid contains **tiers** of two types:

1. **Interval tiers**: Contiguous segments with start/end times and labels
   - Use cases: words, phones, syllables, utterances, speaker turns
   - Each interval has: start time, end time, text label
   
2. **Point tiers**: Discrete time points with labels
   - Use cases: landmarks (e.g., burst release, vowel onset), events, boundaries
   - Each point has: time, text label

### Common Workflows

**Manual annotation**:
- Create TextGrid in Praat
- Manually segment and label
- Load in R for analysis

**Forced alignment**:
- Use tools like Montreal Forced Aligner, WebMAUS, or Gentle
- Generates word- and phone-level TextGrids automatically
- Load in R for quality control and analysis

**Hybrid approach**:
- Automatic alignment for initial segmentation
- Manual correction in Praat
- Batch analysis in R

## Part 1: Creating TextGrids

### Creating from Scratch

```{r create_textgrid}
library(pladdrr)

# Create a TextGrid with multiple tiers
tg <- TextGrid$create(
  tmin = 0,
  tmax = 10,
  tier_names = "utterances words phones",
  point_tiers = "events"  # Space-separated list of point tier names
)

cat("Created TextGrid with", tg$get_number_of_tiers(), "tiers\n")
cat("Duration:", tg$get_total_duration(), "seconds\n")
```

### Adding Interval Boundaries

```{r add_boundaries}
# Add word boundaries to the "words" tier (tier 2)
word_tier <- 2

# Insert boundaries at specific times
tg$insert_boundary(word_tier, time = 1.0)
tg$insert_boundary(word_tier, time = 2.5)
tg$insert_boundary(word_tier, time = 4.0)
tg$insert_boundary(word_tier, time = 6.5)
tg$insert_boundary(word_tier, time = 8.0)

# Label the created intervals
tg$set_interval_text(word_tier, 1, "")
tg$set_interval_text(word_tier, 2, "hello")
tg$set_interval_text(word_tier, 3, "world")
tg$set_interval_text(word_tier, 4, "this")
tg$set_interval_text(word_tier, 5, "is")
tg$set_interval_text(word_tier, 6, "test")

cat("Added", tg$get_number_of_intervals(word_tier), "intervals to words tier\n")
```

### Adding Point Annotations

```{r add_points, eval=FALSE}
# Add events to the point tier (tier 4)
# Note: Point tier operations require proper tier setup
events_tier <- 4

# tg$insert_point(events_tier, time = 0.5, mark = "recording_start")
# tg$insert_point(events_tier, time = 5.0, mark = "sentence_boundary")
# tg$insert_point(events_tier, time = 9.5, mark = "recording_end")

# cat("Added", tg$get_number_of_points(events_tier), "points to events tier\n")
```

### Loading from File

```{r load_textgrid, eval=FALSE}
# Load existing TextGrid
tg <- TextGrid$new("path/to/annotation.TextGrid")

# Or from package example data
tg_file <- system.file("extdata", "benchmarkdata1min.TextGrid", package = "pladdrr")
tg <- TextGrid$new(tg_file)
```

## Part 2: Querying TextGrids

### Tier Information

```{r query_tiers}
# Get tier names and types
tier_names <- tg$get_tier_names()
n_tiers <- tg$get_number_of_tiers()

for (i in 1:n_tiers) {
  is_interval <- tg$tier_is_interval_tier(i)
  tier_type <- if (is_interval) "IntervalTier" else "PointTier"
  
  n_items <- if (is_interval) {
    tg$get_number_of_intervals(i)
  } else {
    tg$get_number_of_points(i)
  }
  
  cat(sprintf("Tier %d: %s (%s) - %d items\n", 
              i, tier_names[i], tier_type, n_items))
}
```

### Extracting Interval Information

```{r query_intervals}
# Get all intervals from a specific tier
tier_idx <- 2  # words tier
n_intervals <- tg$get_number_of_intervals(tier_idx)

# Create data frame with interval information
intervals_df <- data.frame(
  index = 1:n_intervals,
  start = numeric(n_intervals),
  end = numeric(n_intervals),
  duration = numeric(n_intervals),
  label = character(n_intervals),
  stringsAsFactors = FALSE
)

for (i in 1:n_intervals) {
  intervals_df$start[i] <- tg$get_interval_start_time(tier_idx, i)
  intervals_df$end[i] <- tg$get_interval_end_time(tier_idx, i)
  intervals_df$duration[i] <- intervals_df$end[i] - intervals_df$start[i]
  intervals_df$label[i] <- tg$get_interval_text(tier_idx, i)
}

print(intervals_df)
```

### Time-Based Queries

```{r time_queries}
# Find interval at specific time
query_time <- 3.0
interval_idx <- tg$get_interval_at_time(tier_idx, query_time)

if (!is.na(interval_idx)) {
  label <- tg$get_interval_text(tier_idx, interval_idx)
  start <- tg$get_interval_start_time(tier_idx, interval_idx)
  end <- tg$get_interval_end_time(tier_idx, interval_idx)
  
  cat(sprintf("At time %.2f s: '%s' [%.2f - %.2f]\n", 
              query_time, label, start, end))
}

# Find interval by label
target_label <- "world"
for (i in 1:n_intervals) {
  if (tg$get_interval_text(tier_idx, i) == target_label) {
    cat(sprintf("Found '%s' at interval %d\n", target_label, i))
    break
  }
}
```

### Point Tier Queries

```{r query_points}
# Get all points from events tier
points_tier <- 4
n_points <- tg$get_number_of_points(points_tier)

for (i in 1:n_points) {
  time <- tg$get_point_time(points_tier, i)
  mark <- tg$get_point_label(points_tier, i)
  cat(sprintf("Point %d: %.3f s - '%s'\n", i, time, mark))
}
```

## Part 3: Modifying TextGrids

### Inserting and Removing Boundaries

```{r modify_boundaries}
# Insert a new boundary in the middle of an interval
tier_idx <- 2
tg$insert_boundary(tier_idx, time = 3.25)

# This splits an existing interval into two
# Label the newly created intervals
new_interval <- tg$get_interval_at_time(tier_idx, 3.26)
tg$set_interval_label(tier_idx, new_interval, label = "new")

cat("Inserted boundary, now have", 
    tg$get_number_of_intervals(tier_idx), "intervals\n")

# Remove a boundary (merges adjacent intervals)
# tg$remove_boundary(tier_idx, time = 3.25)
```

### Relabeling

```{r relabel}
# Change interval label
tier_idx <- 2
interval_idx <- 3
tg$set_interval_label(tier_idx, interval_idx, label = "WORLD")

# Change point label
points_tier <- 4
tg$set_point_label(points_tier, point_number = 2, label = "BOUNDARY")

cat("Updated labels\n")
```

### Adding and Removing Tiers

```{r modify_tiers}
# Add a new interval tier
tg$add_interval_tier(name = "syllables")
cat("Added 'syllables' tier, now have", tg$get_number_of_tiers(), "tiers\n")

# Add a new point tier
tg$add_point_tier(name = "landmarks")
cat("Added 'landmarks' tier, now have", tg$get_number_of_tiers(), "tiers\n")

# Remove a tier (use with caution!)
# tg$remove_tier(tier_number = 6)

# Rename a tier
tg$set_tier_name(tier_number = 3, name = "segments")
cat("Renamed tier 3 to 'segments'\n")
```

### Duplicating Tiers

```{r duplicate_tier}
# Duplicate a tier for manual editing or comparison
# tg$duplicate_tier(tier_number = 2, new_name = "words_corrected")
# cat("Duplicated 'words' tier\n")
```

## Part 4: Large-Scale Corpus Analysis

### Loading Large TextGrids

For large TextGrids (many intervals/points), consider performance:

```{r large_textgrid, eval=FALSE}
# Load large file
tg_file <- system.file("extdata", "benchmarkdata10min.TextGrid", package = "pladdrr")

# Measure load time
start_time <- Sys.time()
tg <- TextGrid$new(tg_file)
load_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))

cat(sprintf("Loaded in %.3f seconds\n", load_time))
cat(sprintf("Duration: %.2f minutes\n", tg$get_total_duration() / 60))
```

### Efficient Data Extraction

Extract all data at once using the built-in conversion:

```{r efficient_extraction, eval=FALSE}
# Convert entire TextGrid to data frame
tg_df <- tg$to_data_frame()

# Or specific tiers only
tier_numbers <- c(1, 2)  # First two tiers
tg_df <- tg$to_data_frame(tier_numbers = tier_numbers)

# Now use standard R operations
summary(tg_df)
table(tg_df$tier_name, tg_df$label)
```

### Sampling Large Datasets

For extremely large TextGrids (e.g., multi-hour recordings):

```{r sampling_strategy, eval=FALSE}
# Sample intervals instead of processing all
tier_idx <- 1
n_intervals <- tg$get_number_of_intervals(tier_idx)

# Random sample of 1000 intervals (or all if fewer)
sample_size <- min(1000, n_intervals)
sampled_indices <- sample(1:n_intervals, sample_size)

# Process only sampled intervals
sampled_data <- data.frame()
for (i in sampled_indices) {
  # Extract features...
  # sampled_data <- rbind(sampled_data, ...)
}
```

### Batch Processing Multiple Files

Process an entire corpus of TextGrids:

```{r batch_processing, eval=FALSE}
# List all TextGrid files
textgrid_dir <- "path/to/corpus/"
tg_files <- list.files(textgrid_dir, pattern = "\\.TextGrid$", full.names = TRUE)

# Initialize results
all_data <- data.frame()

# Process each file
for (tg_file in tg_files) {
  cat("Processing:", basename(tg_file), "\n")
  
  tg <- TextGrid$new(tg_file)
  
  # Extract features (example: interval durations)
  tier_idx <- 1
  n_intervals <- tg$get_number_of_intervals(tier_idx)
  
  for (i in 1:n_intervals) {
    start <- tg$get_interval_start_time(tier_idx, i)
    end <- tg$get_interval_end_time(tier_idx, i)
    label <- tg$get_interval_text(tier_idx, i)
    
    all_data <- rbind(all_data, data.frame(
      file = basename(tg_file),
      interval = i,
      label = label,
      duration = end - start
    ))
  }
}

# Analyze corpus-wide
summary(all_data$duration)
table(all_data$label)
```

## Part 5: Integration with Forced Alignment

### Montreal Forced Aligner (MFA) Output

MFA produces TextGrids with word and phone tiers:

```{r mfa_integration, eval=FALSE}
# Load MFA output
mfa_tg <- TextGrid$new("speaker01_aligned.TextGrid")

# Typical MFA structure:
# Tier 1: words
# Tier 2: phones

# Extract phone durations
phone_tier <- 2
n_phones <- mfa_tg$get_number_of_intervals(phone_tier)

phone_durations <- data.frame(
  phone = character(n_phones),
  duration = numeric(n_phones),
  stringsAsFactors = FALSE
)

for (i in 1:n_phones) {
  phone_durations$phone[i] <- mfa_tg$get_interval_text(phone_tier, i)
  start <- mfa_tg$get_interval_start_time(phone_tier, i)
  end <- mfa_tg$get_interval_end_time(phone_tier, i)
  phone_durations$duration[i] <- end - start
}

# Analyze by phone class
aggregate(duration ~ phone, data = phone_durations, FUN = mean)
```

### Quality Control for Forced Alignment

Check alignment quality:

```{r alignment_qa, eval=FALSE}
# Flag suspiciously short or long intervals
min_duration <- 0.020  # 20 ms
max_duration <- 0.500  # 500 ms

short_intervals <- phone_durations$duration < min_duration
long_intervals <- phone_durations$duration > max_duration

cat("Intervals needing review:\n")
cat("  Too short:", sum(short_intervals), "\n")
cat("  Too long:", sum(long_intervals), "\n")

# Export for manual correction
flagged <- phone_durations[short_intervals | long_intervals, ]
write.csv(flagged, "alignment_review.csv", row.names = FALSE)
```

## Part 6: Data Export and Visualization

### Export to CSV

```{r export_csv, eval=FALSE}
# Export interval data
write.csv(intervals_df, "textgrid_intervals.csv", row.names = FALSE)

# Export with additional metadata
intervals_df$file <- "recording01.wav"
intervals_df$speaker <- "S01"
intervals_df$tier <- "words"

write.csv(intervals_df, "corpus_annotations.csv", row.names = FALSE)
```

### Save Modified TextGrid

```{r save_textgrid, eval=FALSE}
# Save to file
tg$save("modified_annotation.TextGrid")

# Praat can now open and display your changes
```

### Visualization

For comprehensive examples of TextGrid visualization including tier displays, interval duration distributions, label frequency plots, and multi-tier layouts, see `vignette("visualization")`.

## Best Practices

### 1. Consistent Labeling

Use standardized labels for reproducibility:

```{r labeling_standards, eval=FALSE}
# Define label conventions
vowels <- c("i", "e", "a", "o", "u")
consonants <- c("p", "t", "k", "b", "d", "g", "m", "n")

# Validate labels
for (i in 1:n_intervals) {
  label <- tg$get_interval_text(tier_idx, i)
  if (!(label %in% c(vowels, consonants, ""))) {
    cat("Warning: Non-standard label at interval", i, ":", label, "\n")
  }
}
```

### 2. Handle Empty Intervals

Empty labels (silence, pauses) are common:

```{r handle_empty, eval=FALSE}
# Filter out empty intervals
non_empty <- intervals_df[intervals_df$label != "", ]

# Or analyze empty intervals separately
empty_intervals <- intervals_df[intervals_df$label == "", ]
cat("Mean silence duration:", mean(empty_intervals$duration), "s\n")
```

### 3. Time Precision

Be aware of floating-point precision:

```{r time_precision, eval=FALSE}
# Round times for comparison
time_precision <- 0.001  # 1 ms

boundary1 <- 1.0001
boundary2 <- 1.0009

if (abs(boundary1 - boundary2) < time_precision) {
  cat("Boundaries are effectively identical\n")
}
```

### 4. Backup Before Editing

Always save a copy before modifying:

```{r backup, eval=FALSE}
# Save original
original_tg <- tg  # R6 objects are passed by reference, so duplicate:
# (In practice, save to disk before editing)

# Edit safely
tg$insert_boundary(tier_idx, time = 5.0)

# Revert if needed
# tg <- TextGrid$new("original_file.TextGrid")
```

## Real-World Applications

### 1. Prosodic Analysis

Extract F0 contours aligned to intonational phrases:

```{r prosodic_analysis, eval=FALSE}
# Assume tier 1 has intonational phrase boundaries
sound <- Sound$new("recording.wav")
pitch <- sound$to_pitch()

phrase_tier <- 1
n_phrases <- tg$get_number_of_intervals(phrase_tier)

for (i in 1:n_phrases) {
  start <- tg$get_interval_start_time(phrase_tier, i)
  end <- tg$get_interval_end_time(phrase_tier, i)
  
  f0_mean <- pitch$get_mean(from_time = start, to_time = end, unit = "hertz")
  cat(sprintf("Phrase %d: Mean F0 = %.1f Hz\n", i, f0_mean))
}
```

### 2. Voice Onset Time (VOT) Measurement

Measure time between burst and voicing onset:

```{r vot_measurement, eval=FALSE}
# Assume point tier with "burst" and "voicing_onset" marks
events_tier <- tg$get_tier_number("events")

n_points <- tg$get_number_of_points(events_tier)
burst_times <- c()
voicing_times <- c()

for (i in 1:n_points) {
  mark <- tg$get_point_label(events_tier, i)
  time <- tg$get_point_time(events_tier, i)
  
  if (mark == "burst") burst_times <- c(burst_times, time)
  if (mark == "voicing_onset") voicing_times <- c(voicing_times, time)
}

# Calculate VOTs (assumes paired marks)
vots <- voicing_times - burst_times
cat("Mean VOT:", mean(vots) * 1000, "ms\n")
```

### 3. Automated Tier Creation

Generate derived tiers programmatically:

```{r auto_tier_creation, eval=FALSE}
# Create syllable tier from phone tier
phone_tier <- tg$get_tier_number("phones")
tg$add_interval_tier("syllables")
syllable_tier <- tg$get_number_of_tiers()

# Define syllable nuclei (vowels)
vowels <- c("i", "e", "a", "o", "u")

# Group phones into syllables (simplified logic)
current_syllable_start <- 0
for (i in 1:tg$get_number_of_intervals(phone_tier)) {
  phone <- tg$get_interval_text(phone_tier, i)
  
  if (phone %in% vowels) {
    # Found vowel - mark syllable boundary
    end_time <- tg$get_interval_end_time(phone_tier, i)
    tg$insert_boundary(syllable_tier, time = end_time)
  }
}
```

## Summary

The **pladdrr** package provides comprehensive TextGrid support:

- **Flexible creation** (from scratch or from files)
- **Efficient querying** (intervals, points, time-based)
- **Complete editing** (insert, delete, modify, manage tiers)
- **Large-scale corpus processing** with performance optimization
- **Seamless integration** with forced alignment tools
- **R ecosystem compatibility** for analysis and visualization

TextGrids in pladdrr enable:
- Systematic annotation workflows
- Integration with acoustic analysis
- Reproducible phonetic research
- Large-scale corpus studies

For related workflows, see:

- `vignette("integrated-phonetic-analysis")` - TextGrid + acoustic analysis
- `vignette("vowel-space-analysis")` - Formant extraction pipelines
- `inst/examples/06_textgrid_analysis.R` - Basic operations
- `inst/examples/08_textgrid_corpus_analysis.R` - Large-scale processing

## References

- Boersma, P., & Weenink, D. (2023). *Praat: doing phonetics by computer*. http://www.praat.org/
- McAuliffe, M., Socolof, M., Mihuc, S., Wagner, M., & Sonderegger, M. (2017). Montreal Forced Aligner: Trainable text-speech alignment using Kaldi. *Interspeech 2017*.
- Kisler, T., Reichel, U., & Schiel, F. (2017). Multilingual processing of speech via web services. *Computer Speech & Language*, 45, 326-347.
